### 2023-04-30 R/analysis.qmd
🚀 initial commit

---
title: "Supplement: Fremont Models"
date: "`r Sys.Date()`"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  eval: true
  echo: true
  warning: false
  error: false
---

## Overview

**Goal**: model the distribution of Fremont sites across watersheds.

**Data**: (i) site counts per watershed based on records from the State Historic Preservation Offices of Utah and Nevada and (ii) environmental data from the Oregon State University PRISM Group, as well as the USGS.

**Method**: a GAMM with a negative binomial distribution for count data with dispersion, an an offset to account for the area of each watershed, and an exponential covariance matrix to account for residual spatial autocorrelation.

## R Preamble

```{r libraries}

library(GGally)
library(ggeffects)
library(gratia)
library(gt)
library(here)
library(mgcv)
library(patchwork)
library(sf)
library(sfdep)
library(tidyverse)
library(viridis)

```

## Data

Unit of analysis: watersheds

Dependent variable:  
- a count of **sites** per watershed

Independent variables:  
- **elevation**  
- maize growing degree days (**gdd**, in Celsius)  
- **precipitation** (millimeters)  
- cost-distance (in hours) to **springs** and **streams**  
- **protected** status  
- cost-distance (in hours) to **roads**

The climate variables, precipitation and gdd, were hindcasted for each watershed using Kyle Bocinsky's [paleocar](https://github.com/bocinsky/paleocar) package and averaged over the approximately 1,000 year occupational sequence of the Fremont in the project area. The cost-distance variables are derived from elevation data using Campbell's hiking function and Djikstra's search algorithm and averaged over each watershed. Protected status refers to the proportion of each watershed that is federal land. Note that it and cost-distance to roads are not explanatory variables, but rather controls on potential sources of sampling bias, mostly due to taphonomic processes operating on the archaeological record - basically, modern farms and cities.

Offsets:  
- **area** of each watershed

This is included to account for sampling bias owing to the size of each watershed. The other two are there to account for sampling bias owing to variable survey intensity, as well as the potential for human impacts to cultural resources.

**Note:** not all of these variables make it into the final model as we first evaluate them for potential colinearity and concurvity. An additional concurvity test is performed after fitting the final model. Not all of them have smooths applied in the final model, either.

```{r}

gpkg <- here("data", "western-fremont.gpkg")

utah <- read_sf(gpkg, "utah")

watersheds <- read_sf(gpkg, "watersheds") |> 
  mutate(
    huc4 = str_sub(id, 1, 4),
    huc4 = factor(huc4)
  )

```

```{r}
#| fig-height: 4
#| code-fold: true
#| fig-align: center

max_density_obs <- with(watersheds, max(sites/area))

obs <- ggplot() +
  geom_sf(
    data = utah,
    fill = "gray95"
  ) +
  geom_sf(
    data = watersheds, 
    aes(fill = (sites/area)/max_density_obs),
    color = "gray90", 
    linewidth = 0.2
  ) +
  scale_fill_viridis(
    name = "Relative\nSite Density",
    option = "mako",
    limits = c(0, 1)
  ) +
  # coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.background = element_rect(fill = "gray95", color = "transparent"),
    legend.justification = c("left", "top"),
    legend.position = c(0.66, 0.57)
  )

obs

```

This map shows the relative density of sites based on their observed frequency in the study area. If site density is the ratio of the number of sites in a watershed to the area of that watershed, the _relative_ density is the density estimate for each watershed divided by the maximum density estimate across all watersheds. It is more appropriate to visualize this value as it doesn't have the same strong implications as visualizing the counts. (1) By visualizing the _density_, it does not make assumptions about the absence of sampling bias owing to the area of each watershed. (2) By visualizing the _relative_ density, it does not make assumptions about the total occurrence rate of sites across the study area, i,e, it does not include an estimate of the intercept.

### Test for colinearity

```{r}
#| fig-height: 5
#| fig-asp: 1
#| code-fold: true
#| out-width: "75%"
#| fig-align: center

diagonal_labels <- function(data, mapping, ...) {

  GGally::ggally_text(
    rlang::as_label(mapping$x),
    col = "black",
    size = 4
  )

}

correlations <- function(data, mapping, color = "black", ...) {

  # get the x and y data to use the other code
  x <- GGally::eval_data_col(data, mapping$x)
  y <- GGally::eval_data_col(data, mapping$y)

  ct <- cor.test(x,y)

  r <- unname(ct$estimate)
  rt <- format(r, digits=2)[1]
  tt <- as.character(rt)

  tt <- stringr::str_c(
    tt, 
    GGally::signif_stars(ct$p.value)
  )

  # plot the cor value
  ggally_text(
   label = tt, 
   mapping = aes(),
   xP = 0.5, 
   yP = 0.5, 
   size = 4,
   color = color,
   ...
  )

}

# there's a bug in ggpairs() where it mislabels the y-axis, so we remove the
# text and ticks from this plot

watersheds |> 
  st_drop_geometry() |> 
  select(elevation, precipitation, gdd, streams, springs, protected, roads) |> 
  ggpairs(
    diag = list(continuous = diagonal_labels),
    lower = list(continuous = wrap("points", alpha = 0.3)),
    upper = list(continuous = correlations)
  ) +
  theme_bw(12) +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```

As you can see, there are strong, linear correlations between all the variables, including the familiar inverse relationship between precipitation and gdd, which is driven by elevation in the American West. Based on this and a previous post-hoc concurvity test, we will drop springs.

### Map of covariates

```{r}
#| fig-width: 8
#| fig-asp: 0.9
#| code-fold: true
#| fig-align: center

make_map <- function(fill, name) {

  ggplot() +
    geom_sf(
      data = utah,
      fill = "gray95"
    ) +
    geom_sf(
      data = watersheds, 
      aes(fill = {{fill}}),
      color = "gray90", 
      linewidth = 0.2
    ) +
    scale_fill_viridis(
      name = name,
      option = "mako"
    ) +
    theme_void() +
    theme(
      legend.background = element_rect(fill = "gray95", color = "transparent"),
      legend.justification = c("left", "bottom"),
      legend.position = c(0.65, 0.19)
    )

}

for_print <- 
  (make_map(gdd, "GDD (°C)") + make_map(precipitation, "PPT (mm)")) /
  (make_map(streams, "Streams\n(hours)") + make_map(protected, "Protected"))

bb8 <- st_union(watersheds, utah) |> st_bbox()

dy <- bb8[["ymax"]] - bb8[["ymin"]]
dx <- bb8[["xmax"]] - bb8[["xmin"]]

ggsave(
  plot = for_print,
  filename = here("figures", "covariates.png"),
  width = 7,
  height = 7 * (dy/dx),
  dpi = 300
)

(make_map(gdd, "GDD (°C)") + 
  make_map(precipitation, "PPT (mm)") +
  make_map(streams, "Streams\n(hours)")) /
  (make_map(springs, "Springs\n(hours)") +
     make_map(protected, "Protected") +
     make_map(roads, "Roads\n(hours)"))

```

### Bocinsky 2014 Thresholds

```{r}
#| fig-width: 8
#| fig-asp: 0.45
#| code-fold: true
#| fig-align: center

thresholds <- tibble(
  xintercept = c(300, 1000),
  x = c(320, 1020),
  y = 72,
  variable = c("precipitation", "gdd"),
  label = c("Minimum threshold\n300 mm", "Minimum threshold\n1000°C GDD")
)

values <- watersheds |>
  st_drop_geometry() |> 
  select(precipitation, gdd) |> 
  pivot_longer(
    everything(),
    names_to = "variable"
  )

pretty_breaks <- function(x) {

  xmin <- round(min(x), -2)
  xmax <- round(max(x), -2)

  seq(xmin, xmax, by = 200)

}

threshold_dists <- ggplot() +
  geom_histogram(
    data = values,
    aes(value),
    color = "white",
    binwidth = 100,
    center = 50
  ) +
  geom_vline(
    data = thresholds,
    aes(xintercept = xintercept),
    linewidth = 1.2,
    color = "darkred"
  ) +
  geom_text(
    data = thresholds,
    aes(x, y, label = label),
    color = "darkred",
    hjust = 0,
    vjust = 1
  ) +
  facet_wrap(
    ~variable, 
    scale = "free_x",
    labeller = labeller(
      variable = c(precipitation = "Precipitation (mm)", 
                   gdd = "Maize GDD (°C)")
    ),
    strip.position = "bottom"
  ) +
  scale_x_continuous(breaks = pretty_breaks) +
  labs(
    x = NULL,
    y = "Count of watersheds"
  ) +
  theme_bw(12) +
  theme(
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size = rel(1)),
    strip.placement = "outside"
  )

ggsave(
  plot = threshold_dists,
  filename = here("figures", "threshold_distributions.png"),
  width = 7,
  height = 7 * (dy/(2*dx)),
  dpi = 300
)

threshold_dists

```

Here's what that threshold looks like in geographic space.

```{r}

watersheds <- watersheds |> 
  mutate(
    in_niche = ifelse(gdd >= 1000 & precipitation >= 300, "Yes", "No"),
    in_niche = factor(in_niche)
  )

```

```{r}
#| fig-width: 8
#| fig-asp: 0.45
#| code-fold: true
#| fig-align: center

threshold_map <- ggplot() +
    geom_sf(
      data = utah,
      fill = "gray95"
    ) +
    geom_sf(
      data = watersheds, 
      aes(fill = in_niche),
      color = "gray90", 
      linewidth = 0.2
    ) +
    scale_fill_viridis(
      name = "In Niche?",
      option = "mako",
      end = 0.75,
      discrete = TRUE
    ) +
    theme_void() +
    theme(
      legend.background = element_rect(fill = "gray95", color = "transparent"),
      legend.justification = c("left", "top"),
      legend.position = c(0.66, 0.57)
    )

# ggsave(
#   plot = (obs + threshold_map),
#   filename = here("figures", "threshold_map.png"),
#   width = 7,
#   height = 7 * (dy/(2*dx)),
#   dpi = 300
# )

obs + threshold_map

```

Is there a significant difference in elevation between those in the niche and those outside it?

```{r}

with(watersheds, t.test(elevation ~ in_niche))

```

## Analysis

Here, we go through iterations of model fitting and evaluation, with changes to correct for problems like autocorrelation and concurvity (the smoothed version of co-linearity in the predictors). For each fitted model, we provide a model summary, diagnostic plots, and visualizations of partial effects (sometimes called marginal responses) for each smooth term.

```{r}

fm <- sites ~ 
  s(precipitation, k=5) +
  s(gdd, k=5) +
  s(streams, k=5) +
  s(protected, k=5) + 
  offset(log(area))

fremont <- gam(
  fm, 
  data = watersheds,
  family = poisson, 
  select = TRUE
)

```

```{r}
#| out-width: "80%"
#| fig-align: center

summary(fremont)

appraise(fremont)

draw(fremont, residuals = TRUE)

```

### Dispersion

First, an h-test for scaled dispersion in a poisson model. In a poisson model, the variance is supposed to equal the mean. If this holds, then $\alpha$ in $Var(y) = E(y) + \alpha \cdot E(y)$ should equal zero. A simple, intercept-only linear model can test this idea.

```{r}

# code adopted from AER::dispersiontest()

observed <- model.response(model.frame(fremont))
estimate <- fitted(fremont)

variance <- ((observed - estimate)^2 - observed)/estimate

dispersion_model <- lm(variance ~ 1)

```

```{r}
#| code-fold: true

tibble(
  estimate = as.vector(coefficients(dispersion_model)[1]),
  statistic = as.vector(summary(dispersion_model)$coef[1, 3]),
  null = 0,
  p.value = pnorm(statistic, lower.tail = FALSE)
) |> 
  mutate(across(everything(), round, digits = 4)) |> 
  gt(
    rowname_col = NULL
  ) |> 
  tab_caption(
    caption = gt::html("<p>t-test for Overdispersion<br>Alternative: &alpha; > 0</p>")
  ) |> 
  tab_options(
    table.align = "left",
    container.width = pct(35)
  )

```

Clearly, there is over-dispersion in this model. To correct for this, we re-run the model with a `negative binomial` error distribution, letting the model adjust to $\alpha$. We choose a negative binomial over a quasipoisson as it uses maximum likelihood.

```{r}

fremont <- gam(
  fm, 
  data = watersheds,
  family = nb, 
  select = TRUE
)

```

```{r}
#| out-width: "80%"
#| fig-align: center

summary(fremont)

appraise(fremont)

draw(fremont, residuals = TRUE)

```

### Residual Spatial Autocorrelation

Monte Carlo simulations of Moran's I to test for spatial autocorrelation in the residuals (using sfdep).

```{r}

neighbors <- st_contiguity(watersheds)
weights <- st_weights(neighbors)

global_moran_perm(
  residuals(fremont),
  neighbors,
  weights
)

```

There's significant spatial autocorrelation. To fix this, we will use a generalized additive mixed-model (GAMM) with an exponential spatial correlation structure. This is not strictly correct, as it assumes spatial continuity of points, not polygons, but it's a useful first approximation.

### Spatial Correlation Structure

Add coordinates for watershed centroids to the dataset.

```{r}

xy <- watersheds |> 
  st_centroid() |> 
  st_coordinates() |> 
  as_tibble() |> 
  rename_with(tolower)

watersheds <- bind_cols(watersheds, xy)

```

```{r}

fremont <- gamm(
  fm,
  correlation = corExp(form = ~x+y),
  data = watersheds,
  family = nb,
  verbosePQL = FALSE
)

```

```{r}
#| out-width: "80%"
#| fig-align: center

summary(fremont$lme)

summary(fremont$gam)

appraise(fremont$gam)

draw(fremont$gam)

```

By accounting for spatial autocorrelation, the expected degrees of freedom for gdd and cost-distance to streams are now both 1, which suggests that there are no non-linear effects for those covariates.

### Concurvity

A post-hoc test of non-linear correlations in the predictors.

```{r}

concurvity(fremont$gam)

```

Not good...

### Final model

Cost-distance to roads is non-significant, so we'll remove that variable from the model entirely.

```{r}

fm <- sites ~ 
  s(precipitation, k=5) +
  gdd +
  streams +
  protected +
  offset(log(area))

fremont <- gamm(
  fm,
  correlation = corExp(form = ~x+y),
  data = watersheds,
  family = nb,
  verbosePQL = FALSE
)

```

```{r}
#| out-width: "80%"
#| fig-align: center

summary(fremont$lme)

summary(fremont$gam)

appraise(fremont$gam)

draw(parametric_effects(fremont$gam))

draw(fremont$gam)

```

### Final Moran's I

One more Moran's I test to see if we really got a hold of the spatial autocorrelation. We'll do this with the normalized residuals of the lme inside the GAMM as that incorporates the exponential covariance matrix.

```{r}

global_moran_perm(
  residuals(fremont$lme, type = "normalized"),
  neighbors,
  weights
)

```

Looks good!

### Variance Inflation Factor

There can't be concurvity in the model anymore, as there is only one smoothed term. However, we should now check for potential multi-collinearity in the parameteric terms.

```{r}

# borrowing this code from mgcv.helper
# https://github.com/samclifford/mgcv.helper/blob/master/R/vif.gam.R

vif <- function(object){

  obj.sum <- mgcv::summary.gam(object)

  # estimate of standard deviation of residuals
  s2 <- object$sig2

  # data used to fit the model
  X <- object$model

  # n observations
  n <- nrow(X)

  # omit the intercept term, it can't inflate variance
  v <- -1

  # variance in estimates
  varbeta <- obj.sum$p.table[v,2]^2

  selected_col <- row.names(obj.sum$p.table)[v]
  selected_col <- gsub("TRUE", "", selected_col)

  # variance of all the explanatory variables
  varXj <- apply(X=X[, selected_col],MARGIN=2, var)

  # the variance inflation factor, obtained by rearranging
  # var(beta_j) = s^2/(n-1) * 1/var(X_j) * VIF_j
  VIF <- varbeta/(s2/(n-1)*1/varXj)

  tibble::tibble(
    variable = names(VIF),
    vif = VIF
  )

}

vif(fremont$gam)

```

The VIF for gdd is a little high, but within acceptable limits.

## Results

### Marginal response

Here we estimate the marginal responses (also known as partial effects) for each covariate by letting the target covariate vary and holding the other covariates at their means and predicting the site count (the response). Here we use the {ggeffects} package.

```{r}
#| code-fold: true
#| fig-align: center
#| fig-height: 6
#| fig.asp: 0.93
#| out-width: "70%"

margins <- ggpredict(fremont) |>
  unclass() |>
  bind_rows() |>
  filter(group != "area") |>
  rename(
    "y" = predicted,
    "ymin" = conf.low,
    "ymax" = conf.high,
    "variable" = group
  ) |> 
  mutate(
    # hacky way to "clip" the ribbon before expanding in ggplot
    ymax = ifelse(ymax > 100, 100, ymax)
  )

response_labels <- tibble(
  variable = c("precipitation", "gdd", "streams", "protected"),
  label = c("Precipitation\n(mm)", 
            "Maize GDD\n(°C)", 
            "Streams\n(hours)", 
            "Protected\n(Federal Land)"),
  x = with(watersheds, c(max(precipitation), min(gdd), max(streams), min(protected))),
  y = max(margins$ymax),
  hjust = c(0.95,0.05,0.95,0.05)
)

scale_pretty <- function(x, n = 4L) {

  brks <- pretty(x, n = n)

  nbrks <- length(brks)

  i <- ifelse(
    nbrks > n,
    c(1, nbrks),
    nbrks
  )

  brks[-i]

}

marginal_response <- ggplot(margins, aes(x, y)) +
  geom_ribbon(
    aes(ymin = ymin, ymax = ymax, fill = variable)
  ) + 
  scale_fill_manual(
    values = c("#4F3824", "#D1603D", "#DDB967", "#848FA5")
  ) +
  geom_line(
    linewidth = 2,
    color = alpha('white', 0.5),
    lineend = "round"
  ) +
  geom_line(
    linewidth = 1,
    lineend = "round"
  ) +
  geom_label(
    data = response_labels,
    aes(x, y, label = label),
    size = 4,
    hjust = response_labels$hjust,
    vjust = 1,
    label.size = NA,
    alpha = 0.5,
    lineheight = 0.88
  ) +
  labs(
    x = NULL,
    y = "Number of sites"
  ) +
  facet_wrap(
    ~variable, 
    scale = "free_x"
  ) +
  scale_x_continuous(breaks = scale_pretty) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw(12) +
  theme(
    axis.title.y = element_text(size = 13),
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.text = element_blank()
  )

ggsave(
  plot = marginal_response,
  filename = here("figures", "marginal_response.png"),
  width = 5,
  height = 5 * 0.93,
  dpi = 300
)

marginal_response

```

### Map

```{r}

watersheds <- watersheds |> 
  mutate(
    estimate = fitted(fremont$gam) |> as.vector(),
    residuals = residuals(fremont$lme, type = "normalized")
  )

```

```{r}
#| fig-width: 8
#| fig-asp: 0.9
#| code-fold: true
#| fig-align: center

basemap <- ggplot() +
  geom_sf(
    data = utah,
    fill = "gray95"
  ) +
  theme_void() +
  theme(
    legend.background = element_rect(fill = "gray95", 
                                     color = "transparent"),
    legend.justification = c("left", "bottom"),
    legend.position = c(0.65, 0.19)
  )

obs <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = (sites/area)/max_density_obs),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Observed",
    option = "mako",
    limits = c(0, 1)
  )

max_density_est <- with(watersheds, max(estimate/area))

estimate <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = (estimate/area)/max_density_est),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Estimated",
    option = "mako",
    limits = c(0, 1)
  )

res <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = residuals),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Residuals",
    option = "mako"
  )

ggsave(
  plot = (obs + estimate),
  filename = here("figures", "model-map.png"),
  width = 7,
  height = 7 * (dy/(2*dx)),
  dpi = 300
)

ggsave(
  plot = (estimate + threshold_map),
  filename = here("figures", "threshold_map.png"),
  width = 7,
  height = 7 * (dy/(2*dx)),
  dpi = 300
)

(obs + estimate) / (res + threshold_map)

```

Please note that the residuals are still being shown for the count-level data.

```{r}
#| echo: false

# save model 
saveRDS(fremont, file = here("data", "western-fremont-model.Rds"))

```

## Session Info

```{r}
#| code-fold: true

# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  quarto::quarto_version(), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_sesh

```

### 2023-04-30 manuscript/manuscript.qmd
🚀 initial commit

---
title: The Fremont Frontier
subtitle: Living at the Margins of Maize Farming
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Peter M. Yaworsky
    affiliations:
      - id: aarhus-culture
        address: |
          Department of Culture and Heritage Studies, 
          School of Culture and Society, 
          Aarhus University, 
          Moesgård Allé 20, building 4216, 
          8270 Højbjerg, Denmark   
      - id: aarhus-biodiversity
        address: |
          Center for Biodiversity Dynamics in a Changing World, 
          Department of Biology, 
          Aarhus University, 
          Ny Munkegade 114-116, 
          8000 Aarhus C, Denmark
  - name: Weston McCool
    affiliations:
      - id: utah-anthropology
        address: | 
          University of Utah Department of Anthropology, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
  - name: Jerry D. Spangler
    affiliations:
      - ref: utah-anthropology
  - name: Simon Brewer
    affiliations:
      - id: utah-geography
        address: |
          University of Utah Department of Geography, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
      - id: gcsc
        address: |
          University of Utah Global Change and Sustainability Center, 
          115 South 1460 East, RM 234 FASB, 
          Salt Lake City, UT 84112
  - name: Brian F. Codding
    affiliations:
      - ref: utah-anthropology
      - ref: gcsc
abstract: |
  The Fremont provide an important case study to examine the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear. Using a tree-ring and simulation-based reconstruction of average annual precipitation and maize growing degree days, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin. The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
keywords:
  - Ideal Free Distribution
  - Great Basin
  - Paleoclimate Reconstruction
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Quaternary International
      formatting: preprint
      model: 3p
execute: 
  echo: false
  error: false
  warning: false
---

# Introduction

Archaeological populations of subsistence maize farmers known collectively as "the Fremont" lived at the far northern periphery of maize farming in western North America, in an area encompassing much of the modern state of Utah north of the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989; @madsen1998]. That the Fremont managed to thrive for as long as they did in this part of the world poses an interesting question, for conditions there are extremely harsh and inhospitable for maize farming. Archaeologists interested in that question have typically focused on strategies directly related to farming and subsistence, like irrigation methods, trade, and seasonal high elevation hunting [e.g., @barlow2008; @boomgarden2019; @janetski2002; @hart2021; @madsen1998; @metcalfe1985; @patterson2010; @spangler1993; @morgan2012]. Here, however, we come at the problem from a slightly different angle. We seek to understand not so much *what* they did to make farming effective, but *where* they chose to do it, and why they chose those places over others [e.g., @bocinsky2014; @thomson2020]. Those are not entirely separate questions, of course, as they trade-off each other, but we think a geographic analysis can actually help illuminate some of the costs and benefits of the other strategies adopted by the Fremont. A geographic or spatial model will also help us to investigate the ways that climate change might have structured those costs and benefits, and by extension the settlement decisions of Fremont farmers, which we consider a long-term record of human adaptation to climate change.

As a test case, we focus on those Fremont living in the eastern Great Basin of western Utah, which we refer to as the western Fremont (see @fig-overview). The extreme desert conditions that the western Fremont adapted to provide a useful backdrop for exploring Fremont settlement, for variation in that environment exhibits abrupt and quite dramatic changes over short distances of space and time, thus heightening the differences between areas the western Fremont did and did not occupy. We explore those differences from a socio-ecological perspective, focusing on dynamic interactions between competing individuals on the one hand and individuals and their environment on the other [@bird2006; @kennett2006a]. This dynamic is captured nicely by the Ideal Free Distribution model from Behavioral Ecology, as it represents individual decisions about where to live as a choice between habitats whose benefits to the individual may be constrained by what other individuals are doing. The technical term for these benefits is 'suitability', which the IFD defines as some function of both the habitat's intrinsic or pristine environmental quality and its current population size [@fretwell1969; @codding2015; @winterhalder2010]. The fundamental assumption underlying the IFD is that individuals will behave optimally, that they will choose to settle the habitat with the highest suitability first, meaning the one that offers them the greatest ratio of benefits to costs relative to the available alternatives.

Assuming that western Fremont individuals were adept at responding to variation in their local ecology, in particular, that they optimized their choice of habitats, we can use an empirical or inductive species distribution model of Fremont sites to reconstruct the ecological niche for maize farming [@vernon2021; @yaworsky2020; @elith2009]. This is sometimes referred to as ecological niche modeling. The basic idea here is that we can look at the distribution of Fremont sites as a reliable indicator of the sorts of available environmental conditions that would best promote maize farming in an arid landscape. In effect, we are helping ourselves to a little reverse engineering of the archaeological record [@dennett1995]. We are assuming that whatever strategy - whatever habitat - the Fremont chose was the best alternative available to them, not in any absolute sense, of course, just the best given the constraints and trade-offs Fremont farmers faced in those environments [@jochim2022]. In other words, we are treating their settlement decisions themselves as maize farming adaptations.

![
On the left is an overview map showing the project area with modern satellite imagery. Dark green colors represent high elevation, high precipitation areas. The large beige area in the northwest is the Bonneville Basin. What remains of the Great Salt Lake can be seen just west of Salt Lake City. For visualization purposes, the map on the right shows the relative density of archaeological sites (see the section on site data below for details on how we calculate relative density). Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
](../figures/overview.png){#fig-overview fig-align="center" width="6in"}

# Background

## The Fremont

Archaeologists have nominated a number of artifact types and artifact properties as candidate traits to distinguish the Fremont from their neighbors in space and time, including metates with secondary grinding surfaces, large semi-subterranean pithouses, various clay figurines, trapezoidal rock art, one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and elongated corner-notched arrow points [@madsen1998]. None of these, however, nor any combination of them, has yet proven up to the task of bringing these diverse people under a single, all-encompassing definition [@madsen1998]. Nevertheless, the Fremont do stand out, especially at their population apex roughly 1000 years BP.

Around this time, the Fremont likely reached their greatest geographic extent, inhabiting an area that encompassed most of the modern state of Utah [@janetski2011], with varying levels of support for occupations in Idaho [@dean1992prehistoric], Nevada [@hockett1998], Colorado [@baker1999], and Wyoming [@hakiel1987archery; @smith1992fremont]. These modern political boundaries roughly overlap with the eastern Great Basin and Colorado Plateau physiographic provinces.

![
Linear responses of average water-year precipitation (mm), average growing-season maize gdd (C), and cost-distance to streams (hours) to changes in elevation (m). See methods for how these values were calculated and results for more details.
](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}

A basin and range topography dominates the landscape of the Great Basin, with large, dry valleys and endorheic basins punctuated by north-south trending mountain ranges. Changes in elevation in this setting are quite dramatic, on the order of several thousand meters over a short east-west transect. As elevation drives virtually every climatological process in the region, the local ecology is characterized by equally dramatic extremes [@billings1951; @grayson2011]. As shown in @fig-elevation, net water-year precipitation in the eastern Great Basin tends to increase with elevation, while maize growing degree days (GDD) over the growing-season tends to decrease. An important trade-off thus exists between temperature and precipitation, with warmer and drier conditions at lower elevations and cooler and wetter conditions at higher elevations.

These are the environments into which the Fremont first emerged in western Utah. Models for describing their origins, while diverse, tend to fall into the familiar categories of migration and diffusion. Where models of the former migration variety insist that Ancestral Puebloan farmers migrated northward from the US Southwest, bringing with them drought-tolerant maize varieties and maize farming [@madsen1975; @kidder1924], models of the latter diffusion variety suggest that the Fremont first emerged within these areas as local Archaic populations transitioned away from mobile foraging, adopting semi-sedentary agricultural practices from the southwest in a piecemeal fashion over several hundred years [@jennings1978; @winter1974]. While archaeologists have marshaled an impressive array of arguments and counter-arguments for these two classes of models, a general consensus appears to be coalescing around the idea that the truth lies somewhere in between, that the interaction between a nascent Archaic population existing at extremely low population levels and a burgeoning community of migrant farmers together produced the complex we now call the Fremont [@patterson2015; @simms1986; @spangler1993; @spangler2000; @spangler2013].

Models that try to explain *why* Fremont maize farmers emerged in the far northern periphery of the US Southwest tend to agree - more or less - that at some point farming became a more economical alternative to foraging in the region. Where they differ, it is in whether things got better for farming starting roughly 2000 years BP [@benson2011; @coltrain2002; @matson1988] or worse for foraging [@barlow2002; @broughton2010; @cannon2000; @cannon2001]. In support of the *farming-more-better* model, archaeologists draw on evidence for a suite of favorable climatic conditions that came together around 1600 years BP, including warmer temperatures and increased precipitation (both winter and summer), as well as the expansion of grasslands [@grayson2006; @hemphill1995; @rhode2000]. The greatest expansion of the Fremont, in fact, occurred around 1000 years BP, around the time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures that would have increased the intensity of summer monsoons in the southwest, pushing them further north into the Great Basin and onto the Colorado Plateau [@grayson2006], a pattern reflected in numerous tree ring records [e.g., @graybill1990; @leavitt1994; @stine1990; @salzer2014], though some complications do exist for this suggested pattern [@hart2021].

In response to these sorts of arguments, those who favor the *foraging-more-worse* model [@hart2021; @barlow2002] argue that conditions promoting maize farming would also tend to make foraging a more productive strategy. Plus, farming is an extremely costly endeavor, especially when compared to foraging, including lots of upfront investment of time and energy, so improving conditions favorable to maize agriculture are not by themselves sufficient to explain its origins. To fill that lacuna, archaeologists point to accumulating evidence suggesting that the Fremont transition to agriculture accompanied periods of sustained population growth [@barlow2008; @codding2021; @spangler2013] that likely resulted in the depressed availability of wild resources [@barlow2002; @janetski1997; @simms1986]. So, demographic pressure reduced the efficiency of foraging strategies, thus making maize agriculture the optimal strategy. And that, in turn, led the Fremont to adopt farming as their primary mode of subsistence.

Once they entered the farming niche, the Fremont quickly developed several adaptations for coping with the marginal and stochastic conditions in the eastern Great Basin and the Colorado Plateau, the most important almost certainly being irrigation (for more on this, see the discussion). On the Colorado Plateau, the eastern Fremont settled deep within the steep canyons and washes that flanked the tributaries of major rivers like the Green and Yampa [@yaworsky2021; see also, Yaworsky this issue], in places like Nine Mile Canyon [@spangler1993] and Range Creek Canyon [@hart2021]. In the eastern Great Basin, the western Fremont chose to settle on open, alluvial fans and stream terraces along the lower slopes of most ranges [@janetski2000; @madsen1998; @simms2008], at places like Nephi Mounds [@sharrock1967], Pharo Village [@marwitt1968], Median Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock Village [@aikens1967].

Given the extreme aridity of the west, they were still susceptible to drought, of course, even in those more suitable areas, with a single short growing season making agricultural returns highly variable. To cope with these uncertainties, Fremont farmers may have adopted additional risk-mitigation strategies like spreading farm plots among different micro-environments, a technique otherwise known as plot diversification [@spangler1993; @spangler2013; @patterson2015]. The Fremont also relied heavily on crop storage both to provide a surplus during the winter season and to make-up for poor yields [@spangler2019].

## The Ideal Free Distribution

In population ecology, settlement decisions are represented as a choice between habitats, where each habitat $i$ has some net biological value or benefit, known as its suitability ($S_i$). Typically, though not always, this value exhibits negative density dependence, declining as a function of population size ($N_i$), perhaps owing to increased competition or resource depression. When $N_i$ is very small, the suitability of a habitat approaches its pristine or intrinsic quality ($Q_i$), or its gross value independent of all settlement costs [@greene2001]. Together, these parameters suggest a very simple model of habitat suitability:

$$
S_i = Q_i - f(N_i)
$$

This is known as the Ideal Free Distribution (IFD) model [@fretwell1969]. The IFD model assumes that individuals will choose habitats that maximize $S_i$, settling the most suitable habitat first then shifting to the next most suitable habitat when those become equal, and thereafter infilling each at an equal rate. The aggregate effect is a population distributed at equilibrium, meaning suitability is equal for all habitats. When conditions are ideal (individuals have perfect knowledge) and free (individuals face no additional costs to settlement), this distribution will also satisfy the "input matching" rule, with populations occurring at larger numbers in habitats with greater intrinsic value [@parker1978].

Two circumstances provide important exceptions to this simple model. The first concerns economies of scale, or situations in which $S_i$ exhibits positive density dependence. In ecology, these are known as Allee effects, and they are assumed to occur at low density [@allee1949; @fretwell1969]. The second concerns additional settlement costs, or costs associated with moving from one habitat to another. This is a violation of the free element of the IFD. When it occurs, the result is an ideal despotic distribution (IDD) [@fretwell1969]. This has the consequence of reducing the total population in the more suitable habitat at equilibrium.

Many social and ecological variables may be relevant to defining a general measure of $S_i$, but here we are focusing on ecological determinants of intrinsic habitat quality, $Q_i$, that are specific to maize farmers. Our reasoning here is fairly straightforward. If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters. That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium. So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.

# Materials and Methods

## The Project Area

Because we are focusing on the western Fremont, we constrain the current project area to the eastern Great Basin in western Utah, as shown in @fig-overview, where the Great Basin is defined in the hydrologic sense using watershed boundaries [@grayson2011]. On the far eastern edge of this region, a series of north-south ranges, including the Wasatch, Pahvant, and Tushar Mountains, together stand as the primary physiographic boundaries separating the Great Basin from the Colorado Plateau. On its far western edge, a series of north-south ranges straddle the Nevada state line or lie just inside it, including the Snake Range where Great Basin National Park is located. Its northern periphery is the Great Salt Lake and the West Desert, its southern boundary the Pine Valley Mountains that separate the Great Basin from the Virgin River watershed, just north of St. George, UT. This is a region approximately 107,000 km$^2$ in area, encompassing virtually all of the Bonneville Basin in western Utah.

These boundaries also serve as geographic borders between the western Fremont and other contemporaneous populations. Around the Great Salt Lake and the Utah-Nevada border, Fremont farming gives way to Archaic populations engaged in intensive foraging modes of subsistence. The Wasatch Range and Tushar Mountains separate the western from the eastern Fremont on the Colorado Plateau, and the Pine Valley Mountains serve as a border between the Fremont and the Virgin Ancestral Puebloan to the south [@simms2008].

## The Unit of Analysis

The unit of analysis for this research is the HUC10 watershed (n = 183), as defined by the Watershed Boundary Dataset (WBD) developed by the US Geological Survey and the Natural Resources Conservation Service within the US Department of Agriculture [@usgs2013]. These provide a spatially-explicit proxy for specifically human habitats in the proposed study area. There are two reasons for this. First, the topography of the region is such that the cost of travel between watersheds is often quite significant. As a consequence, individuals are expected to spend more time traveling within watersheds than between them, all else being equal. Second, watersheds are by definition water sinks, funneling all available run-off into their respective stream networks. Given the extreme aridity of the proposed project area, this also makes them critical resource sinks, especially for maize farmers, as watersheds determine how much water might accumulate at a location, as well as its potential for irrigation.

Aggregating to the level of the watershed also reduces the computational burdens of this modeling exercise. In particular, it makes paleoclimate reconstructions more tractable, as we are only estimating the means within each watershed through time (see below for details).

## Site Data

To identify Fremont sites in the project area, we rely on site records and cultural resource reports hosted by the cultural resource information systems in Nevada (NVCRIS) and Utah (Sego) with permission from the respective State Historic Preservation Offices. The Sego system also includes an attribute table with affiliation data that we use to filter Fremont sites where possible. For a previous project, we also performed a stratified random sample of all archaeological sites by county in western Utah and recorded affiliation and other attribute data from their site forms. For this analysis, we compare the results of that previous effort with the sites identified by filtering the Sego system as a form of quality control. The resulting dataset contains 2,248 individual Fremont sites.

These site data include everything from small ceramic scatters to massive Fremont villages like Five Finger Ridge [@janetski1998; @janetski2000]. For this analysis, we assume that these reflect different levels of population size and settlement intensity. All else being equal, scatters should follow from shorter stays by fewer people, villages from longer stays by more people. This is important because population size correlates with a habitat's total suitability according to the IFD model, and our attempts to reconstruct the ecological and geographic borders of the Fremont rely heavily on the input-matching rule, or the idea that habitats with greater intrinsic quality will have larger populations at equilibrium. We, therefore, weight the count of sites in each watershed by the total number of residential features that they contain (+1 for sites with no features). So, a site with no residential features counts as one site, a site consisting of, say, four pithouses counts as four unique sites, and a site containing a roomblock with seven rooms counts as seven sites. This brings the total weighted site count to 2,447, with minimum and maximum values of 0 and 119, respectively.

@fig-overview shows the relative density of Fremont sites across watersheds. The *relative* density refers to the density of feature-weighted sites ($D_i = N_i/area_i$) divided by the maximum density ($max\;D$) across all watersheds, thus scaling the density estimates to the unit interval [0,1]. Using the relative density is just a way of avoiding any implication that we have an independent estimate of the absolute number of Fremont sites in the area, and not just an estimate based on a thinned sample.

## Environmental Covariates

For this analysis, we use both topographic and climatological covariates. The topographic covariate is cost-distance to perennial streams (measured in hours). This provides a coarse grained estimate of water availability, as well as the potential costs of irrigation. To calculate this, we first use the R package FedData [@bocinsky2020] to download perennial stream features from the US National Hydrography Dataset [@usgs2022] and a digital elevation model (DEM) from the US National Elevation Dataset [@usgs2022a]. We then apply Campbell's hiking function [@campbell2019] to slope estimates derived from the DEM using the R package hiker [@vernon2021a]. This allows us to generate estimates of travel time between grid cells in the DEM and then calculate the accumulated cost of travel from each perennial stream to any grid cell within each watershed. We then aggregate these values to the watershed level by taking their mean.

Climate covariates include net water-year precipitation (precipitation, in millimeters, mm) from October to September and maize growing degree days (GDD, in Celsius, C) over the growing season from May to September, as these have a large effect on maize productivity, thus providing important constraints on the patterns of settlement for those who relied on maize for subsistence. We caution that GDD is calculated in such a way as to be insensitive to extreme high and low temperatures (extreme heat and frost, in effect). It is also conceptually understood to be a measure of accumulated temperature between the last and first frost-free days, but in our analysis is calculated using hard start and end dates of May 1 and September 30, under the assumption that these dates will roughly coincide with those frost-free days.

![
Distribution of covariates across watersheds. These include maize growing degree days (GDD), annual precipitation (precipitation), and cost-distance to streams (Streams).
](../figures/covariates.png){#fig-covariates fig-align="center" width="6in"}

We hindcast these climate covariates for each watershed and year over the Fremont sequence from 1550 to 550 years BP (400 to 1400 CE) using the Correlation Adjusted corRelation (CAR) method implemented in the R package paleocar [@bocinsky2019]. This method, in effect, regresses modern climate data against the width of younger tree rings, then uses older rings to predict past climate trends [for further details, see @bocinsky2014; @bocinsky2016]. In this case, we use interpolated precipitation and temperature estimates from the \~800 m resolution PRISM dataset [@prismclimategroup2014] and all tree-ring samples in North America, which were obtained from the International Tree Ring Database using FedData. For each watershed, we took the median values of precipitation and GDD over the Fremont sequence.

@fig-covariates shows the geographic distributions of these covariates over watersheds. As these are all in one way or another driven by elevation, it is reasonable to expect that they will correlate. As part of the exploratory phase of this analysis, we, therefore, perform a series of pair-wise tests of correlation measured using Pearson's R. We also regress the scaled and centered values (or z-scores) of maize GDD, precipitation, and cost-distance to streams on elevation using ordinary least squares (OLS). No doubt, this violates many assumptions of OLS, particularly the identity and independence of the errors, but explanation was not our main goal with these simple linear models. We are only trying to get a rough sense of their general response to changes in elevation, which OLS offers (see @fig-elevation).

Given that our primary covariates all correlate with elevation (reported below), readers may wonder why we do not simply use elevation in our main analysis. The reason for this is two-fold. First, using elevation alone would obscure the trade-off between water availability and temperature. Second, elevation is not explanatory of settlement behavior. A maize farmer does not choose a location for farming *because* it is at, say, 1200 m above sea level. They choose that location *because* it has the right combination of water availability and temperature, that is to say, because conditions there are optimal for maize farming. Were those conditions found at some other elevation, the farmer would be expected to farm there.

## Statistical Modeling and Model Evaluation

The response variable is a count $N$ of sites per watershed $i$, so it should arise as the result of a Poisson process. Given the trade-offs outlined above, we also expect it to exhibit a non-linear response to precipitation, GDD, and cost-distance to streams, so we fit a generalized additive model (GAM) with a log-link to the data using the R package mgcv [@wood2004]. Although we expect the outcome to be Poisson distributed, we use a negative binomial distribution here, as this includes a parameter, $r$, that can account for potential dispersion without giving up the inferential benefits of maximum likelihood.

$$
\begin{aligned}
N_i \sim& \;NB\,(\lambda, r)\\
E\,(N_i) =& \;\lambda_i\\
log\, (\lambda_i) =& \;log\, (area_i) + s(precipitation_i) + streams_i + gdd_i + protected_i
\end{aligned}
$$

Importantly, we add to the model specification a constant offset for the log of the area of each watershed, in effect making this a model of population *density*. This is meant to address the idea that larger watersheds will have more sites just as a matter of chance. While our model as currently specified includes the intercept, we caution that such an estimate is not strictly meaningful for this kind of data, as it assumes that the absolute site counts can be inferred from the sample, which requires that the sample be unbiased - an implausible condition to meet. On that note, we also include a parametric term for the proportion of each watershed classified as federally protected land. This is meant to address potential sampling biases owing to taphonomic processes operating on the archaeological record. Notably, for these data, those are almost entirely the result of Euro-American colonization of the area, in particular, the impacts of modern farming techniques and the development of the Wasatch Front as the urban core of the modern state of Utah. This is a coarse metric, but our expectation is that the greater the share of a watershed with federal protections, the smaller the effect of these anthropogenic impacts, the more sites we should observe.

Model evaluation includes checks for concurvity or non-linear correlation in the smooth terms of the GAM, as well as a Variance Inflation Factor test on the parametric or linear terms. We also test for spatial autocorrelation in the untransformed residuals using Monte Carlo simulations of Moran's I and incorporate an exponential covariance matrix to remove that residual autocorrelation. Several smooth terms had effective degrees of freedom (EDF) equal to one, suggesting no non-linear response in the data. We, therefore, remove the smooth terms for those covariates in the final model, as expressed in the formula above.

All analyses are conducted in the R programming language and environment [@rcoreteam2022]. For details of this analysis, please see the Supplement.

# Results

```{r}
#| label: tbl-results
#| tbl-cap: Model Results
#| tbl-subcap:
#|   - "Parametric Terms"
#|   - "Smooth Terms"
#| layout-nrow: 2
#| layout: "[[1], [-1], [1]]"
#| out-width: 50%

# returns parametric and smooth regression tables
# and q, a list with character strings for reporting estimates in text
here::here("R", "regression-table.R") |> source()

parametric_table
smooth_table

```

As shown in @tbl-results, all linear (or parametric) coefficients are significant in the final model. The intercept `r q[["Intercept"]]` is close to zero when transformed back onto the response scale, likely reflecting the large number of watersheds with zero Fremont sites, though as mentioned above, the intercept estimate is not strictly meaningful for these data. The proportion of watersheds falling under federal protections `r q[["Protected"]]` has a positive effect on site counts, thus confirming our expectation that more sites occur in watersheds with a larger proportion of federal land. Site counts increase in watersheds with larger GDD values, though the effect is small `r q[["Maize GDD"]]`, and site counts decrease with distance from perennial streams, though again the effect is small `r q[["CD to Streams"]]`. The only smooth term in the final model is average precipitation over the water-year, which is significant `r q[["s(Precipitation)"]]`. The effective degrees of freedom (EDF) for precipitation suggests a strong non-linear effect of that covariate on site counts, which increases up to about 450 mm and decreases thereafter.

A concurvity test on the final model was not conducted as there was only one smooth term. As expected, a VIF test for linear correlation in the parametric terms of the final model shows some evidence of correlation, particularly for GDD, which was close to 6, but this value is within acceptable limits at a moderate threshold. The Moran's I test suggests that there is no spatial autocorrelation in the untransformed residuals of the final model (I = 0.233, rank = 381, p = 0.476).

![
Marginal response plots. Each plot shows the response of site counts to changes in a target covariate while holding all other variables at their mean.
](../figures/marginal_response.png){#fig-responses fig-align="center" width="5in"}

# Discussion

Our results indicate that Fremont maize farmers generally preferred warmer watersheds with longer growing seasons, moderate levels of precipitation (in the range of 400-600 mm), and greater access to perennial streams, as shown by the marginal response plots in Fig. 3. While site counts exhibit a positive linear response to maize GDD, for maize has an upper threshold temperature at which point it no longer grows. Plus, increased temperatures trade-off with water availability, with warmer conditions being better for plant growth but at the same time increasing water demand on the plant to keep up with growth [@ramankutty2002]. The linear relationship between site counts and maize GDD is, thus, relative to the observed distribution of that variable within the study area and should not be used to generalize to a larger geographic extent.

Another curious result of our analysis is that site counts do not increase linearly as a function of precipitation, but instead decline after 500 mm. We believe this is likely an artifact of trade-offs related to elevation, specifically the trade-off between precipitation and temperature. We discuss this more below, along with other important implications our results have for the regional prehistory of the eastern Great Basin, particularly for the spread and eventual collapse of maize agriculture.

## A Goldilocks Zone?

Provided that the trade-off between precipitation and temperature is driven largely by elevation in this region, it would seem likely that the western Fremont chose site locations at elevations that maximized agricultural suitability, as evidenced by the estimated distribution of sites across watersheds (see Fig. 4), though we caution that watersheds are coarse grained units of analysis. Given the topography of the study area, this trade-off would also suggest that areas suitable for maize agriculture take the form of a thin band around the lower slopes of mountain ranges (or lower canyons in the Uintas). Such a band would be akin to a Goldilocks zone for maize agriculture (Yaworsky et al, this issue), a place where the optima for GDD and precipitation overlap, though we should not make too much of this idea, for it is an area of overlap for continuous ecological gradients, not a region demarcated by anything like a perfect line. It also must be emphasized that what matters most for maize agriculture is not necessarily where precipitation falls, but where it accumulates once it enters the streamflow network. Of course, that will largely be a function of elevation in this region, or changes in elevation, so it will still be in proximity to the Goldilocks zone, though perhaps not perfectly coincident with it. At any rate, given these considerations, it would perhaps be more accurate to characterize the trade-off confronting the Fremont as one between what we might call water availability (including general precipitation across the streamflow network, water runoff, proximity to perennial surficial sources, and irrigation potential, among other things) and temperature (including, but not limited to, the length of the growing season and the number of frost-free days).

## Irrigation and the Maize Farming Niche

Based on previous research [@benson2011; @adams2006; @bellorado2010], Bocinsky and Kohler [@bocinsky2014] place minimum thresholds for precipitation and maize GDD at 300 mm and 1000°C GDD, respectively. Importantly, these are thresholds for dry-farming, which is typically defined as farming free of irrigation [@benson2011; @cordell2012; @varien1999], so they can be loosely interpreted as minimum temperature and precipitation levels required to grow maize through nothing more than planting and harvesting. The right panel in @fig-thresholds shows the watersheds whose median values for maize GDD and precipitation over the Fremont sequence are above those values. This is similar to Bocinsky's concept of "refugia," but rather than calculate the proportion of years in niche, we are using median values, so refugia are defined here as watersheds that spend at least half the Fremont sequence in the rain-fed maize farming niche ($\geq$ 500 years, in this case).

![
The map on the left shows the relative density of feature-weighted site counts as estimated by the model, with lighter colors representing larger relative densities, darker colors representing smaller relative densities. The map on the right shows which watersheds are in the rain-fed maize farming niche for at least half of the Fremont sequence. The light green color indicates those that are in the niche.
](../figures/threshold_map.png){#fig-thresholds fig-align="center" width="6in"}

It appears that these models tend to agree on the best watersheds for maize farming, though there are some notable exceptions. The western slopes of the Stansbury Range in the north central part of the study area and parts of the Parowan Valley and surrounding areas in the southeast part of the study area are estimated by our model to have non-negligible relative densities, but they fall outside of the niche. Other areas, like much of the Salt Lake valley appear to be in the niche, but are estimated by our model to have relatively low densities. These discrepancies are at least partially explained by limitations of our methodology. For one thing, including the proportion of federal land as a covariate is a very crude way of accounting for sampling bias, so our model might not be picking up on the fact that the Salt Lake valley was actually a highly suitable area for maize agriculture, as evidence to that effect has largely been destroyed. The discrepancy may also be explained by the fact that we have aggregated to the watershed level, thus obscuring important environmental variation across both the landscape in general and site locations in particular. Were we to build a model of the disaggregated data, we would likely find that many agricultural sites do, in fact, fall in the rain-fed maize agricultural niche, even if the watersheds that contain them do not.

But setting aside those methodological issues, some much more interesting explanations may be offered, at least from the perspective of theory. In particular, the spatial covariance structure in our model may be picking up on the structure of the streamflow network, specifically the accumulation of water run-off as a function of slope. This would suggest that if there are more people in one watershed, there should be more more people in the downstream watershed, too, at least relative to other watersheds with similar precipitation and GDD values. In the language of the IFD model, the suitability of the watershed habitat should propagate through the streamflow network. Exploring that intuition in a more systematic way, however, would require much more careful consideration of streamflow. Unfortunately, that is well beyond the scope of the current paper.

Still, the distribution of the rain-fed niche may offer additional insights. Consider the fact that of the 183 watersheds in the study area, 87% (n=159) have median GDD values greater than 1000°C, but only 54% (n=98) have median precipitation values above 300 mm, and a scant 40% (n=74) meet both requirements. This would suggest that precipitation is the primary limiting factor when it comes to the choice of watersheds in which to dry farm, for there are many more watersheds that meet the minimum GDD requirements than meet the minimum precipitation requirements. One important implication of this is that dry farming would have been very nearly impossible across most of the study area over the Fremont sequence and irrigation more or less necessary. We note that even within the niche, which we have defined somewhat liberally as those watersheds that are included for at least half the Fremont sequence, irrigation would almost certainly have been a necessity [@boomgarden2019; @matson1990; @spangler2010]. Of course, the type and intensity of irrigation would probably differ between sites, with more intensive irrigation occurring where there is less precipitation and less intensive irrigation where there is more precipitation, whether that precipitation falls in the watersheds themselves or in the upstream watersheds that feed into them.

## The Spread of Maize Agriculture

Our results comport with previous work suggesting that maize farmers in the eastern Great Basin preferentially targeted a certain elevation band that coincides with alluvial fans, forming a rim around the lower slopes of mountains and ridges on or near important drainages and floodplains. The ruggedness and aridity of this region led to extreme circumscription, with the costs of intensive farming outside these areas being so severe as to render that alternative virtually unsustainable. For maize farmers, the Great Basin should, thus, have the look and feel of an island biogeography, with mountains rising up like islands out of an ocean of desert sand and sagebrush.

To account for the timing and tempo of island colonization among food producers in Oceania, @kennett2006 offer an extension of the IFD framework that might actually apply to the Great Basin. Their model depends crucially on the potential for farming populations to introduce economies of scale or Allee effects, with suitability increasing with increasing population at low densities. They also assume that individuals will prefer nearer habitats to those that are farther away, under the assumption that greater distances impose greater settlement costs, all else being equal. Together, these assumptions suggest a pulsating pattern of maize spread into the North American Southwest, with maize agriculture growing, then spreading, growing, then spreading, as appears to have happened in Oceania.

This need not be a wholesale movement or adoption of agricultural practices, either. @madsen1998 argue that Fremont subsistence largely varies along two dimensions: levels of residential mobility and levels of cultigen adoption. While some were more mobile, moving from isolated patch to isolated patch, others were more sedentary, typically tethered to a productive wetland or riparian area. As local populations increase, the benefits of increased sedentism and cultigen adoption would slowly begin to outweigh the benefits of a more mobile and foraging-centered diet. According to our model, individuals moving in this direction would tend to concentrate more of their time and energy in the maize farming niche. This would then have the effect of increasing population size within a smaller area, possibly leading to Allee effects, which would invite further sedentism from those in the surrounding area. Following this logic, the patchy adoption of agriculture makes more sense. The degree and speed with which it occurs is simply a function of how fast habitat quality declines and how quickly populations grow.

![
Geographic distribution of Fremont sites across watersheds. The left panel shows the relative density derived from observed site counts, and the right panel shows the relative density from the estimated site counts.
](../figures/model-map.png){#fig-model-map fig-align="center" width="6in"}

## The Collapse of Maize Agriculture

After farming these regions successfully for over a thousand years, the Fremont began an abrupt process of abandonment starting around 700 years BP. The cause of this event is still a matter of some dispute, though a growing consensus appears to be coalescing around climatic changes driving severe reductions in agricultural productivity [@benson2007; @spangler2019; @thomson2019; @thomson2020; @finley2020]. We follow @lindsay1986 in thinking this was likely owing to climatic events that, in effect, tore the Goldilocks zone apart, pulling higher temperatures further down in elevation and pushing higher precipitation levels further up.

Lindsay suggests that the end of the Medieval Warm Period may have been responsible for this decoupling, as it involved higher temperatures and more growing-season precipitation owing to the northward intrusion of summer monsoons. This period lasted from roughly 1,000 to 600 years BP [@grayson2011; @graumlich1993], so it does coincide with the height of the Fremont complex. Given the logic of IFD, this likely led to population spillovers into less suitable habitats [@codding2021]. However, given the extreme circumscription we have highlighted in the Great Basin, the available alternatives would have been limited, potentially leading to increasing population packing across all habitats during this time, thus making the Fremont especially sensitive to climatic downturns.

The well-documented megadrought at the end of the thirteenth century [@cook2004] would have been catastrophic in this context, reducing precipitation levels across the region [@benson2007], but also leading to reductions in water accumulation within the streamflow network. So, just as temperatures are beginning to cool, precipitation levels collapse. By extension, this would have decreased overall water availability and, thus, the potential for irrigation. Opportunities for the Fremont to recover from such a drought would have been limited, too, as the end of the Medieval Warm Period was also the beginning of the Little Ice Age, a time during which conditions were generally colder and wetter [@mann2009], further pulling apart the Goldilocks zone upon which the Fremont so long depended.

### 2023-04-30 manuscript/manuscript.tex
🚀 initial commit

--- 2023-04-30 manuscript/manuscript.qmd
@@
----
-title: The Fremont Frontier
-subtitle: Living at the Margins of Maize Farming
-date: today
-author:
-  - name: Kenneth B. Vernon
-    email: Kenneth.Vernon@colorado.edu
-    affiliations: 
-      - id: colorado
-        address: | 
-          Center for Collaborative Synthesis in Archaeology,  
-          Institute of Behavioral Science,  
-          University of Colorado, Boulder,  
-          1440 15th Street,  
-          Boulder, CO 80302
-    attributes:
-        corresponding: true 
-  - name: Peter M. Yaworsky
-    affiliations:
-      - id: aarhus-culture
-        address: |
-          Department of Culture and Heritage Studies, 
-          School of Culture and Society, 
-          Aarhus University, 
-          Moesgård Allé 20, building 4216, 
-          8270 Højbjerg, Denmark   
-      - id: aarhus-biodiversity
-        address: |
-          Center for Biodiversity Dynamics in a Changing World, 
-          Department of Biology, 
-          Aarhus University, 
-          Ny Munkegade 114-116, 
-          8000 Aarhus C, Denmark
-  - name: Weston McCool
-    affiliations:
-      - id: utah-anthropology
-        address: | 
-          University of Utah Department of Anthropology, 
-          260 S. Central Campus Drive, RM 4625, 
-          Salt Lake City, UT 84112
-  - name: Jerry D. Spangler
-    affiliations:
-      - ref: utah-anthropology
-  - name: Simon Brewer
-    affiliations:
-      - id: utah-geography
-        address: |
-          University of Utah Department of Geography, 
-          260 S. Central Campus Drive, RM 4625, 
-          Salt Lake City, UT 84112
-      - id: gcsc
-        address: |
-          University of Utah Global Change and Sustainability Center, 
-          115 South 1460 East, RM 234 FASB, 
-          Salt Lake City, UT 84112
-  - name: Brian F. Codding
-    affiliations:
-      - ref: utah-anthropology
-      - ref: gcsc
-abstract: |
-  The Fremont provide an important case study to examine the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear.
-Using a tree-ring and simulation-based reconstruction of average annual precipitation and maize growing degree days, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin.
-The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
-keywords:
-  - Ideal Free Distribution
-  - Great Basin
-  - Paleoclimate Reconstruction
-bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Quaternary International
-      formatting: preprint
-      model: 3p
-execute: 
-  echo: false
-  error: false
-  warning: false
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  number,
+  preprint,
+  3p]{elsarticle}
 
-# Introduction
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
-Archaeological populations of subsistence maize farmers known collectively as "the Fremont" lived at the far northern periphery of maize farming in western North America, in an area encompassing much of the modern state of Utah north of the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989; @madsen1998].
-That the Fremont managed to thrive for as long as they did in this part of the world poses an interesting question, for conditions there are extremely harsh and inhospitable for maize farming.
-Archaeologists interested in that question have typically focused on strategies directly related to farming and subsistence, like irrigation methods, trade, and seasonal high elevation hunting [e.g., @barlow2008; @boomgarden2019; @janetski2002; @hart2021; @madsen1998; @metcalfe1985; @patterson2010; @spangler1993; @morgan2012].
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\journal{Quaternary International}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-num}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={The Fremont Frontier},
+  pdfauthor={Kenneth B. Vernon; Peter M. Yaworsky; Weston McCool; Jerry D. Spangler; Simon Brewer; Brian F. Codding},
+  pdfkeywords={Ideal Free Distribution, Great Basin, Paleoclimate
+Reconstruction},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\setlength{\parindent}{6pt}
+\begin{document}
+
+\begin{frontmatter}
+\title{The Fremont Frontier \\\large{Living at the Margins of Maize
+Farming} }
+\author[1]{Kenneth B. Vernon%
+\corref{cor1}%
+}
+ \ead{Kenneth.Vernon@colorado.edu} 
+\author[2,3]{Peter M. Yaworsky%
+%
+}
+
+\author[4]{Weston McCool%
+%
+}
+
+\author[4]{Jerry D. Spangler%
+%
+}
+
+\author[5,6]{Simon Brewer%
+%
+}
+
+\author[4,6]{Brian F. Codding%
+%
+}
+
+\affiliation[1]{organization={},addressline={Center for Collaborative
+Synthesis in Archaeology, Institute of Behavioral Science, University of
+Colorado, Boulder, 1440 15th Street, Boulder, CO 80302},postcodesep={}}
+\affiliation[2]{organization={},addressline={Department of Culture and
+Heritage Studies, School of Culture and Society, Aarhus University,
+Moesgård Allé 20, building 4216, 8270 Højbjerg, Denmark},postcodesep={}}
+\affiliation[3]{organization={},addressline={Center for Biodiversity
+Dynamics in a Changing World, Department of Biology, Aarhus University,
+Ny Munkegade 114-116, 8000 Aarhus C, Denmark},postcodesep={}}
+\affiliation[4]{organization={},addressline={University of Utah
+Department of Anthropology, 260 S. Central Campus Drive, RM 4625, Salt
+Lake City, UT 84112},postcodesep={}}
+\affiliation[5]{organization={},addressline={University of Utah
+Department of Geography, 260 S. Central Campus Drive, RM 4625, Salt Lake
+City, UT 84112},postcodesep={}}
+\affiliation[6]{organization={},addressline={University of Utah Global
+Change and Sustainability Center, 115 South 1460 East, RM 234 FASB, Salt
+Lake City, UT 84112},postcodesep={}}
+
+\cortext[cor1]{Corresponding author}
+
+\begin{abstract}
+The Fremont provide an important case study to examine the resilience of
+ancient farmers to climatic downturns, for they lived at the far
+northern margin of intensive maize agriculture in the American West,
+where the constraints on maize production are made abundantly clear.
+Using a tree-ring and simulation-based reconstruction of average annual
+precipitation and maize growing degree days, along with cost-distance to
+perennial streams, we model spatial variability in Fremont site density
+in the eastern Great Basin. The results of our analysis have
+implications for defining the ecological envelope in which farming is a
+viable strategy across this arid region and can be used to predict where
+and why maize farming strategies might evolve and eventually collapse as
+climate changes over time.
+\end{abstract}
+
+\begin{keyword}
+    Ideal Free Distribution \sep Great Basin \sep 
+    Paleoclimate Reconstruction
+\end{keyword}
+\end{frontmatter}
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Archaeological populations of subsistence maize farmers known
+collectively as ``the Fremont'' lived at the far northern periphery of
+maize farming in western North America, in an area encompassing much of
+the modern state of Utah north of the Virgin and Colorado Rivers, from
+roughly 2000 to 700 years BP \citep{madsen1989, madsen1998}. That the
+Fremont managed to thrive for as long as they did in this part of the
+world poses an interesting question, for conditions there are extremely
+harsh and inhospitable for maize farming. Archaeologists interested in
+that question have typically focused on strategies directly related to
+farming and subsistence, like irrigation methods, trade, and seasonal
+high elevation hunting
+\citep[e.g.,][]{barlow2008, boomgarden2019, janetski2002, hart2021, madsen1998, metcalfe1985, patterson2010, spangler1993, morgan2012}.
 Here, however, we come at the problem from a slightly different angle.
-We seek to understand not so much *what* they did to make farming effective, but *where* they chose to do it, and why they chose those places over others [e.g., @bocinsky2014; @thomson2020].
-Those are not entirely separate questions, of course, as they trade-off each other, but we think a geographic analysis can actually help illuminate some of the costs and benefits of the other strategies adopted by the Fremont.
-A geographic or spatial model will also help us to investigate the ways that climate change might have structured those costs and benefits, and by extension the settlement decisions of Fremont farmers, which we consider a long-term record of human adaptation to climate change.
+We seek to understand not so much \emph{what} they did to make farming
+effective, but \emph{where} they chose to do it, and why they chose
+those places over others \citep[e.g.,][]{bocinsky2014, thomson2020}.
+Those are not entirely separate questions, of course, as they trade-off
+each other, but we think a geographic analysis can actually help
+illuminate some of the costs and benefits of the other strategies
+adopted by the Fremont. A geographic or spatial model will also help us
+to investigate the ways that climate change might have structured those
+costs and benefits, and by extension the settlement decisions of Fremont
+farmers, which we consider a long-term record of human adaptation to
+climate change.
 
-As a test case, we focus on those Fremont living in the eastern Great Basin of western Utah, which we refer to as the western Fremont (see @fig-overview).
-The extreme desert conditions that the western Fremont adapted to provide a useful backdrop for exploring Fremont settlement, for variation in that environment exhibits abrupt and quite dramatic changes over short distances of space and time, thus heightening the differences between areas the western Fremont did and did not occupy.
-We explore those differences from a socio-ecological perspective, focusing on dynamic interactions between competing individuals on the one hand and individuals and their environment on the other [@bird2006; @kennett2006a].
-This dynamic is captured nicely by the Ideal Free Distribution model from Behavioral Ecology, as it represents individual decisions about where to live as a choice between habitats whose benefits to the individual may be constrained by what other individuals are doing.
-The technical term for these benefits is 'suitability', which the IFD defines as some function of both the habitat's intrinsic or pristine environmental quality and its current population size [@fretwell1969; @codding2015; @winterhalder2010].
-The fundamental assumption underlying the IFD is that individuals will behave optimally, that they will choose to settle the habitat with the highest suitability first, meaning the one that offers them the greatest ratio of benefits to costs relative to the available alternatives.
+As a test case, we focus on those Fremont living in the eastern Great
+Basin of western Utah, which we refer to as the western Fremont (see
+Figure~\ref{fig-overview}). The extreme desert conditions that the
+western Fremont adapted to provide a useful backdrop for exploring
+Fremont settlement, for variation in that environment exhibits abrupt
+and quite dramatic changes over short distances of space and time, thus
+heightening the differences between areas the western Fremont did and
+did not occupy. We explore those differences from a socio-ecological
+perspective, focusing on dynamic interactions between competing
+individuals on the one hand and individuals and their environment on the
+other \citep{bird2006, kennett2006a}. This dynamic is captured nicely by
+the Ideal Free Distribution model from Behavioral Ecology, as it
+represents individual decisions about where to live as a choice between
+habitats whose benefits to the individual may be constrained by what
+other individuals are doing. The technical term for these benefits is
+`suitability', which the IFD defines as some function of both the
+habitat's intrinsic or pristine environmental quality and its current
+population size \citep{fretwell1969, codding2015, winterhalder2010}. The
+fundamental assumption underlying the IFD is that individuals will
+behave optimally, that they will choose to settle the habitat with the
+highest suitability first, meaning the one that offers them the greatest
+ratio of benefits to costs relative to the available alternatives.
 
-Assuming that western Fremont individuals were adept at responding to variation in their local ecology, in particular, that they optimized their choice of habitats, we can use an empirical or inductive species distribution model of Fremont sites to reconstruct the ecological niche for maize farming [@vernon2021; @yaworsky2020; @elith2009].
-This is sometimes referred to as ecological niche modeling.
-The basic idea here is that we can look at the distribution of Fremont sites as a reliable indicator of the sorts of available environmental conditions that would best promote maize farming in an arid landscape.
-In effect, we are helping ourselves to a little reverse engineering of the archaeological record [@dennett1995].
-We are assuming that whatever strategy - whatever habitat - the Fremont chose was the best alternative available to them, not in any absolute sense, of course, just the best given the constraints and trade-offs Fremont farmers faced in those environments [@jochim2022].
-In other words, we are treating their settlement decisions themselves as maize farming adaptations.
+Assuming that western Fremont individuals were adept at responding to
+variation in their local ecology, in particular, that they optimized
+their choice of habitats, we can use an empirical or inductive species
+distribution model of Fremont sites to reconstruct the ecological niche
+for maize farming \citep{vernon2021, yaworsky2020, elith2009}. This is
+sometimes referred to as ecological niche modeling. The basic idea here
+is that we can look at the distribution of Fremont sites as a reliable
+indicator of the sorts of available environmental conditions that would
+best promote maize farming in an arid landscape. In effect, we are
+helping ourselves to a little reverse engineering of the archaeological
+record \citep{dennett1995}. We are assuming that whatever strategy -
+whatever habitat - the Fremont chose was the best alternative available
+to them, not in any absolute sense, of course, just the best given the
+constraints and trade-offs Fremont farmers faced in those environments
+\citep{jochim2022}. In other words, we are treating their settlement
+decisions themselves as maize farming adaptations.
 
-![
-On the left is an overview map showing the project area with modern satellite imagery.
-Dark green colors represent high elevation, high precipitation areas.
-The large beige area in the northwest is the Bonneville Basin.
-What remains of the Great Salt Lake can be seen just west of Salt Lake City.
-For visualization purposes, the map on the right shows the relative density of archaeological sites (see the section on site data below for details on how we calculate relative density).
-Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
-](../figures/overview.png){#fig-overview fig-align="center" width="6in"}
+\begin{figure}
 
-# Background
+{\centering \includegraphics[width=6in,height=\textheight]{../figures/overview.png}
 
-## The Fremont
+}
 
-Archaeologists have nominated a number of artifact types and artifact properties as candidate traits to distinguish the Fremont from their neighbors in space and time, including metates with secondary grinding surfaces, large semi-subterranean pithouses, various clay figurines, trapezoidal rock art, one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and elongated corner-notched arrow points [@madsen1998].
-None of these, however, nor any combination of them, has yet proven up to the task of bringing these diverse people under a single, all-encompassing definition [@madsen1998].
-Nevertheless, the Fremont do stand out, especially at their population apex roughly 1000 years BP.
+\caption{\label{fig-overview}On the left is an overview map showing the
+project area with modern satellite imagery. Dark green colors represent
+high elevation, high precipitation areas. The large beige area in the
+northwest is the Bonneville Basin. What remains of the Great Salt Lake
+can be seen just west of Salt Lake City. For visualization purposes, the
+map on the right shows the relative density of archaeological sites (see
+the section on site data below for details on how we calculate relative
+density). Please note that although the map appears to show many empty
+watersheds, most do in fact contain archaeological sites, just at
+extremely low densities.}
 
-Around this time, the Fremont likely reached their greatest geographic extent, inhabiting an area that encompassed most of the modern state of Utah [@janetski2011], with varying levels of support for occupations in Idaho [@dean1992prehistoric], Nevada [@hockett1998], Colorado [@baker1999], and Wyoming [@hakiel1987archery; @smith1992fremont].
-These modern political boundaries roughly overlap with the eastern Great Basin and Colorado Plateau physiographic provinces.
+\end{figure}
 
-![
-Linear responses of average water-year precipitation (mm), average growing-season maize gdd (C), and cost-distance to streams (hours) to changes in elevation (m).
-See methods for how these values were calculated and results for more details.
-](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
-A basin and range topography dominates the landscape of the Great Basin, with large, dry valleys and endorheic basins punctuated by north-south trending mountain ranges.
-Changes in elevation in this setting are quite dramatic, on the order of several thousand meters over a short east-west transect.
-As elevation drives virtually every climatological process in the region, the local ecology is characterized by equally dramatic extremes [@billings1951; @grayson2011].
-As shown in @fig-elevation, net water-year precipitation in the eastern Great Basin tends to increase with elevation, while maize growing degree days (GDD) over the growing-season tends to decrease.
-An important trade-off thus exists between temperature and precipitation, with warmer and drier conditions at lower elevations and cooler and wetter conditions at higher elevations.
+\hypertarget{the-fremont}{%
+\subsection{The Fremont}\label{the-fremont}}
 
-These are the environments into which the Fremont first emerged in western Utah.
-Models for describing their origins, while diverse, tend to fall into the familiar categories of migration and diffusion.
-Where models of the former migration variety insist that Ancestral Puebloan farmers migrated northward from the US Southwest, bringing with them drought-tolerant maize varieties and maize farming [@madsen1975; @kidder1924], models of the latter diffusion variety suggest that the Fremont first emerged within these areas as local Archaic populations transitioned away from mobile foraging, adopting semi-sedentary agricultural practices from the southwest in a piecemeal fashion over several hundred years [@jennings1978; @winter1974].
-While archaeologists have marshaled an impressive array of arguments and counter-arguments for these two classes of models, a general consensus appears to be coalescing around the idea that the truth lies somewhere in between, that the interaction between a nascent Archaic population existing at extremely low population levels and a burgeoning community of migrant farmers together produced the complex we now call the Fremont [@patterson2015; @simms1986; @spangler1993; @spangler2000; @spangler2013].
+Archaeologists have nominated a number of artifact types and artifact
+properties as candidate traits to distinguish the Fremont from their
+neighbors in space and time, including metates with secondary grinding
+surfaces, large semi-subterranean pithouses, various clay figurines,
+trapezoidal rock art, one-rod-and-bundle basketry, painted white and
+corrugated gray ceramics, and elongated corner-notched arrow points
+\citep{madsen1998}. None of these, however, nor any combination of them,
+has yet proven up to the task of bringing these diverse people under a
+single, all-encompassing definition \citep{madsen1998}. Nevertheless,
+the Fremont do stand out, especially at their population apex roughly
+1000 years BP.
 
-Models that try to explain *why* Fremont maize farmers emerged in the far northern periphery of the US Southwest tend to agree - more or less - that at some point farming became a more economical alternative to foraging in the region.
-Where they differ, it is in whether things got better for farming starting roughly 2000 years BP [@benson2011; @coltrain2002; @matson1988] or worse for foraging [@barlow2002; @broughton2010; @cannon2000; @cannon2001].
-In support of the *farming-more-better* model, archaeologists draw on evidence for a suite of favorable climatic conditions that came together around 1600 years BP, including warmer temperatures and increased precipitation (both winter and summer), as well as the expansion of grasslands [@grayson2006; @hemphill1995; @rhode2000].
-The greatest expansion of the Fremont, in fact, occurred around 1000 years BP, around the time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures that would have increased the intensity of summer monsoons in the southwest, pushing them further north into the Great Basin and onto the Colorado Plateau [@grayson2006], a pattern reflected in numerous tree ring records [e.g., @graybill1990; @leavitt1994; @stine1990; @salzer2014], though some complications do exist for this suggested pattern [@hart2021].
+Around this time, the Fremont likely reached their greatest geographic
+extent, inhabiting an area that encompassed most of the modern state of
+Utah \citep{janetski2011}, with varying levels of support for
+occupations in Idaho \citep{dean1992prehistoric}, Nevada
+\citep{hockett1998}, Colorado \citep{baker1999}, and Wyoming
+\citep{hakiel1987archery, smith1992fremont}. These modern political
+boundaries roughly overlap with the eastern Great Basin and Colorado
+Plateau physiographic provinces.
 
-In response to these sorts of arguments, those who favor the *foraging-more-worse* model [@hart2021; @barlow2002] argue that conditions promoting maize farming would also tend to make foraging a more productive strategy.
-Plus, farming is an extremely costly endeavor, especially when compared to foraging, including lots of upfront investment of time and energy, so improving conditions favorable to maize agriculture are not by themselves sufficient to explain its origins.
-To fill that lacuna, archaeologists point to accumulating evidence suggesting that the Fremont transition to agriculture accompanied periods of sustained population growth [@barlow2008; @codding2021; @spangler2013] that likely resulted in the depressed availability of wild resources [@barlow2002; @janetski1997; @simms1986].
-So, demographic pressure reduced the efficiency of foraging strategies, thus making maize agriculture the optimal strategy.
-And that, in turn, led the Fremont to adopt farming as their primary mode of subsistence.
+\begin{figure}
 
-Once they entered the farming niche, the Fremont quickly developed several adaptations for coping with the marginal and stochastic conditions in the eastern Great Basin and the Colorado Plateau, the most important almost certainly being irrigation (for more on this, see the discussion).
-On the Colorado Plateau, the eastern Fremont settled deep within the steep canyons and washes that flanked the tributaries of major rivers like the Green and Yampa [@yaworsky2021; see also, Yaworsky this issue], in places like Nine Mile Canyon [@spangler1993] and Range Creek Canyon [@hart2021].
-In the eastern Great Basin, the western Fremont chose to settle on open, alluvial fans and stream terraces along the lower slopes of most ranges [@janetski2000; @madsen1998; @simms2008], at places like Nephi Mounds [@sharrock1967], Pharo Village [@marwitt1968], Median Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock Village [@aikens1967].
+{\centering \includegraphics[width=4in,height=\textheight]{../figures/elevation-everything.png}
 
-Given the extreme aridity of the west, they were still susceptible to drought, of course, even in those more suitable areas, with a single short growing season making agricultural returns highly variable.
-To cope with these uncertainties, Fremont farmers may have adopted additional risk-mitigation strategies like spreading farm plots among different micro-environments, a technique otherwise known as plot diversification [@spangler1993; @spangler2013; @patterson2015].
-The Fremont also relied heavily on crop storage both to provide a surplus during the winter season and to make-up for poor yields [@spangler2019].
+}
 
-## The Ideal Free Distribution
+\caption{\label{fig-elevation}Linear responses of average water-year
+precipitation (mm), average growing-season maize gdd (C), and
+cost-distance to streams (hours) to changes in elevation (m). See
+methods for how these values were calculated and results for more
+details.}
 
-In population ecology, settlement decisions are represented as a choice between habitats, where each habitat $i$ has some net biological value or benefit, known as its suitability ($S_i$).
-Typically, though not always, this value exhibits negative density dependence, declining as a function of population size ($N_i$), perhaps owing to increased competition or resource depression.
-When $N_i$ is very small, the suitability of a habitat approaches its pristine or intrinsic quality ($Q_i$), or its gross value independent of all settlement costs [@greene2001].
-Together, these parameters suggest a very simple model of habitat suitability:
+\end{figure}
 
-$$
+A basin and range topography dominates the landscape of the Great Basin,
+with large, dry valleys and endorheic basins punctuated by north-south
+trending mountain ranges. Changes in elevation in this setting are quite
+dramatic, on the order of several thousand meters over a short east-west
+transect. As elevation drives virtually every climatological process in
+the region, the local ecology is characterized by equally dramatic
+extremes \citep{billings1951, grayson2011}. As shown in
+Figure~\ref{fig-elevation}, net water-year precipitation in the eastern
+Great Basin tends to increase with elevation, while maize growing degree
+days (GDD) over the growing-season tends to decrease. An important
+trade-off thus exists between temperature and precipitation, with warmer
+and drier conditions at lower elevations and cooler and wetter
+conditions at higher elevations.
+
+These are the environments into which the Fremont first emerged in
+western Utah. Models for describing their origins, while diverse, tend
+to fall into the familiar categories of migration and diffusion. Where
+models of the former migration variety insist that Ancestral Puebloan
+farmers migrated northward from the US Southwest, bringing with them
+drought-tolerant maize varieties and maize farming
+\citep{madsen1975, kidder1924}, models of the latter diffusion variety
+suggest that the Fremont first emerged within these areas as local
+Archaic populations transitioned away from mobile foraging, adopting
+semi-sedentary agricultural practices from the southwest in a piecemeal
+fashion over several hundred years \citep{jennings1978, winter1974}.
+While archaeologists have marshaled an impressive array of arguments and
+counter-arguments for these two classes of models, a general consensus
+appears to be coalescing around the idea that the truth lies somewhere
+in between, that the interaction between a nascent Archaic population
+existing at extremely low population levels and a burgeoning community
+of migrant farmers together produced the complex we now call the Fremont
+\citep{patterson2015, simms1986, spangler1993, spangler2000, spangler2013}.
+
+Models that try to explain \emph{why} Fremont maize farmers emerged in
+the far northern periphery of the US Southwest tend to agree - more or
+less - that at some point farming became a more economical alternative
+to foraging in the region. Where they differ, it is in whether things
+got better for farming starting roughly 2000 years BP
+\citep{benson2011, coltrain2002, matson1988} or worse for foraging
+\citep{barlow2002, broughton2010, cannon2000, cannon2001}. In support of
+the \emph{farming-more-better} model, archaeologists draw on evidence
+for a suite of favorable climatic conditions that came together around
+1600 years BP, including warmer temperatures and increased precipitation
+(both winter and summer), as well as the expansion of grasslands
+\citep{grayson2006, hemphill1995, rhode2000}. The greatest expansion of
+the Fremont, in fact, occurred around 1000 years BP, around the time the
+Medieval Climate Anomaly (MCA) introduced warmer summer temperatures
+that would have increased the intensity of summer monsoons in the
+southwest, pushing them further north into the Great Basin and onto the
+Colorado Plateau \citep{grayson2006}, a pattern reflected in numerous
+tree ring records
+\citep[e.g.,][]{graybill1990, leavitt1994, stine1990, salzer2014},
+though some complications do exist for this suggested pattern
+\citep{hart2021}.
+
+In response to these sorts of arguments, those who favor the
+\emph{foraging-more-worse} model \citep{hart2021, barlow2002} argue that
+conditions promoting maize farming would also tend to make foraging a
+more productive strategy. Plus, farming is an extremely costly endeavor,
+especially when compared to foraging, including lots of upfront
+investment of time and energy, so improving conditions favorable to
+maize agriculture are not by themselves sufficient to explain its
+origins. To fill that lacuna, archaeologists point to accumulating
+evidence suggesting that the Fremont transition to agriculture
+accompanied periods of sustained population growth
+\citep{barlow2008, codding2021, spangler2013} that likely resulted in
+the depressed availability of wild resources
+\citep{barlow2002, janetski1997, simms1986}. So, demographic pressure
+reduced the efficiency of foraging strategies, thus making maize
+agriculture the optimal strategy. And that, in turn, led the Fremont to
+adopt farming as their primary mode of subsistence.
+
+Once they entered the farming niche, the Fremont quickly developed
+several adaptations for coping with the marginal and stochastic
+conditions in the eastern Great Basin and the Colorado Plateau, the most
+important almost certainly being irrigation (for more on this, see the
+discussion). On the Colorado Plateau, the eastern Fremont settled deep
+within the steep canyons and washes that flanked the tributaries of
+major rivers like the Green and Yampa {[}\citep{yaworsky2021}; see also,
+Yaworsky this issue{]}, in places like Nine Mile Canyon
+\citep{spangler1993} and Range Creek Canyon \citep{hart2021}. In the
+eastern Great Basin, the western Fremont chose to settle on open,
+alluvial fans and stream terraces along the lower slopes of most ranges
+\citep{janetski2000, madsen1998, simms2008}, at places like Nephi Mounds
+\citep{sharrock1967}, Pharo Village \citep{marwitt1968}, Median Village
+\citep{marwitt1970}, Evans Mound \citep{berry1972, dodd1982}, and Snake
+Rock Village \citep{aikens1967}.
+
+Given the extreme aridity of the west, they were still susceptible to
+drought, of course, even in those more suitable areas, with a single
+short growing season making agricultural returns highly variable. To
+cope with these uncertainties, Fremont farmers may have adopted
+additional risk-mitigation strategies like spreading farm plots among
+different micro-environments, a technique otherwise known as plot
+diversification \citep{spangler1993, spangler2013, patterson2015}. The
+Fremont also relied heavily on crop storage both to provide a surplus
+during the winter season and to make-up for poor yields
+\citep{spangler2019}.
+
+\hypertarget{the-ideal-free-distribution}{%
+\subsection{The Ideal Free
+Distribution}\label{the-ideal-free-distribution}}
+
+In population ecology, settlement decisions are represented as a choice
+between habitats, where each habitat \(i\) has some net biological value
+or benefit, known as its suitability (\(S_i\)). Typically, though not
+always, this value exhibits negative density dependence, declining as a
+function of population size (\(N_i\)), perhaps owing to increased
+competition or resource depression. When \(N_i\) is very small, the
+suitability of a habitat approaches its pristine or intrinsic quality
+(\(Q_i\)), or its gross value independent of all settlement costs
+\citep{greene2001}. Together, these parameters suggest a very simple
+model of habitat suitability:
+
+\[
 S_i = Q_i - f(N_i)
-$$
+\]
 
-This is known as the Ideal Free Distribution (IFD) model [@fretwell1969].
-The IFD model assumes that individuals will choose habitats that maximize $S_i$, settling the most suitable habitat first then shifting to the next most suitable habitat when those become equal, and thereafter infilling each at an equal rate.
-The aggregate effect is a population distributed at equilibrium, meaning suitability is equal for all habitats.
-When conditions are ideal (individuals have perfect knowledge) and free (individuals face no additional costs to settlement), this distribution will also satisfy the "input matching" rule, with populations occurring at larger numbers in habitats with greater intrinsic value [@parker1978].
+This is known as the Ideal Free Distribution (IFD) model
+\citep{fretwell1969}. The IFD model assumes that individuals will choose
+habitats that maximize \(S_i\), settling the most suitable habitat first
+then shifting to the next most suitable habitat when those become equal,
+and thereafter infilling each at an equal rate. The aggregate effect is
+a population distributed at equilibrium, meaning suitability is equal
+for all habitats. When conditions are ideal (individuals have perfect
+knowledge) and free (individuals face no additional costs to
+settlement), this distribution will also satisfy the ``input matching''
+rule, with populations occurring at larger numbers in habitats with
+greater intrinsic value \citep{parker1978}.
 
-Two circumstances provide important exceptions to this simple model.
-The first concerns economies of scale, or situations in which $S_i$ exhibits positive density dependence.
-In ecology, these are known as Allee effects, and they are assumed to occur at low density [@allee1949; @fretwell1969].
-The second concerns additional settlement costs, or costs associated with moving from one habitat to another.
-This is a violation of the free element of the IFD.
-When it occurs, the result is an ideal despotic distribution (IDD) [@fretwell1969].
-This has the consequence of reducing the total population in the more suitable habitat at equilibrium.
+Two circumstances provide important exceptions to this simple model. The
+first concerns economies of scale, or situations in which \(S_i\)
+exhibits positive density dependence. In ecology, these are known as
+Allee effects, and they are assumed to occur at low density
+\citep{allee1949, fretwell1969}. The second concerns additional
+settlement costs, or costs associated with moving from one habitat to
+another. This is a violation of the free element of the IFD. When it
+occurs, the result is an ideal despotic distribution (IDD)
+\citep{fretwell1969}. This has the consequence of reducing the total
+population in the more suitable habitat at equilibrium.
 
-Many social and ecological variables may be relevant to defining a general measure of $S_i$, but here we are focusing on ecological determinants of intrinsic habitat quality, $Q_i$, that are specific to maize farmers.
-Our reasoning here is fairly straightforward.
-If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters.
-That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium.
-So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.
+Many social and ecological variables may be relevant to defining a
+general measure of \(S_i\), but here we are focusing on ecological
+determinants of intrinsic habitat quality, \(Q_i\), that are specific to
+maize farmers. Our reasoning here is fairly straightforward. If we
+assume that the distribution of Fremont populations across habitats is
+at equilibrium, then we can combine this concept of maize-specific
+suitability with the input matching rule to reconstruct the ranking of
+habitats on the basis of ecologically-relevant parameters. That is, it
+follows from their being at equilibrium that any observed differences in
+population size must be the result of differences in \(Q_i\), that
+\(Q_i \approx f(N_i)\) at equilibrium. So, if we have some proxy for
+\(N_i\), like the distribution of Fremont sites across habitats, we can
+model that value as a function of ecological covariates \(X_i\), then
+use the resulting model to reconstruct \(Q_i \approx f(X_i)\), giving us
+the maize niche for Fremont farmers, both in geographic and ecological
+space.
 
-# Materials and Methods
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
 
-## The Project Area
+\hypertarget{the-project-area}{%
+\subsection{The Project Area}\label{the-project-area}}
 
-Because we are focusing on the western Fremont, we constrain the current project area to the eastern Great Basin in western Utah, as shown in @fig-overview, where the Great Basin is defined in the hydrologic sense using watershed boundaries [@grayson2011].
-On the far eastern edge of this region, a series of north-south ranges, including the Wasatch, Pahvant, and Tushar Mountains, together stand as the primary physiographic boundaries separating the Great Basin from the Colorado Plateau.
-On its far western edge, a series of north-south ranges straddle the Nevada state line or lie just inside it, including the Snake Range where Great Basin National Park is located.
-Its northern periphery is the Great Salt Lake and the West Desert, its southern boundary the Pine Valley Mountains that separate the Great Basin from the Virgin River watershed, just north of St.
-George, UT.
-This is a region approximately 107,000 km$^2$ in area, encompassing virtually all of the Bonneville Basin in western Utah.
+Because we are focusing on the western Fremont, we constrain the current
+project area to the eastern Great Basin in western Utah, as shown in
+Figure~\ref{fig-overview}, where the Great Basin is defined in the
+hydrologic sense using watershed boundaries \citep{grayson2011}. On the
+far eastern edge of this region, a series of north-south ranges,
+including the Wasatch, Pahvant, and Tushar Mountains, together stand as
+the primary physiographic boundaries separating the Great Basin from the
+Colorado Plateau. On its far western edge, a series of north-south
+ranges straddle the Nevada state line or lie just inside it, including
+the Snake Range where Great Basin National Park is located. Its northern
+periphery is the Great Salt Lake and the West Desert, its southern
+boundary the Pine Valley Mountains that separate the Great Basin from
+the Virgin River watershed, just north of St.~George, UT. This is a
+region approximately 107,000 km\(^2\) in area, encompassing virtually
+all of the Bonneville Basin in western Utah.
 
-These boundaries also serve as geographic borders between the western Fremont and other contemporaneous populations.
-Around the Great Salt Lake and the Utah-Nevada border, Fremont farming gives way to Archaic populations engaged in intensive foraging modes of subsistence.
-The Wasatch Range and Tushar Mountains separate the western from the eastern Fremont on the Colorado Plateau, and the Pine Valley Mountains serve as a border between the Fremont and the Virgin Ancestral Puebloan to the south [@simms2008].
+These boundaries also serve as geographic borders between the western
+Fremont and other contemporaneous populations. Around the Great Salt
+Lake and the Utah-Nevada border, Fremont farming gives way to Archaic
+populations engaged in intensive foraging modes of subsistence. The
+Wasatch Range and Tushar Mountains separate the western from the eastern
+Fremont on the Colorado Plateau, and the Pine Valley Mountains serve as
+a border between the Fremont and the Virgin Ancestral Puebloan to the
+south \citep{simms2008}.
 
-## The Unit of Analysis
+\hypertarget{the-unit-of-analysis}{%
+\subsection{The Unit of Analysis}\label{the-unit-of-analysis}}
 
-The unit of analysis for this research is the HUC10 watershed (n = 183), as defined by the Watershed Boundary Dataset (WBD) developed by the US Geological Survey and the Natural Resources Conservation Service within the US Department of Agriculture [@usgs2013].
-These provide a spatially-explicit proxy for specifically human habitats in the proposed study area.
-There are two reasons for this.
-First, the topography of the region is such that the cost of travel between watersheds is often quite significant.
-As a consequence, individuals are expected to spend more time traveling within watersheds than between them, all else being equal.
-Second, watersheds are by definition water sinks, funneling all available run-off into their respective stream networks.
-Given the extreme aridity of the proposed project area, this also makes them critical resource sinks, especially for maize farmers, as watersheds determine how much water might accumulate at a location, as well as its potential for irrigation.
+The unit of analysis for this research is the HUC10 watershed (n = 183),
+as defined by the Watershed Boundary Dataset (WBD) developed by the US
+Geological Survey and the Natural Resources Conservation Service within
+the US Department of Agriculture \citep{usgs2013}. These provide a
+spatially-explicit proxy for specifically human habitats in the proposed
+study area. There are two reasons for this. First, the topography of the
+region is such that the cost of travel between watersheds is often quite
+significant. As a consequence, individuals are expected to spend more
+time traveling within watersheds than between them, all else being
+equal. Second, watersheds are by definition water sinks, funneling all
+available run-off into their respective stream networks. Given the
+extreme aridity of the proposed project area, this also makes them
+critical resource sinks, especially for maize farmers, as watersheds
+determine how much water might accumulate at a location, as well as its
+potential for irrigation.
 
-Aggregating to the level of the watershed also reduces the computational burdens of this modeling exercise.
-In particular, it makes paleoclimate reconstructions more tractable, as we are only estimating the means within each watershed through time (see below for details).
+Aggregating to the level of the watershed also reduces the computational
+burdens of this modeling exercise. In particular, it makes paleoclimate
+reconstructions more tractable, as we are only estimating the means
+within each watershed through time (see below for details).
 
-## Site Data
+\hypertarget{site-data}{%
+\subsection{Site Data}\label{site-data}}
 
-To identify Fremont sites in the project area, we rely on site records and cultural resource reports hosted by the cultural resource information systems in Nevada (NVCRIS) and Utah (Sego) with permission from the respective State Historic Preservation Offices.
-The Sego system also includes an attribute table with affiliation data that we use to filter Fremont sites where possible.
-For a previous project, we also performed a stratified random sample of all archaeological sites by county in western Utah and recorded affiliation and other attribute data from their site forms.
-For this analysis, we compare the results of that previous effort with the sites identified by filtering the Sego system as a form of quality control.
-The resulting dataset contains 2,248 individual Fremont sites.
+To identify Fremont sites in the project area, we rely on site records
+and cultural resource reports hosted by the cultural resource
+information systems in Nevada (NVCRIS) and Utah (Sego) with permission
+from the respective State Historic Preservation Offices. The Sego system
+also includes an attribute table with affiliation data that we use to
+filter Fremont sites where possible. For a previous project, we also
+performed a stratified random sample of all archaeological sites by
+county in western Utah and recorded affiliation and other attribute data
+from their site forms. For this analysis, we compare the results of that
+previous effort with the sites identified by filtering the Sego system
+as a form of quality control. The resulting dataset contains 2,248
+individual Fremont sites.
 
-These site data include everything from small ceramic scatters to massive Fremont villages like Five Finger Ridge [@janetski1998; @janetski2000].
-For this analysis, we assume that these reflect different levels of population size and settlement intensity.
-All else being equal, scatters should follow from shorter stays by fewer people, villages from longer stays by more people.
-This is important because population size correlates with a habitat's total suitability according to the IFD model, and our attempts to reconstruct the ecological and geographic borders of the Fremont rely heavily on the input-matching rule, or the idea that habitats with greater intrinsic quality will have larger populations at equilibrium.
-We, therefore, weight the count of sites in each watershed by the total number of residential features that they contain (+1 for sites with no features).
-So, a site with no residential features counts as one site, a site consisting of, say, four pithouses counts as four unique sites, and a site containing a roomblock with seven rooms counts as seven sites.
-This brings the total weighted site count to 2,447, with minimum and maximum values of 0 and 119, respectively.
+These site data include everything from small ceramic scatters to
+massive Fremont villages like Five Finger Ridge
+\citep{janetski1998, janetski2000}. For this analysis, we assume that
+these reflect different levels of population size and settlement
+intensity. All else being equal, scatters should follow from shorter
+stays by fewer people, villages from longer stays by more people. This
+is important because population size correlates with a habitat's total
+suitability according to the IFD model, and our attempts to reconstruct
+the ecological and geographic borders of the Fremont rely heavily on the
+input-matching rule, or the idea that habitats with greater intrinsic
+quality will have larger populations at equilibrium. We, therefore,
+weight the count of sites in each watershed by the total number of
+residential features that they contain (+1 for sites with no features).
+So, a site with no residential features counts as one site, a site
+consisting of, say, four pithouses counts as four unique sites, and a
+site containing a roomblock with seven rooms counts as seven sites. This
+brings the total weighted site count to 2,447, with minimum and maximum
+values of 0 and 119, respectively.
 
-@fig-overview shows the relative density of Fremont sites across watersheds.
-The *relative* density refers to the density of feature-weighted sites ($D_i = N_i/area_i$) divided by the maximum density ($max\;D$) across all watersheds, thus scaling the density estimates to the unit interval [0,1].
-Using the relative density is just a way of avoiding any implication that we have an independent estimate of the absolute number of Fremont sites in the area, and not just an estimate based on a thinned sample.
+Figure~\ref{fig-overview} shows the relative density of Fremont sites
+across watersheds. The \emph{relative} density refers to the density of
+feature-weighted sites (\(D_i = N_i/area_i\)) divided by the maximum
+density (\(max\;D\)) across all watersheds, thus scaling the density
+estimates to the unit interval {[}0,1{]}. Using the relative density is
+just a way of avoiding any implication that we have an independent
+estimate of the absolute number of Fremont sites in the area, and not
+just an estimate based on a thinned sample.
 
-## Environmental Covariates
+\hypertarget{environmental-covariates}{%
+\subsection{Environmental Covariates}\label{environmental-covariates}}
 
-For this analysis, we use both topographic and climatological covariates.
-The topographic covariate is cost-distance to perennial streams (measured in hours).
-This provides a coarse grained estimate of water availability, as well as the potential costs of irrigation.
-To calculate this, we first use the R package FedData [@bocinsky2020] to download perennial stream features from the US National Hydrography Dataset [@usgs2022] and a digital elevation model (DEM) from the US National Elevation Dataset [@usgs2022a].
-We then apply Campbell's hiking function [@campbell2019] to slope estimates derived from the DEM using the R package hiker [@vernon2021a].
-This allows us to generate estimates of travel time between grid cells in the DEM and then calculate the accumulated cost of travel from each perennial stream to any grid cell within each watershed.
-We then aggregate these values to the watershed level by taking their mean.
+For this analysis, we use both topographic and climatological
+covariates. The topographic covariate is cost-distance to perennial
+streams (measured in hours). This provides a coarse grained estimate of
+water availability, as well as the potential costs of irrigation. To
+calculate this, we first use the R package FedData \citep{bocinsky2020}
+to download perennial stream features from the US National Hydrography
+Dataset \citep{usgs2022} and a digital elevation model (DEM) from the US
+National Elevation Dataset \citep{usgs2022a}. We then apply Campbell's
+hiking function \citep{campbell2019} to slope estimates derived from the
+DEM using the R package hiker \citep{vernon2021a}. This allows us to
+generate estimates of travel time between grid cells in the DEM and then
+calculate the accumulated cost of travel from each perennial stream to
+any grid cell within each watershed. We then aggregate these values to
+the watershed level by taking their mean.
 
-Climate covariates include net water-year precipitation (precipitation, in millimeters, mm) from October to September and maize growing degree days (GDD, in Celsius, C) over the growing season from May to September, as these have a large effect on maize productivity, thus providing important constraints on the patterns of settlement for those who relied on maize for subsistence.
-We caution that GDD is calculated in such a way as to be insensitive to extreme high and low temperatures (extreme heat and frost, in effect).
-It is also conceptually understood to be a measure of accumulated temperature between the last and first frost-free days, but in our analysis is calculated using hard start and end dates of May 1 and September 30, under the assumption that these dates will roughly coincide with those frost-free days.
+Climate covariates include net water-year precipitation (precipitation,
+in millimeters, mm) from October to September and maize growing degree
+days (GDD, in Celsius, C) over the growing season from May to September,
+as these have a large effect on maize productivity, thus providing
+important constraints on the patterns of settlement for those who relied
+on maize for subsistence. We caution that GDD is calculated in such a
+way as to be insensitive to extreme high and low temperatures (extreme
+heat and frost, in effect). It is also conceptually understood to be a
+measure of accumulated temperature between the last and first frost-free
+days, but in our analysis is calculated using hard start and end dates
+of May 1 and September 30, under the assumption that these dates will
+roughly coincide with those frost-free days.
 
-![
-Distribution of covariates across watersheds. These include maize growing degree days (GDD), annual precipitation (precipitation), and cost-distance to streams (Streams).
-](../figures/covariates.png){#fig-covariates fig-align="center" width="6in"}
+\begin{figure}
 
-We hindcast these climate covariates for each watershed and year over the Fremont sequence from 1550 to 550 years BP (400 to 1400 CE) using the Correlation Adjusted corRelation (CAR) method implemented in the R package paleocar [@bocinsky2019].
-This method, in effect, regresses modern climate data against the width of younger tree rings, then uses older rings to predict past climate trends [for further details, see @bocinsky2014; @bocinsky2016].
-In this case, we use interpolated precipitation and temperature estimates from the \~800 m resolution PRISM dataset [@prismclimategroup2014] and all tree-ring samples in North America, which were obtained from the International Tree Ring Database using FedData.
-For each watershed, we took the median values of precipitation and GDD over the Fremont sequence.
+{\centering \includegraphics[width=6in,height=\textheight]{../figures/covariates.png}
 
-@fig-covariates shows the geographic distributions of these covariates over watersheds.
-As these are all in one way or another driven by elevation, it is reasonable to expect that they will correlate.
-As part of the exploratory phase of this analysis, we, therefore, perform a series of pair-wise tests of correlation measured using Pearson's R.
-We also regress the scaled and centered values (or z-scores) of maize GDD, precipitation, and cost-distance to streams on elevation using ordinary least squares (OLS).
-No doubt, this violates many assumptions of OLS, particularly the identity and independence of the errors, but explanation was not our main goal with these simple linear models.
-We are only trying to get a rough sense of their general response to changes in elevation, which OLS offers (see @fig-elevation).
+}
 
-Given that our primary covariates all correlate with elevation (reported below), readers may wonder why we do not simply use elevation in our main analysis.
-The reason for this is two-fold.
-First, using elevation alone would obscure the trade-off between water availability and temperature.
-Second, elevation is not explanatory of settlement behavior.
-A maize farmer does not choose a location for farming *because* it is at, say, 1200 m above sea level.
-They choose that location *because* it has the right combination of water availability and temperature, that is to say, because conditions there are optimal for maize farming.
-Were those conditions found at some other elevation, the farmer would be expected to farm there.
+\caption{\label{fig-covariates}Distribution of covariates across
+watersheds. These include maize growing degree days (GDD), annual
+precipitation (precipitation), and cost-distance to streams (Streams).}
 
-## Statistical Modeling and Model Evaluation
+\end{figure}
 
-The response variable is a count $N$ of sites per watershed $i$, so it should arise as the result of a Poisson process.
-Given the trade-offs outlined above, we also expect it to exhibit a non-linear response to precipitation, GDD, and cost-distance to streams, so we fit a generalized additive model (GAM) with a log-link to the data using the R package mgcv [@wood2004].
-Although we expect the outcome to be Poisson distributed, we use a negative binomial distribution here, as this includes a parameter, $r$, that can account for potential dispersion without giving up the inferential benefits of maximum likelihood.
+We hindcast these climate covariates for each watershed and year over
+the Fremont sequence from 1550 to 550 years BP (400 to 1400 CE) using
+the Correlation Adjusted corRelation (CAR) method implemented in the R
+package paleocar \citep{bocinsky2019}. This method, in effect, regresses
+modern climate data against the width of younger tree rings, then uses
+older rings to predict past climate trends \citep[for further details,
+see][]{bocinsky2014, bocinsky2016}. In this case, we use interpolated
+precipitation and temperature estimates from the \textasciitilde800 m
+resolution PRISM dataset \citep{prismclimategroup2014} and all tree-ring
+samples in North America, which were obtained from the International
+Tree Ring Database using FedData. For each watershed, we took the median
+values of precipitation and GDD over the Fremont sequence.
 
-$$
+Figure~\ref{fig-covariates} shows the geographic distributions of these
+covariates over watersheds. As these are all in one way or another
+driven by elevation, it is reasonable to expect that they will
+correlate. As part of the exploratory phase of this analysis, we,
+therefore, perform a series of pair-wise tests of correlation measured
+using Pearson's R. We also regress the scaled and centered values (or
+z-scores) of maize GDD, precipitation, and cost-distance to streams on
+elevation using ordinary least squares (OLS). No doubt, this violates
+many assumptions of OLS, particularly the identity and independence of
+the errors, but explanation was not our main goal with these simple
+linear models. We are only trying to get a rough sense of their general
+response to changes in elevation, which OLS offers (see
+Figure~\ref{fig-elevation}).
+
+Given that our primary covariates all correlate with elevation (reported
+below), readers may wonder why we do not simply use elevation in our
+main analysis. The reason for this is two-fold. First, using elevation
+alone would obscure the trade-off between water availability and
+temperature. Second, elevation is not explanatory of settlement
+behavior. A maize farmer does not choose a location for farming
+\emph{because} it is at, say, 1200 m above sea level. They choose that
+location \emph{because} it has the right combination of water
+availability and temperature, that is to say, because conditions there
+are optimal for maize farming. Were those conditions found at some other
+elevation, the farmer would be expected to farm there.
+
+\hypertarget{statistical-modeling-and-model-evaluation}{%
+\subsection{Statistical Modeling and Model
+Evaluation}\label{statistical-modeling-and-model-evaluation}}
+
+The response variable is a count \(N\) of sites per watershed \(i\), so
+it should arise as the result of a Poisson process. Given the trade-offs
+outlined above, we also expect it to exhibit a non-linear response to
+precipitation, GDD, and cost-distance to streams, so we fit a
+generalized additive model (GAM) with a log-link to the data using the R
+package mgcv \citep{wood2004}. Although we expect the outcome to be
+Poisson distributed, we use a negative binomial distribution here, as
+this includes a parameter, \(r\), that can account for potential
+dispersion without giving up the inferential benefits of maximum
+likelihood.
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
+\]
 
-Importantly, we add to the model specification a constant offset for the log of the area of each watershed, in effect making this a model of population *density*.
-This is meant to address the idea that larger watersheds will have more sites just as a matter of chance.
-While our model as currently specified includes the intercept, we caution that such an estimate is not strictly meaningful for this kind of data, as it assumes that the absolute site counts can be inferred from the sample, which requires that the sample be unbiased - an implausible condition to meet.
-On that note, we also include a parametric term for the proportion of each watershed classified as federally protected land.
-This is meant to address potential sampling biases owing to taphonomic processes operating on the archaeological record.
-Notably, for these data, those are almost entirely the result of Euro-American colonization of the area, in particular, the impacts of modern farming techniques and the development of the Wasatch Front as the urban core of the modern state of Utah.
-This is a coarse metric, but our expectation is that the greater the share of a watershed with federal protections, the smaller the effect of these anthropogenic impacts, the more sites we should observe.
+Importantly, we add to the model specification a constant offset for the
+log of the area of each watershed, in effect making this a model of
+population \emph{density}. This is meant to address the idea that larger
+watersheds will have more sites just as a matter of chance. While our
+model as currently specified includes the intercept, we caution that
+such an estimate is not strictly meaningful for this kind of data, as it
+assumes that the absolute site counts can be inferred from the sample,
+which requires that the sample be unbiased - an implausible condition to
+meet. On that note, we also include a parametric term for the proportion
+of each watershed classified as federally protected land. This is meant
+to address potential sampling biases owing to taphonomic processes
+operating on the archaeological record. Notably, for these data, those
+are almost entirely the result of Euro-American colonization of the
+area, in particular, the impacts of modern farming techniques and the
+development of the Wasatch Front as the urban core of the modern state
+of Utah. This is a coarse metric, but our expectation is that the
+greater the share of a watershed with federal protections, the smaller
+the effect of these anthropogenic impacts, the more sites we should
+observe.
 
-Model evaluation includes checks for concurvity or non-linear correlation in the smooth terms of the GAM, as well as a Variance Inflation Factor test on the parametric or linear terms.
-We also test for spatial autocorrelation in the untransformed residuals using Monte Carlo simulations of Moran's I and incorporate an exponential covariance matrix to remove that residual autocorrelation.
-Several smooth terms had effective degrees of freedom (EDF) equal to one, suggesting no non-linear response in the data.
-We, therefore, remove the smooth terms for those covariates in the final model, as expressed in the formula above.
+Model evaluation includes checks for concurvity or non-linear
+correlation in the smooth terms of the GAM, as well as a Variance
+Inflation Factor test on the parametric or linear terms. We also test
+for spatial autocorrelation in the untransformed residuals using Monte
+Carlo simulations of Moran's I and incorporate an exponential covariance
+matrix to remove that residual autocorrelation. Several smooth terms had
+effective degrees of freedom (EDF) equal to one, suggesting no
+non-linear response in the data. We, therefore, remove the smooth terms
+for those covariates in the final model, as expressed in the formula
+above.
 
-All analyses are conducted in the R programming language and environment [@rcoreteam2022]. For details of this analysis, please see the Supplement.
+All analyses are conducted in the R programming language and environment
+\citep{rcoreteam2022}. For details of this analysis, please see the
+Supplement.
 
-# Results
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-```{r}
-#| label: tbl-results
-#| tbl-cap: Model Results
-#| tbl-subcap:
-#|   - "Parametric Terms"
-#|   - "Smooth Terms"
-#| layout-nrow: 2
-#| layout: "[[1], [-1], [1]]"
-#| out-width: 50%
+\begin{table}
 
-# returns parametric and smooth regression tables
-# and q, a list with character strings for reporting estimates in text
-here::here("R", "regression-table.R") |> source()
+\caption{\label{tbl-results}Model
+Results}\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-results-1}Parametric Terms }
 
-parametric_table
-smooth_table
+{\centering
 
-```
+\setlength{\LTpost}{0mm}
+\begin{longtable}{lcccc}
+\tabularnewline
 
-As shown in @tbl-results, all linear (or parametric) coefficients are significant in the final model.
-The intercept `r q[["Intercept"]]` is close to zero when transformed back onto the response scale, likely reflecting the large number of watersheds with zero Fremont sites, though as mentioned above, the intercept estimate is not strictly meaningful for these data.
-The proportion of watersheds falling under federal protections `r q[["Protected"]]` has a positive effect on site counts, thus confirming our expectation that more sites occur in watersheds with a larger proportion of federal land.
-Site counts increase in watersheds with larger GDD values, though the effect is small `r q[["Maize GDD"]]`, and site counts decrease with distance from perennial streams, though again the effect is small `r q[["CD to Streams"]]`.
-The only smooth term in the final model is average precipitation over the water-year, which is significant `r q[["s(Precipitation)"]]`.
-The effective degrees of freedom (EDF) for precipitation suggests a strong non-linear effect of that covariate on site counts, which increases up to about 450 mm and decreases thereafter.
+\toprule
+ & \textbf{exp(Beta)} & \textbf{SE}\textsuperscript{\textit{1}} & \textbf{t} & \textbf{p-value} \\ 
+\midrule
+Intercept & 0.00 & 1.40 & -7.13 & <0.001 \\ 
+Maize GDD & 1.00 & 0.001 & 4.15 & <0.001 \\ 
+CD to Streams & 0.67 & 0.156 & -2.57 & 0.011 \\ 
+Protected & 4.82 & 0.497 & 3.16 & 0.002 \\ 
+\bottomrule
+\end{longtable}
+\begin{minipage}{\linewidth}
+\textsuperscript{\textit{1}}SE = Standard Error\\
+\end{minipage}
 
-A concurvity test on the final model was not conducted as there was only one smooth term.
-As expected, a VIF test for linear correlation in the parametric terms of the final model shows some evidence of correlation, particularly for GDD, which was close to 6, but this value is within acceptable limits at a moderate threshold.
-The Moran's I test suggests that there is no spatial autocorrelation in the untransformed residuals of the final model (I = 0.233, rank = 381, p = 0.476).
+}
 
-![
-Marginal response plots. Each plot shows the response of site counts to changes in a target covariate while holding all other variables at their mean.
-](../figures/marginal_response.png){#fig-responses fig-align="center" width="5in"}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{\linewidth}
+\subcaption{\label{tbl-results-2}Smooth Terms }
 
-# Discussion
+{\centering
 
-Our results indicate that Fremont maize farmers generally preferred warmer watersheds with longer growing seasons, moderate levels of precipitation (in the range of 400-600 mm), and greater access to perennial streams, as shown by the marginal response plots in Fig.
-3.
-While site counts exhibit a positive linear response to maize GDD, for maize has an upper threshold temperature at which point it no longer grows.
-Plus, increased temperatures trade-off with water availability, with warmer conditions being better for plant growth but at the same time increasing water demand on the plant to keep up with growth [@ramankutty2002].
-The linear relationship between site counts and maize GDD is, thus, relative to the observed distribution of that variable within the study area and should not be used to generalize to a larger geographic extent.
+\begin{longtable}{lcccc}
+\tabularnewline
 
-Another curious result of our analysis is that site counts do not increase linearly as a function of precipitation, but instead decline after 500 mm.
-We believe this is likely an artifact of trade-offs related to elevation, specifically the trade-off between precipitation and temperature.
-We discuss this more below, along with other important implications our results have for the regional prehistory of the eastern Great Basin, particularly for the spread and eventual collapse of maize agriculture.
+\toprule
+ & \textbf{edf} & \textbf{ref.df} & \textbf{F} & \textbf{p-value} \\ 
+\midrule
+s(Precipitation) & 3.25 & 3.25 & 11.1 & <0.001 \\ 
+\bottomrule
+\end{longtable}
 
-## A Goldilocks Zone?
+}
 
-Provided that the trade-off between precipitation and temperature is driven largely by elevation in this region, it would seem likely that the western Fremont chose site locations at elevations that maximized agricultural suitability, as evidenced by the estimated distribution of sites across watersheds (see Fig.
-4), though we caution that watersheds are coarse grained units of analysis.
-Given the topography of the study area, this trade-off would also suggest that areas suitable for maize agriculture take the form of a thin band around the lower slopes of mountain ranges (or lower canyons in the Uintas).
-Such a band would be akin to a Goldilocks zone for maize agriculture (Yaworsky et al, this issue), a place where the optima for GDD and precipitation overlap, though we should not make too much of this idea, for it is an area of overlap for continuous ecological gradients, not a region demarcated by anything like a perfect line.
-It also must be emphasized that what matters most for maize agriculture is not necessarily where precipitation falls, but where it accumulates once it enters the streamflow network.
-Of course, that will largely be a function of elevation in this region, or changes in elevation, so it will still be in proximity to the Goldilocks zone, though perhaps not perfectly coincident with it.
-At any rate, given these considerations, it would perhaps be more accurate to characterize the trade-off confronting the Fremont as one between what we might call water availability (including general precipitation across the streamflow network, water runoff, proximity to perennial surficial sources, and irrigation potential, among other things) and temperature (including, but not limited to, the length of the growing season and the number of frost-free days).
+\end{minipage}%
 
-## Irrigation and the Maize Farming Niche
+\end{table}
 
-Based on previous research [@benson2011; @adams2006; @bellorado2010], Bocinsky and Kohler [@bocinsky2014] place minimum thresholds for precipitation and maize GDD at 300 mm and 1000°C GDD, respectively.
-Importantly, these are thresholds for dry-farming, which is typically defined as farming free of irrigation [@benson2011; @cordell2012; @varien1999], so they can be loosely interpreted as minimum temperature and precipitation levels required to grow maize through nothing more than planting and harvesting.
-The right panel in @fig-thresholds shows the watersheds whose median values for maize GDD and precipitation over the Fremont sequence are above those values.
-This is similar to Bocinsky's concept of "refugia," but rather than calculate the proportion of years in niche, we are using median values, so refugia are defined here as watersheds that spend at least half the Fremont sequence in the rain-fed maize farming niche ($\geq$ 500 years, in this case).
+As shown in Table~\ref{tbl-results}, all linear (or parametric)
+coefficients are significant in the final model. The intercept
+(exp\(\,\beta\) = 0.00, p \textless{} 0.001) is close to zero when
+transformed back onto the response scale, likely reflecting the large
+number of watersheds with zero Fremont sites, though as mentioned above,
+the intercept estimate is not strictly meaningful for these data. The
+proportion of watersheds falling under federal protections
+(exp\(\,\beta\) = 4.82, p = 0.002) has a positive effect on site counts,
+thus confirming our expectation that more sites occur in watersheds with
+a larger proportion of federal land. Site counts increase in watersheds
+with larger GDD values, though the effect is small (exp\(\,\beta\) =
+1.00, p \textless{} 0.001), and site counts decrease with distance from
+perennial streams, though again the effect is small (exp\(\,\beta\) =
+0.67, p = 0.011). The only smooth term in the final model is average
+precipitation over the water-year, which is significant (edf = 3.25, p
+\textless{} 0.001). The effective degrees of freedom (EDF) for
+precipitation suggests a strong non-linear effect of that covariate on
+site counts, which increases up to about 450 mm and decreases
+thereafter.
 
-![
-The map on the left shows the relative density of feature-weighted site counts as estimated by the model, with lighter colors representing larger relative densities, darker colors representing smaller relative densities.
-The map on the right shows which watersheds are in the rain-fed maize farming niche for at least half of the Fremont sequence.
-The light green color indicates those that are in the niche.
-](../figures/threshold_map.png){#fig-thresholds fig-align="center" width="6in"}
+A concurvity test on the final model was not conducted as there was only
+one smooth term. As expected, a VIF test for linear correlation in the
+parametric terms of the final model shows some evidence of correlation,
+particularly for GDD, which was close to 6, but this value is within
+acceptable limits at a moderate threshold. The Moran's I test suggests
+that there is no spatial autocorrelation in the untransformed residuals
+of the final model (I = 0.233, rank = 381, p = 0.476).
 
-It appears that these models tend to agree on the best watersheds for maize farming, though there are some notable exceptions.
-The western slopes of the Stansbury Range in the north central part of the study area and parts of the Parowan Valley and surrounding areas in the southeast part of the study area are estimated by our model to have non-negligible relative densities, but they fall outside of the niche.
-Other areas, like much of the Salt Lake valley appear to be in the niche, but are estimated by our model to have relatively low densities.
-These discrepancies are at least partially explained by limitations of our methodology.
-For one thing, including the proportion of federal land as a covariate is a very crude way of accounting for sampling bias, so our model might not be picking up on the fact that the Salt Lake valley was actually a highly suitable area for maize agriculture, as evidence to that effect has largely been destroyed.
-The discrepancy may also be explained by the fact that we have aggregated to the watershed level, thus obscuring important environmental variation across both the landscape in general and site locations in particular.
-Were we to build a model of the disaggregated data, we would likely find that many agricultural sites do, in fact, fall in the rain-fed maize agricultural niche, even if the watersheds that contain them do not.
+\begin{figure}
 
-But setting aside those methodological issues, some much more interesting explanations may be offered, at least from the perspective of theory.
-In particular, the spatial covariance structure in our model may be picking up on the structure of the streamflow network, specifically the accumulation of water run-off as a function of slope.
-This would suggest that if there are more people in one watershed, there should be more more people in the downstream watershed, too, at least relative to other watersheds with similar precipitation and GDD values.
-In the language of the IFD model, the suitability of the watershed habitat should propagate through the streamflow network.
-Exploring that intuition in a more systematic way, however, would require much more careful consideration of streamflow.
-Unfortunately, that is well beyond the scope of the current paper.
+{\centering \includegraphics[width=5in,height=\textheight]{../figures/marginal_response.png}
 
-Still, the distribution of the rain-fed niche may offer additional insights.
-Consider the fact that of the 183 watersheds in the study area, 87% (n=159) have median GDD values greater than 1000°C, but only 54% (n=98) have median precipitation values above 300 mm, and a scant 40% (n=74) meet both requirements.
-This would suggest that precipitation is the primary limiting factor when it comes to the choice of watersheds in which to dry farm, for there are many more watersheds that meet the minimum GDD requirements than meet the minimum precipitation requirements.
-One important implication of this is that dry farming would have been very nearly impossible across most of the study area over the Fremont sequence and irrigation more or less necessary.
-We note that even within the niche, which we have defined somewhat liberally as those watersheds that are included for at least half the Fremont sequence, irrigation would almost certainly have been a necessity [@boomgarden2019; @matson1990; @spangler2010].
-Of course, the type and intensity of irrigation would probably differ between sites, with more intensive irrigation occurring where there is less precipitation and less intensive irrigation where there is more precipitation, whether that precipitation falls in the watersheds themselves or in the upstream watersheds that feed into them.
+}
 
-## The Spread of Maize Agriculture
+\caption{\label{fig-responses}Marginal response plots. Each plot shows
+the response of site counts to changes in a target covariate while
+holding all other variables at their mean.}
 
-Our results comport with previous work suggesting that maize farmers in the eastern Great Basin preferentially targeted a certain elevation band that coincides with alluvial fans, forming a rim around the lower slopes of mountains and ridges on or near important drainages and floodplains.
-The ruggedness and aridity of this region led to extreme circumscription, with the costs of intensive farming outside these areas being so severe as to render that alternative virtually unsustainable.
-For maize farmers, the Great Basin should, thus, have the look and feel of an island biogeography, with mountains rising up like islands out of an ocean of desert sand and sagebrush.
+\end{figure}
 
-To account for the timing and tempo of island colonization among food producers in Oceania, @kennett2006 offer an extension of the IFD framework that might actually apply to the Great Basin.
-Their model depends crucially on the potential for farming populations to introduce economies of scale or Allee effects, with suitability increasing with increasing population at low densities.
-They also assume that individuals will prefer nearer habitats to those that are farther away, under the assumption that greater distances impose greater settlement costs, all else being equal.
-Together, these assumptions suggest a pulsating pattern of maize spread into the North American Southwest, with maize agriculture growing, then spreading, growing, then spreading, as appears to have happened in Oceania.
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
-This need not be a wholesale movement or adoption of agricultural practices, either.
-@madsen1998 argue that Fremont subsistence largely varies along two dimensions: levels of residential mobility and levels of cultigen adoption.
-While some were more mobile, moving from isolated patch to isolated patch, others were more sedentary, typically tethered to a productive wetland or riparian area.
-As local populations increase, the benefits of increased sedentism and cultigen adoption would slowly begin to outweigh the benefits of a more mobile and foraging-centered diet.
-According to our model, individuals moving in this direction would tend to concentrate more of their time and energy in the maize farming niche.
-This would then have the effect of increasing population size within a smaller area, possibly leading to Allee effects, which would invite further sedentism from those in the surrounding area.
-Following this logic, the patchy adoption of agriculture makes more sense.
-The degree and speed with which it occurs is simply a function of how fast habitat quality declines and how quickly populations grow.
+Our results indicate that Fremont maize farmers generally preferred
+warmer watersheds with longer growing seasons, moderate levels of
+precipitation (in the range of 400-600 mm), and greater access to
+perennial streams, as shown by the marginal response plots in Fig. 3.
+While site counts exhibit a positive linear response to maize GDD, for
+maize has an upper threshold temperature at which point it no longer
+grows. Plus, increased temperatures trade-off with water availability,
+with warmer conditions being better for plant growth but at the same
+time increasing water demand on the plant to keep up with growth
+\citep{ramankutty2002}. The linear relationship between site counts and
+maize GDD is, thus, relative to the observed distribution of that
+variable within the study area and should not be used to generalize to a
+larger geographic extent.
 
-![
-Geographic distribution of Fremont sites across watersheds.
-The left panel shows the relative density derived from observed site counts, and the right panel shows the relative density from the estimated site counts.
-](../figures/model-map.png){#fig-model-map fig-align="center" width="6in"}
+Another curious result of our analysis is that site counts do not
+increase linearly as a function of precipitation, but instead decline
+after 500 mm. We believe this is likely an artifact of trade-offs
+related to elevation, specifically the trade-off between precipitation
+and temperature. We discuss this more below, along with other important
+implications our results have for the regional prehistory of the eastern
+Great Basin, particularly for the spread and eventual collapse of maize
+agriculture.
 
-## The Collapse of Maize Agriculture
+\hypertarget{a-goldilocks-zone}{%
+\subsection{A Goldilocks Zone?}\label{a-goldilocks-zone}}
 
-After farming these regions successfully for over a thousand years, the Fremont began an abrupt process of abandonment starting around 700 years BP.
-The cause of this event is still a matter of some dispute, though a growing consensus appears to be coalescing around climatic changes driving severe reductions in agricultural productivity [@benson2007; @spangler2019; @thomson2019; @thomson2020; @finley2020].
-We follow @lindsay1986 in thinking this was likely owing to climatic events that, in effect, tore the Goldilocks zone apart, pulling higher temperatures further down in elevation and pushing higher precipitation levels further up.
+Provided that the trade-off between precipitation and temperature is
+driven largely by elevation in this region, it would seem likely that
+the western Fremont chose site locations at elevations that maximized
+agricultural suitability, as evidenced by the estimated distribution of
+sites across watersheds (see Fig. 4), though we caution that watersheds
+are coarse grained units of analysis. Given the topography of the study
+area, this trade-off would also suggest that areas suitable for maize
+agriculture take the form of a thin band around the lower slopes of
+mountain ranges (or lower canyons in the Uintas). Such a band would be
+akin to a Goldilocks zone for maize agriculture (Yaworsky et al, this
+issue), a place where the optima for GDD and precipitation overlap,
+though we should not make too much of this idea, for it is an area of
+overlap for continuous ecological gradients, not a region demarcated by
+anything like a perfect line. It also must be emphasized that what
+matters most for maize agriculture is not necessarily where
+precipitation falls, but where it accumulates once it enters the
+streamflow network. Of course, that will largely be a function of
+elevation in this region, or changes in elevation, so it will still be
+in proximity to the Goldilocks zone, though perhaps not perfectly
+coincident with it. At any rate, given these considerations, it would
+perhaps be more accurate to characterize the trade-off confronting the
+Fremont as one between what we might call water availability (including
+general precipitation across the streamflow network, water runoff,
+proximity to perennial surficial sources, and irrigation potential,
+among other things) and temperature (including, but not limited to, the
+length of the growing season and the number of frost-free days).
 
-Lindsay suggests that the end of the Medieval Warm Period may have been responsible for this decoupling, as it involved higher temperatures and more growing-season precipitation owing to the northward intrusion of summer monsoons.
-This period lasted from roughly 1,000 to 600 years BP [@grayson2011; @graumlich1993], so it does coincide with the height of the Fremont complex.
-Given the logic of IFD, this likely led to population spillovers into less suitable habitats [@codding2021].
-However, given the extreme circumscription we have highlighted in the Great Basin, the available alternatives would have been limited, potentially leading to increasing population packing across all habitats during this time, thus making the Fremont especially sensitive to climatic downturns.
+\hypertarget{irrigation-and-the-maize-farming-niche}{%
+\subsection{Irrigation and the Maize Farming
+Niche}\label{irrigation-and-the-maize-farming-niche}}
 
-The well-documented megadrought at the end of the thirteenth century [@cook2004] would have been catastrophic in this context, reducing precipitation levels across the region [@benson2007], but also leading to reductions in water accumulation within the streamflow network.
-So, just as temperatures are beginning to cool, precipitation levels collapse.
-By extension, this would have decreased overall water availability and, thus, the potential for irrigation.
-Opportunities for the Fremont to recover from such a drought would have been limited, too, as the end of the Medieval Warm Period was also the beginning of the Little Ice Age, a time during which conditions were generally colder and wetter [@mann2009], further pulling apart the Goldilocks zone upon which the Fremont so long depended.
+Based on previous research \citep{benson2011, adams2006, bellorado2010},
+Bocinsky and Kohler \citep{bocinsky2014} place minimum thresholds for
+precipitation and maize GDD at 300 mm and 1000°C GDD, respectively.
+Importantly, these are thresholds for dry-farming, which is typically
+defined as farming free of irrigation
+\citep{benson2011, cordell2012, varien1999}, so they can be loosely
+interpreted as minimum temperature and precipitation levels required to
+grow maize through nothing more than planting and harvesting. The right
+panel in Figure~\ref{fig-thresholds} shows the watersheds whose median
+values for maize GDD and precipitation over the Fremont sequence are
+above those values. This is similar to Bocinsky's concept of
+``refugia,'' but rather than calculate the proportion of years in niche,
+we are using median values, so refugia are defined here as watersheds
+that spend at least half the Fremont sequence in the rain-fed maize
+farming niche (\(\geq\) 500 years, in this case).
+
+\begin{figure}
+
+{\centering \includegraphics[width=6in,height=\textheight]{../figures/threshold_map.png}
+
+}
+
+\caption{\label{fig-thresholds}The map on the left shows the relative
+density of feature-weighted site counts as estimated by the model, with
+lighter colors representing larger relative densities, darker colors
+representing smaller relative densities. The map on the right shows
+which watersheds are in the rain-fed maize farming niche for at least
+half of the Fremont sequence. The light green color indicates those that
+are in the niche.}
+
+\end{figure}
+
+It appears that these models tend to agree on the best watersheds for
+maize farming, though there are some notable exceptions. The western
+slopes of the Stansbury Range in the north central part of the study
+area and parts of the Parowan Valley and surrounding areas in the
+southeast part of the study area are estimated by our model to have
+non-negligible relative densities, but they fall outside of the niche.
+Other areas, like much of the Salt Lake valley appear to be in the
+niche, but are estimated by our model to have relatively low densities.
+These discrepancies are at least partially explained by limitations of
+our methodology. For one thing, including the proportion of federal land
+as a covariate is a very crude way of accounting for sampling bias, so
+our model might not be picking up on the fact that the Salt Lake valley
+was actually a highly suitable area for maize agriculture, as evidence
+to that effect has largely been destroyed. The discrepancy may also be
+explained by the fact that we have aggregated to the watershed level,
+thus obscuring important environmental variation across both the
+landscape in general and site locations in particular. Were we to build
+a model of the disaggregated data, we would likely find that many
+agricultural sites do, in fact, fall in the rain-fed maize agricultural
+niche, even if the watersheds that contain them do not.
+
+But setting aside those methodological issues, some much more
+interesting explanations may be offered, at least from the perspective
+of theory. In particular, the spatial covariance structure in our model
+may be picking up on the structure of the streamflow network,
+specifically the accumulation of water run-off as a function of slope.
+This would suggest that if there are more people in one watershed, there
+should be more more people in the downstream watershed, too, at least
+relative to other watersheds with similar precipitation and GDD values.
+In the language of the IFD model, the suitability of the watershed
+habitat should propagate through the streamflow network. Exploring that
+intuition in a more systematic way, however, would require much more
+careful consideration of streamflow. Unfortunately, that is well beyond
+the scope of the current paper.
+
+Still, the distribution of the rain-fed niche may offer additional
+insights. Consider the fact that of the 183 watersheds in the study
+area, 87\% (n=159) have median GDD values greater than 1000°C, but only
+54\% (n=98) have median precipitation values above 300 mm, and a scant
+40\% (n=74) meet both requirements. This would suggest that
+precipitation is the primary limiting factor when it comes to the choice
+of watersheds in which to dry farm, for there are many more watersheds
+that meet the minimum GDD requirements than meet the minimum
+precipitation requirements. One important implication of this is that
+dry farming would have been very nearly impossible across most of the
+study area over the Fremont sequence and irrigation more or less
+necessary. We note that even within the niche, which we have defined
+somewhat liberally as those watersheds that are included for at least
+half the Fremont sequence, irrigation would almost certainly have been a
+necessity \citep{boomgarden2019, matson1990, spangler2010}. Of course,
+the type and intensity of irrigation would probably differ between
+sites, with more intensive irrigation occurring where there is less
+precipitation and less intensive irrigation where there is more
+precipitation, whether that precipitation falls in the watersheds
+themselves or in the upstream watersheds that feed into them.
+
+\hypertarget{the-spread-of-maize-agriculture}{%
+\subsection{The Spread of Maize
+Agriculture}\label{the-spread-of-maize-agriculture}}
+
+Our results comport with previous work suggesting that maize farmers in
+the eastern Great Basin preferentially targeted a certain elevation band
+that coincides with alluvial fans, forming a rim around the lower slopes
+of mountains and ridges on or near important drainages and floodplains.
+The ruggedness and aridity of this region led to extreme
+circumscription, with the costs of intensive farming outside these areas
+being so severe as to render that alternative virtually unsustainable.
+For maize farmers, the Great Basin should, thus, have the look and feel
+of an island biogeography, with mountains rising up like islands out of
+an ocean of desert sand and sagebrush.
+
+To account for the timing and tempo of island colonization among food
+producers in Oceania, \citep{kennett2006} offer an extension of the IFD
+framework that might actually apply to the Great Basin. Their model
+depends crucially on the potential for farming populations to introduce
+economies of scale or Allee effects, with suitability increasing with
+increasing population at low densities. They also assume that
+individuals will prefer nearer habitats to those that are farther away,
+under the assumption that greater distances impose greater settlement
+costs, all else being equal. Together, these assumptions suggest a
+pulsating pattern of maize spread into the North American Southwest,
+with maize agriculture growing, then spreading, growing, then spreading,
+as appears to have happened in Oceania.
+
+This need not be a wholesale movement or adoption of agricultural
+practices, either. \citep{madsen1998} argue that Fremont subsistence
+largely varies along two dimensions: levels of residential mobility and
+levels of cultigen adoption. While some were more mobile, moving from
+isolated patch to isolated patch, others were more sedentary, typically
+tethered to a productive wetland or riparian area. As local populations
+increase, the benefits of increased sedentism and cultigen adoption
+would slowly begin to outweigh the benefits of a more mobile and
+foraging-centered diet. According to our model, individuals moving in
+this direction would tend to concentrate more of their time and energy
+in the maize farming niche. This would then have the effect of
+increasing population size within a smaller area, possibly leading to
+Allee effects, which would invite further sedentism from those in the
+surrounding area. Following this logic, the patchy adoption of
+agriculture makes more sense. The degree and speed with which it occurs
+is simply a function of how fast habitat quality declines and how
+quickly populations grow.
+
+\begin{figure}
+
+{\centering \includegraphics[width=6in,height=\textheight]{../figures/model-map.png}
+
+}
+
+\caption{\label{fig-model-map}Geographic distribution of Fremont sites
+across watersheds. The left panel shows the relative density derived
+from observed site counts, and the right panel shows the relative
+density from the estimated site counts.}
+
+\end{figure}
+
+\hypertarget{the-collapse-of-maize-agriculture}{%
+\subsection{The Collapse of Maize
+Agriculture}\label{the-collapse-of-maize-agriculture}}
+
+After farming these regions successfully for over a thousand years, the
+Fremont began an abrupt process of abandonment starting around 700 years
+BP. The cause of this event is still a matter of some dispute, though a
+growing consensus appears to be coalescing around climatic changes
+driving severe reductions in agricultural productivity
+\citep{benson2007, spangler2019, thomson2019, thomson2020, finley2020}.
+We follow \citep{lindsay1986} in thinking this was likely owing to
+climatic events that, in effect, tore the Goldilocks zone apart, pulling
+higher temperatures further down in elevation and pushing higher
+precipitation levels further up.
+
+Lindsay suggests that the end of the Medieval Warm Period may have been
+responsible for this decoupling, as it involved higher temperatures and
+more growing-season precipitation owing to the northward intrusion of
+summer monsoons. This period lasted from roughly 1,000 to 600 years BP
+\citep{grayson2011, graumlich1993}, so it does coincide with the height
+of the Fremont complex. Given the logic of IFD, this likely led to
+population spillovers into less suitable habitats \citep{codding2021}.
+However, given the extreme circumscription we have highlighted in the
+Great Basin, the available alternatives would have been limited,
+potentially leading to increasing population packing across all habitats
+during this time, thus making the Fremont especially sensitive to
+climatic downturns.
+
+The well-documented megadrought at the end of the thirteenth century
+\citep{cook2004} would have been catastrophic in this context, reducing
+precipitation levels across the region \citep{benson2007}, but also
+leading to reductions in water accumulation within the streamflow
+network. So, just as temperatures are beginning to cool, precipitation
+levels collapse. By extension, this would have decreased overall water
+availability and, thus, the potential for irrigation. Opportunities for
+the Fremont to recover from such a drought would have been limited, too,
+as the end of the Medieval Warm Period was also the beginning of the
+Little Ice Age, a time during which conditions were generally colder and
+wetter \citep{mann2009}, further pulling apart the Goldilocks zone upon
+which the Fremont so long depended.
+
+  \bibliography{bibliography.bib}
+
+\end{document}

### 2023-04-30 R/cost-distance.R

@@ -1,0 +1,135 @@
+
+# Site distribution for the western Fremont -- Data Wrangling
+# Vernon 2022
+
+##############################.
+#     Table of Contents
+# 01. Preamble
+# 02. Load data
+# 03. Prepare data 
+# 04. Calculate cost-distance
+# 05. 
+##############################.
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+# Load libraries
+library(furrr)
+library(here)
+library(hiker)
+library(sf)
+library(terra)
+library(tidyverse)
+
+# geopackage database
+gpkg <- here("data", "western-fremont.gpkg")
+
+# Set a "plan" for how the code should run.
+plan(multisession, workers = 4)
+
+# load custom functions
+# doing it this way to mask the functions in the global environment
+sys.source(
+  here("R", "fun-extractors.R"),
+  envir = attach(NULL, name = "extractors")
+)
+
+sys.source(
+  here("R", "fun-cost_distance.R"),
+  envir = attach(NULL, name = "cost")
+)
+
+# 02) LOAD DATA -----------------------------------------------------------
+
+dem <- rast(here("data", "rast-250m_dem.tif"))
+
+window     <- read_sf(gpkg, layer = "window")
+watersheds <- read_sf(gpkg, layer = "watersheds")
+springs    <- read_sf(gpkg, layer = "springs")
+streams    <- read_sf(gpkg, layer = "streams")
+roads      <- read_sf(gpkg, layer = "roads")
+
+# 03) PREPARE DATA --------------------------------------------------------
+
+# aggregate the dem to speed up processing time
+# at the scale, it shouldn't make much difference
+dem <- project(dem, crs(dem), res = c(500,500))
+
+# subset roads to interstate (I), us highway (U), and state highway (S)
+roads <- roads |>
+  filter(rttyp %in% c("U", "I", "S")) |>
+  st_intersection(window)
+
+# hiker only works with point locations, so sample one point every 500 meters (1/500)
+# this will exclude line segments shorter than 500 meters in length
+roads <- roads |>
+  st_cast("LINESTRING") |>
+  st_line_sample(density = 1/500) |>
+  st_cast("POINT") |>
+  st_as_sf() |>
+  (\(x) filter(!st_is_empty(x)))()
+
+streams <- streams |>
[…]

### 2023-04-30 R/data-wrangling.R

@@ -1,0 +1,351 @@
+
+# Site distribution for the western Fremont -- Data Wrangling
+# Vernon 2022
+
+##############################.
+#     Table of Contents
+# 01. Preamble
+# 02. Utah polygon
+# 03. Watersheds 
+# 04. Sites
+# 05. Digital Elevation Model (DEM)
+# 06. Land ownership
+# 07. Water features
+# 08. Roads
+# 09. PRISM
+# 10. Tree Rings (ITRDB)
+# 11. Cost-distance
+# 12. Extract all data into watersheds 
+##############################.
+
+# Note:
+# 1- Sensitive site data cannot be shared directly, so you'll need to get it
+#    from NV and UT SHPOs, or get their permission to get it from us.
+#    You can't run this code without it!
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+# Load libraries
+library(FedData)
+library(here)
+library(sf)
+library(spData)
+library(terra)
+library(tidyverse)
+library(tigris)
+
+# geopackage database
+gpkg <- here("data", "western-fremont.gpkg")
+
+# coordinate reference 
+# EPSG:26912
+
+# load custom functions
+# doing it this way to mask the functions in the global environment
+sys.source(
+  here("R", "fun-download_features.R"), # requires httr and jsonlite
+  envir = attach(NULL, name = "features")
+)
+
+sys.source(
+  here("R", "fun-extractors.R"),
+  envir = attach(NULL, name = "extractors")
+)
+
+# use ls(features) or ls(extractors) to see the functions
+
+# 02) UTAH POLYGON --------------------------------------------------------
+
+utah <- states() |> 
+  filter(NAME == "Utah") |> 
+  select(NAME) |> 
+  rename_with(tolower) |> 
+  st_transform(26912)
+
+write_sf(
+  utah, 
+  dsn = gpkg, 
+  layer = "utah"
+)
+
+# 03) WATERSHEDS ----------------------------------------------------------
+
+watersheds <- utah |> 
+  # to make sure we pick up the HUC12 inside the HUC8 but outside UT
+  st_buffer(30000) |> 
+  get_wbd(label = "western_fremont") |> 
+  st_transform(26912) |> 
+  select(HUC12) |> 
+  rename("id" = HUC12) |> 
[…]

### 2023-04-30 R/fig-elevation-tradeoffs.R

@@ -1,0 +1,135 @@
+
+# Elevation drives everything
+
+##############################.
+#     Table of Contents
+# 01. Preamble
+# 02. Load data
+# 03. PaleoCAR models
+# 04. Collect results in HUC10 watersheds
+# 05. 
+##############################.
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+library(broom)
+library(geomtextpath)
+library(here)
+library(terra)
+library(tidyverse)
+
+# 02) DATA ----------------------------------------------------------------
+
+elevation <- rast(here("data", "rast-250m_dem.tif"))
+streams <- rast(here("data", "rast-500m_cd_streams.tif"))
+streams <- streams/3600 # convert to hours
+
+precipitation <- rast(here("data", "rast-800m_ppt.tif")) |> 
+  project("EPSG:26912") |> 
+  mean()
+
+gdd <- rast(here("data", "rast-800m_gdd.tif")) |> 
+  project("EPSG:26912") |> 
+  mean()
+
+elevation <- resample(elevation, gdd)
+streams <- resample(streams, gdd)
+
+rstack <- c(elevation, streams, precipitation, gdd)
+names(rstack) <- c("elevation", "streams", "precipitation", "gdd")
+
+# 03) OLS -----------------------------------------------------------------
+
+training_data <- spatSample(
+  rstack,
+  size = 10000,
+  method = "regular",
+  na.rm = TRUE
+) |> 
+  na.omit() |> 
+  arrange(elevation)
+
+fit_lm <- function(variable, .data) {
+  
+  .f <- as.formula(paste0(variable, "~elevation"))
+  
+  .m <- lm(.f, .data)
+  
+  broom::tidy(.m) |> 
+    rename(
+      "x" = term,
+      "t.statistic" = statistic
+    ) |> 
+    slice(2) |> 
+    mutate(
+      "y" = variable,
+      "r2" = summary(.m)$r.squared
+    ) |> 
+    select(y, x, everything())
+  
+}
+
+q <- lapply(
+  c("gdd", "precipitation", "streams"), 
+  fit_lm, 
+  .data = training_data
+) |> bind_rows()
+
+bob <- global(c(gdd, precipitation, streams), mean, na.rm = TRUE)[, 1, drop = TRUE]
+
+q <- mutate(q, "y.mean" = round(bob, 2))
+
+write_csv(q, here("data", "elevation-everything.csv"))
+
[…]

### 2023-04-30 R/fig-overview-map.R

@@ -1,0 +1,224 @@
+
+# Create a regional overview map of the project area
+
+####################.
+#    OUTLINE
+# 1) R Preamble
+# 2) features
+# 3) basemap
+# 4) main map
+# 5) inset map
+# 6) site density map
+# 7) save
+####################.
+
+# R preamble --------------------------------------------------------------
+
+library(ggfx)
+library(ggspatial)
+library(here)
+library(patchwork)
+library(sf)
+library(tidyverse)
+library(tigris)
+library(viridis)
+
+gpkg <- here("data", "western-fremont.gpkg")
+
+# load custom functions
+# doing it this way to mask the functions in the global environment
+sys.source(
+  here("R", "fun-get_basemap.R"),
+  envir = attach(NULL, name = "basemap")
+)
+
+# features ----------------------------------------------------------------
+
+state_names <- c(
+  "Idaho", "Wyoming", "Colorado", "New Mexico", "Arizona", "Nevada", "Utah"
+)
+
+states <- tigris::states() |> 
+  rename_with(tolower) |> 
+  filter(name %in% state_names) |> 
+  select(name, stusps) |>
+  st_transform(26912)
+
+utah <- states |> filter(name == "Utah")
+
+window <- read_sf(gpkg, "window")
+
+roads <- read_sf(gpkg, "roads")
+
+roads <- st_intersection(
+  roads, 
+  st_union(utah, window)
+)
+
+watersheds <- read_sf(gpkg, "watersheds")
+
+# basemap -----------------------------------------------------------------
+
+bb8 <- st_union(window, utah) |> st_bbox()
+
+dy <- bb8[["ymax"]] - bb8[["ymin"]]
+dx <- bb8[["xmax"]] - bb8[["xmin"]]
+
+basemap <- get_basemap(
+  bb8, 
+  map = "imagery",
+  size = c(9000, 9000*(dy/dx)),
+  imageSR = 26912
+)
+
+# main map ----------------------------------------------------------------
+
+bob <- ggplot() +
+  as_reference(
+    geom_sf(
+      data = utah,
+      color = "transparent",
+      fill = alpha("white", 0.3)
+    ),
+    id = "utah-mask"
+  ) +
+  with_mask(
+    annotation_raster(
+      basemap,
+      bb8[["xmin"]], bb8[["xmax"]],
+      bb8[["ymin"]], bb8[["ymax"]]
[…]

### 2023-04-30 R/fun-cost_distance.R

@@ -1,0 +1,87 @@
+
+# this is a modification of hf_survey() that loops over origins,
+# running in parallel and returning a vector, 
+# for use in dplyr::mutate()
+# not sure this is the most efficient way to do this
+# would like to try some pre-processing of the graph in the future
+survey_parallel <- function(terrain, from, ngroups = 500){
+  
+  rr <- terra::rast(
+    nrow   = terrain$nrow,
+    ncol   = terrain$ncol,
+    extent = terra::ext(terrain$bb8),
+    crs    = terrain$crs
+  )
+  
+  from_xy <- sf::st_coordinates(from)[, 1:2, drop = FALSE]
+  
+  from_cells <- terra::cellFromXY(rr, from_xy)
+  
+  groups <- as.numeric(
+    gl(
+      n = nrow(from),
+      k = ngroups,
+      length = nrow(from)
+    )
+  )
+  
+  from_cells <- split(from_cells, groups)
+  
+  to_cells <- unique(as.integer(terrain$conductance@j)) + 1
+  
+  graph <- igraph::graph_from_adjacency_matrix(
+    terrain$conductance,
+    mode = "directed",
+    weighted = TRUE
+  )
+  
+  # invert conductance to get travel cost
+  igraph::E(graph)$weight <- (1/igraph::E(graph)$weight)
+  
+  cat("Building cost-distance matrices.\n")
+  
+  p <- progressr::progressor(length(from)/2)
+  
+  cost <- furrr::future_map(
+    from_cells,
+    survey,
+    to = to_cells,
+    graph = graph,
+    p = p
+  )
+  
+  cat("\nBuilding aggregate cost-distance raster.\n\n")
+  
+  rasterize(cost, rr, to_cells)
+  
+}
+
+survey <- function(from, to, graph, p){
+  
+  cost <- igraph::distances(graph, from, to, mode = "out")
+  
+  cost <- apply(cost, 2, min)
+  
+  p()
+  
+  invisible(gc())
+  
+  return(cost)
+  
+}
+
+rasterize <- function(x, r, to_cells){
+  
+  r[] <- NA
+  
+  r <- rep(r, length(x))
+  
+  for (i in seq_along(x)){
+    
+    r[[i]][to_cells] <- x[[i]]
+    
+  }
+  
+  min(r, na.rm = TRUE)
+  
+}

### 2023-04-30 R/fun-download_features.R

@@ -1,0 +1,75 @@
+
+# some functions for working with ArcGIS REST API
+# use feature ids to download features in bundles, work around download size limits
+
+get_ids <- function(envelope, url, where = ""){
+  
+  query <- list(
+    f = "json",
+    where = where,
+    geometry = envelope,
+    geometryType = "esriGeometryEnvelope",
+    inSR = 4326, 
+    returnIdsOnly = TRUE,
+    returnGeometry = FALSE
+  )
+  
+  html <- httr::GET(url, query = query)
+  
+  guts <- httr::content(
+    html, 
+    as = "text", 
+    encoding = "UTF-8"
+  )
+  
+  json <- jsonlite::fromJSON(guts)
+  
+  json[['objectIds']]
+  
+}
+
+get_features <- function(x, url){
+  
+  objectIds <- paste0(x, collapse = ",")
+  
+  query <- list(
+    f = "json", # gotta use json because geojson is generating bad request errors
+    outFields = "*",
+    objectIds = objectIds,
+    outSR = 4326
+  )
+  
+  html <- httr::GET(url, query = query)
+  
+  result <- tryCatch(
+    {
+      features <- httr::content(html, as = "text", encoding = "UTF-8")
+      sf::read_sf(features)
+    },
+    error = function(cond) {
+      message(cond)
+      message(paste("\n\nfor", objectIds))
+      return(NULL)
+    }
+  )
+  
+  result
+  
+}
+
+download_features <- function(envelope, url, where = "", dlsize = 200) {
+  
+  bb8 <- sf::st_transform(envelope, crs = 4326)
+  bb8 <- sf::st_bbox(bb8)
+  bb8 <- round(bb8, 4)
+  bb8 <- paste(bb8, collapse = ",")
+  
+  ids <- get_ids(url = url, envelope = bb8, where = where)
+  
+  ids <- split(ids, ceiling(seq_along(ids)/dlsize))
+  
+  features <- lapply(ids, get_features, url = url)
+  
+  do.call("rbind", features)
+  
+}

### 2023-04-30 R/fun-extractors.R

@@ -1,0 +1,47 @@
+
+# calculate proportion of polygon x that polygon y overlaps
+# vectorized to work with dplyr::mutate
+calculate_overlap <- function(x,  y) {
+  
+  suppressWarnings({
+    
+    area_x <- st_area(x)
+    
+    area_y <- unlist(
+      lapply(1:nrow(x), function(z){
+        
+        clip <- st_intersection(y, x[z,])
+        clip <- st_union(clip)
+        
+        st_area(clip)
+        
+      })
+    )
+    
+  })
+  
+  p <- area_y/area_x
+  
+  as.numeric(p)
+  
+}
+
+
+# have terra::extract() return a vector so it can work in dplyr::mutate()
+extract_value <- function(x, y, .fun){
+  
+  df <- terra::extract(y, vect(x), fun = .fun, na.rm = TRUE)
+  
+  df[, 2, drop = TRUE]
+  
+}
+
+# count the number of site points in each watershed polygon
+# vectorized for use in dplyr::mutate()
+count_sites <- function(x, y) {
+  
+  i <- sf::st_intersects(x, y)
+  
+  lengths(i)
+  
+}

### 2023-04-30 R/fun-get_basemap.R

@@ -1,0 +1,69 @@
+
+get_basemap <- function(x,
+                        map = "physical",
+                        size = c(16000,9000),
+                        dpi = 900,
+                        imageSR = 4326,
+                        native = FALSE){
+
+  x <- st_bbox(x)
+
+  old_ratio <- (x[["xmax"]] - x[["xmin"]]) / (x[["ymax"]] - x[["ymin"]])
+  new_ratio <- (size[[1]] / size[[2]])
+
+  if (!all.equal(old_ratio, new_ratio)) {
+
+    msg <- paste0(
+      "Extent of image (size) differs from extent of x (bbox). ",
+      "Map may be warped."
+    )
+
+    warning(msg, call. = FALSE)
+
+  }
+
+  base_url <- "http://services.arcgisonline.com/arcgis/rest/services/"
+
+  map_name <- switch(
+    map,
+    "hillshade"= "Elevation/World_Hillshade",
+    "dark"     = "Elevation/World_Hillshade_Dark",
+    "natgeo"   = "NatGeo_World_Map",
+    "usa"      = "USA_Topo_Maps",
+    "imagery"  = "World_Imagery",
+    "physical" = "World_Physical_Map",
+    "shaded"   = "World_Shaded_Relief",
+    "street"   = "World_Street_Map",
+    "terrain"  = "World_Terrain_Base",
+    "topo"     = "World_Topo_map",
+    stop(paste0("The ", map, " map is not supported."), call. = FALSE)
+  )
+
+  map_service <- paste0(base_url, map_name, "/MapServer/export?")
+
+  query <- list(
+    bbox    = paste(x, collapse = ","),
+    bboxSR  = st_crs(x)$epsg,
+    imageSR = imageSR,
+    f       = "image",
+    format  = "png",
+    size    = paste(size, collapse = ","),
+    dpi     = dpi
+  )
+
+  request <- httr::GET(
+    map_service,
+    query = query
+  )
+
+  result <- httr::content(request)
+
+  if (native) {
+
+    result <- grDevices::as.raster(result, native = TRUE)
+
+  }
+
+  return(result)
+
+}

### 2023-04-30 R/paleocar.R

@@ -1,0 +1,100 @@
+
+# PaleoCAR climate reconstruction for western United States
+
+##############################.
+#     Table of Contents
+# 01. Preamble
+# 02. Load data
+# 03. PaleoCAR models
+# 04. Collect results in HUC10 watersheds
+# 05. 
+##############################.
+
+# Note: for this analysis, we average the results (collapse the time dimension) 
+# across the entire sequence, as we lack a tight site occupation chronology
+# thus, we are getting something similar to Bocinsky's maize farming "refugia"
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+library(here)
+library(paleocar)
+library(sf)
+library(terra)
+library(tidyverse)
+
+# geopackage database
+gpkg <- here("data", "western-fremont.gpkg")
+
+# coordinate reference 
+# EPSG:4326 (WGS84)
+
+years_cal <- 1924:1983
+years_prd <- 401:1400
+
+# 02) LOAD DATA -----------------------------------------------------------
+
+watersheds <- read_sf(gpkg, layer = "watersheds") |> st_transform(4326)
+
+tree_rings <- read_rds(here("data", "western_fremont_ITRDB.Rds"))
+
+ppt <- here("data", "rast-800m_ppt.tif") |> rast()
+gdd <- here("data", "rast-800m_gdd.tif") |> rast()
+
+# 03) PALEOCAR MODELS -----------------------------------------------------
+
+ppt <- ppt |> 
+  terra::extract(vect(watersheds), fun = mean) |> 
+  select(-ID) |> 
+  t()
+
+pcar_ppt <- paleocar(
+  chronologies      = itrdb,
+  predictands       = ppt,
+  calibration.years = years_cal,
+  prediction.years  = years_prd,
+  min.width         = 5,
+  verbose           = TRUE,
+  label             = "ppt",
+  out.dir           = here("data")
+)
+
+gdd <- gdd |> 
+  terra::extract(vect(watersheds), fun = mean) |> 
+  select(-ID) |> 
+  t()
+
+pcar_gdd <- paleocar(
+  chronologies      = itrdb,
+  predictands       = gdd,
+  calibration.years = years_cal,
+  prediction.years  = years_prd,
+  verbose           = TRUE,
+  label             = "gdd",
+  out.dir           = here("data")
+)
+
[…]

### 2023-04-30 R/prism.R

@@ -1,0 +1,244 @@
+
+# Extract PRISM Data (LT81, 800-m resolution)
+# and prepare it for PaleoCAR climate reconstruction
+
+##############################.
+#     Table of Contents
+# 01. Preamble
+# 02. Western US
+# 03. Precipitation
+# 04. Temperature maximum
+# 05. Temperature minimum
+# 06. Maize GDD
+##############################.
+
+# Note: 
+# 1- LT81 is behind a paywall (http://www.prism.oregonstate.edu),
+#    and you can't run this code without it!
+# 2- This code is adapted from Bocinsky (2016) 
+# 3- The basic idea behind paleocar is to model climate trends in an area of 
+#    interest as a function of tree-ring chronologies over a much, much wider area.
+# 4- The aoi for this project is western Utah.
+
+# 01) PREAMBLE ------------------------------------------------------------
+
+library(here)
+library(sf)
+library(terra)
+library(tidyverse)
+
+# geopackage database
+gpkg <- here("data", "western-fremont.gpkg")
+
+# coordinate reference 
+# EPSG:4326 (WGS84)
+
+# This MUST point to the LT81 dataset 
+prism <- "D:/PRISM_800m_DATA"
+
+# years required, 1924-1983 
+# specified in list.files(pattern = years)
+# complicated to get a regex to handle a range of numbers, so it's a bit messy 
+
+# 02) PROJECT WINDOW ------------------------------------------------------
+
+window <- read_sf(gpkg, layer = "window") |> 
+  st_buffer(1000) |> 
+  st_transform(4326)
+
+watersheds <- read_sf(gpkg, layer = "watersheds")
+
+# 03) PRECIPITATION -------------------------------------------------------
+
+# including 1923 to get full 1924 water-year
+# water-year = October to September 
+
+zipfiles <- prism |> 
+  file.path("ppt") |> 
+  list.files(
+    full.names = TRUE, 
+    pattern = "((198[0-3])|(19[3-7][0-9]{1})|(192[3-9])).zip$"
+  )
+
+exdir <- file.path(tempdir(), "ppt")
+
+dir.create(exdir)
+
+# unzip returns a vector of paths to unzipped files
+files <- lapply(zipfiles, unzip, exdir = exdir)
+
+files <- files |> 
[…]

### 2023-04-30 R/regression-table.R

@@ -1,0 +1,94 @@
+
+library(dplyr)
+library(glue)
+library(gt)
+library(gtsummary)
+library(here)
+library(mgcv)
+
+fremont <- readRDS(here::here("data", "western-fremont-model.Rds"))
+
+cap <- function(x) paste0(
+  "<div style='text-align: left; font-weight: bold; color: grey'>",
+  x, 
+  "</div>"
+)
+
+parametric <- tbl_regression(
+  fremont$gam,
+  label = list(
+    `(Intercept)` = "Intercept",
+    gdd = "Maize GDD",
+    streams = "CD to Streams",
+    protected = "Protected"
+  ),
+  exponentiate = TRUE, 
+  include = !starts_with("s("),
+  intercept = TRUE, 
+  tidy_fun = tidy_gam 
+) |> 
+  modify_column_hide(columns = ci) |> 
+  modify_column_unhide(columns = c(statistic, std.error)) |> 
+  modify_header(
+    label = "",
+    statistic = "**t**"
+  )
+
+smooth <- tbl_regression(
+  fremont$gam,
+  label = list(`s(precipitation)` = "s(Precipitation)"),
+  exponentiate = TRUE, 
+  include = starts_with("s("),
+  intercept = FALSE, 
+  tidy_fun = tidy_gam 
+) |> 
+  modify_column_hide(columns = c(estimate, ci)) |> 
+  modify_column_unhide(columns = c(edf, ref.df, statistic)) |> 
+  modify_table_body(~.x |> relocate(statistic, .before = p.value)) |> 
+  modify_header(
+    label = "",
+    statistic = "**F**",
+    edf = "**edf**",
+    ref.df = "**ref.df**"
+  )
+
+parametric_table <- parametric |> 
+  as_gt() |> 
+  tab_options(
+    table.width = pct(50),
+    table.font.names = "Times New Roman"
+  ) |> 
+  as_latex()
+
+smooth_table <- smooth |> 
+  as_gt() |> 
+  tab_options(
+    table.width = pct(50),
+    table.font.names = "Times New Roman"
+  ) |> 
+  as_latex()
+
+p <- parametric |> 
+  as_tibble() |> 
+  rename("v" = 1, "b" = 2, "p" = 5) |> 
+  select(v, b, p) |> 
+  mutate(x = glue("(exp$\\,\\beta$ = {b}, p = {p})"))
[…]

### 2023-05-01 manuscript/manuscript.qmd
prepare manuscript for QI sub

@@
 #| tbl-cap: Model Results
-#| tbl-subcap:
-#|   - "Parametric Terms"
-#|   - "Smooth Terms"
-#| layout-nrow: 2
-#| layout: "[[1], [-1], [1]]"
-#| out-width: 50%
+#| output: asis
 
-# returns parametric and smooth regression tables
-# and q, a list with character strings for reporting estimates in text
+# tables in quarto are still a hot mess
+cat(
+"
+\\begin{table}[]
+\\centering
+\\begin{tabular}{lllll}
+\\textbf{Parametric Terms}  & \\textbf{exp ($\\beta$)}  & \\textbf{std.er} & \\textbf{t} & \\textbf{p-value} \\\\ \\midrule
+Intercept                   & 0.00                      & 1.40             & -7.13       & \\textless{}0.001 \\\\ 
+Maize GDD                   & 1.00                      & 0.001            & ~4.15       & \\textless{}0.001 \\\\ 
+CD to Streams               & 0.67                      & 0.156            & -2.57       & ~~0.011           \\\\ 
+Protected                   & 4.82                      & 0.497            & ~3.16       & ~~0.002           \\\\ \\bottomrule
+                            &                           &                  &             &                   \\\\ 
+                            &                           &                  &             &                   \\\\ 
+\\textbf{Smooth Terms}      & \\textbf{edf}             & \\textbf{ref.df} & \\textbf{F} & \\textbf{p-value} \\\\ \\midrule
+s(Precipitation)            & 3.25                      & 3.25             & 11.1        & \\textless{}0.001 \\\\ \\bottomrule
+\\end{tabular}
+\\end{table}
+"
+)
+
+```
+
+```{r}
+
 here::here("R", "regression-table.R") |> source()
-
-parametric_table
-smooth_table
 

### 2023-05-01 manuscript/manuscript.tex
prepare manuscript for QI sub

@@
 
-\usepackage{booktabs}
-\usepackage{caption}
-\usepackage{longtable}
 \makeatletter
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\hypertarget{tbl-results}{}
 \begin{table}
-
-\caption{\label{tbl-results}Model
-Results}\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-results-1}Parametric Terms }
-
-{\centering
-
-\setlength{\LTpost}{0mm}
-\begin{longtable}{lcccc}
-\tabularnewline
-
-\toprule
- & \textbf{exp(Beta)} & \textbf{SE}\textsuperscript{\textit{1}} & \textbf{t} & \textbf{p-value} \\ 
-\midrule
-Intercept & 0.00 & 1.40 & -7.13 & <0.001 \\ 
-Maize GDD & 1.00 & 0.001 & 4.15 & <0.001 \\ 
-CD to Streams & 0.67 & 0.156 & -2.57 & 0.011 \\ 
-Protected & 4.82 & 0.497 & 3.16 & 0.002 \\ 
-\bottomrule
-\end{longtable}
-\begin{minipage}{\linewidth}
-\textsuperscript{\textit{1}}SE = Standard Error\\
-\end{minipage}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-\subcaption{\label{tbl-results-2}Smooth Terms }
-
-{\centering
-
-\begin{longtable}{lcccc}
-\tabularnewline
-
-\toprule
- & \textbf{edf} & \textbf{ref.df} & \textbf{F} & \textbf{p-value} \\ 
-\midrule
-s(Precipitation) & 3.25 & 3.25 & 11.1 & <0.001 \\ 
-\bottomrule
-\end{longtable}
-
-}
-
-\end{minipage}%
-
+\caption{\label{tbl-results}Model Results }\tabularnewline
+[]
+\centering
+\begin{tabular}{lllll}
+\textbf{Parametric Terms}  & \textbf{exp ($\beta$)}  & \textbf{std.er} & \textbf{t} & \textbf{p-value} \\ \midrule
+Intercept                   & 0.00                      & 1.40             & -7.13       & \textless{}0.001 \\ 
+Maize GDD                   & 1.00                      & 0.001            & ~4.15       & \textless{}0.001 \\ 
+CD to Streams               & 0.67                      & 0.156            & -2.57       & ~~0.011           \\ 
+Protected                   & 4.82                      & 0.497            & ~3.16       & ~~0.002           \\ \bottomrule
+                            &                           &                  &             &                   \\ 
+                            &                           &                  &             &                   \\ 
+\textbf{Smooth Terms}      & \textbf{edf}             & \textbf{ref.df} & \textbf{F} & \textbf{p-value} \\ \midrule
+s(Precipitation)            & 3.25                      & 3.25             & 11.1        & \textless{}0.001 \\ \bottomrule
+\end{tabular}
 \end{table}

### 2023-05-01 R/regression-table.R

@@ -52,21 +52,21 @@ smooth <- tbl_regression(
     ref.df = "**ref.df**"
   )
 
-parametric_table <- parametric |> 
-  as_gt() |> 
-  tab_options(
-    table.width = pct(50),
-    table.font.names = "Times New Roman"
-  ) |> 
-  as_latex()
+# parametric_table <- parametric |>
+#   as_gt() |>
+#   tab_options(
+#     table.width = pct(50),
+#     table.font.names = "Times New Roman"
+#   ) |>
+#   as_latex()
 
-smooth_table <- smooth |> 
-  as_gt() |> 
-  tab_options(
-    table.width = pct(50),
-    table.font.names = "Times New Roman"
-  ) |> 
-  as_latex()
+# smooth_table <- smooth |> 
+#   as_gt() |> 
+#   tab_options(
+#     table.width = pct(50),
+#     table.font.names = "Times New Roman"
+#   ) |> 
+#   as_latex()
 
 p <- parametric |> 
   as_tibble() |> 

### 2023-06-19 R/analysis.qmd
increase figure resolution and convert to jpg

@@
 title: "Supplement: Fremont Models"
-date: "`r Sys.Date()`"
+date: today
 format:
@@
 library(ggeffects)
-library(gratia)
 library(gt)
@@
   plot = for_print,
-  filename = here("figures", "covariates.png"),
-  width = 7,
-  height = 7 * (dy/dx),
-  dpi = 300
+  filename = here("figures", "covariates.jpg"),
+  width = 5.75,
+  height = 5.75 * (dy/dx),
+  dpi = 600
 )
@@
   plot = threshold_dists,
-  filename = here("figures", "threshold_distributions.png"),
-  width = 7,
-  height = 7 * (dy/(2*dx)),
-  dpi = 300
+  filename = here("figures", "threshold_distributions.jpg"),
+  width = 5.75,
+  height = 5.75 * (dy/(2*dx)),
+  dpi = 600
 )
@@
 
-# ggsave(
-#   plot = (obs + threshold_map),
-#   filename = here("figures", "threshold_map.png"),
-#   width = 7,
-#   height = 7 * (dy/(2*dx)),
-#   dpi = 300
-# )
-
 obs + threshold_map
@@
 
-```
-
-```{r}
-#| out-width: "80%"
-#| fig-align: center
-
 summary(fremont)
-
-appraise(fremont)
-
-draw(fremont, residuals = TRUE)
 
@@
 
-```
-
-```{r}
-#| out-width: "80%"
-#| fig-align: center
-
 summary(fremont$lme)
@@
 summary(fremont$gam)
-
-appraise(fremont$gam)
-
-draw(fremont$gam)
 
@@
 summary(fremont$gam)
-
-appraise(fremont$gam)
-
-draw(parametric_effects(fremont$gam))
-
-draw(fremont$gam)
 
@@
   bind_rows() |>
+  as_tibble() |> # the "ggeffects" class doesn't want to let go
   filter(group != "area") |>
@@
     linewidth = 2,
-    color = alpha('white', 0.5),
+    color = alpha('white', 0.75),
     lineend = "round"
@@
   plot = marginal_response,
-  filename = here("figures", "marginal_response.png"),
+  filename = here("figures", "marginal_response.jpg"),
   width = 5,
   height = 5 * 0.93,
-  dpi = 300
+  dpi = 600
 )
@@
   plot = (obs + estimate),
-  filename = here("figures", "model-map.png"),
-  width = 7,
-  height = 7 * (dy/(2*dx)),
-  dpi = 300
+  filename = here("figures", "model-map.jpg"),
+  width = 5.75,
+  height = 5.75 * (dy/(2*dx)),
+  dpi = 600
 )
@@
   plot = (estimate + threshold_map),
-  filename = here("figures", "threshold_map.png"),
-  width = 7,
-  height = 7 * (dy/(2*dx)),
-  dpi = 300
+  filename = here("figures", "threshold_map.jpg"),
+  width = 5.75,
+  height = 5.75 * (dy/(2*dx)),
+  dpi = 600
 )

### 2023-06-19 manuscript/manuscript.qmd
increase figure resolution and convert to jpg

@@
 bibliography: bibliography.bib
+linenumbers: false
+watermark: false
 format: 
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Quaternary International
-      formatting: preprint
-      model: 3p
+  simple-pdf: default
 execute: 
@@
 Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
-](../figures/overview.png){#fig-overview fig-align="center" width="6in"}
+](../figures/overview.jpg){#fig-overview fig-align="center" width="6in"}
 
@@
 See methods for how these values were calculated and results for more details.
-](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}
+](../figures/elevation-everything.jpg){#fig-elevation fig-align="center" width="4in"}
 
@@
 If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters.
-That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium.
-So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.
+In other words, it follows from their being at equilibrium that any observed differences in population size, $N_i$, must be the result of differences in $Q_i$.
+So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates, $X_i$, then use the resulting model to reconstruct $Q_i \sim f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.
 
@@
 The *relative* density refers to the density of feature-weighted sites ($D_i = N_i/area_i$) divided by the maximum density ($max\;D$) across all watersheds, thus scaling the density estimates to the unit interval [0,1].
-Using the relative density is just a way of avoiding any implication that we have an independent estimate of the absolute number of Fremont sites in the area, and not just an estimate based on a thinned sample.
+Using the relative density is just a visual tool to help avoid any implication that we have an independent estimate of the absolute number of Fremont sites in the area, and not just an estimate based on a thinned sample.
 
@@
 Distribution of covariates across watersheds. These include maize growing degree days (GDD), annual precipitation (precipitation), and cost-distance to streams (Streams).
-](../figures/covariates.png){#fig-covariates fig-align="center" width="6in"}
+](../figures/covariates.jpg){#fig-covariates fig-align="center" width="6in"}
 
@@
 Although we expect the outcome to be Poisson distributed, we use a negative binomial distribution here, as this includes a parameter, $r$, that can account for potential dispersion without giving up the inferential benefits of maximum likelihood.
-
 $$
@@
 Marginal response plots. Each plot shows the response of site counts to changes in a target covariate while holding all other variables at their mean.
-](../figures/marginal_response.png){#fig-responses fig-align="center" width="5in"}
+](../figures/marginal_response.jpg){#fig-responses fig-align="center" width="5in"}
 
@@
 3.
-While site counts exhibit a positive linear response to maize GDD, for maize has an upper threshold temperature at which point it no longer grows.
+While site counts exhibit a positive linear response to maize GDD in our model, this should not be expected to continue indefinitely for the obvious reason that maize has an upper threshold temperature at which point it no longer grows.
 Plus, increased temperatures trade-off with water availability, with warmer conditions being better for plant growth but at the same time increasing water demand on the plant to keep up with growth [@ramankutty2002].
-The linear relationship between site counts and maize GDD is, thus, relative to the observed distribution of that variable within the study area and should not be used to generalize to a larger geographic extent.
+The linear relationship between site counts and maize GDD is, thus, relative to the observed distribution of that variable within the study area and should not be used to generalize outside of that distribution.
 
@@
 Given the topography of the study area, this trade-off would also suggest that areas suitable for maize agriculture take the form of a thin band around the lower slopes of mountain ranges (or lower canyons in the Uintas).
-Such a band would be akin to a Goldilocks zone for maize agriculture (Yaworsky et al, this issue), a place where the optima for GDD and precipitation overlap, though we should not make too much of this idea, for it is an area of overlap for continuous ecological gradients, not a region demarcated by anything like a perfect line.
-It also must be emphasized that what matters most for maize agriculture is not necessarily where precipitation falls, but where it accumulates once it enters the streamflow network.
+Such a band would be akin to a Goldilocks zone for maize agriculture (Yaworsky et al, in press), a place where the optima for GDD and precipitation overlap.
+
+We should take care not to over-interpret this idea, however, for it is an area of overlap for continuous ecological gradients, not a region demarcated by anything like a perfect line.
+It should also be emphasized that what matters most for maize agriculture is not necessarily where precipitation falls, but where it accumulates once it enters the stream flow network.
 Of course, that will largely be a function of elevation in this region, or changes in elevation, so it will still be in proximity to the Goldilocks zone, though perhaps not perfectly coincident with it.
-At any rate, given these considerations, it would perhaps be more accurate to characterize the trade-off confronting the Fremont as one between what we might call water availability (including general precipitation across the streamflow network, water runoff, proximity to perennial surficial sources, and irrigation potential, among other things) and temperature (including, but not limited to, the length of the growing season and the number of frost-free days).
+At any rate, given these considerations, it would perhaps be more accurate to characterize the trade-off confronting the Fremont as one between what we might call water availability (including general precipitation across the stream flow network, water runoff, proximity to perennial surficial sources, and irrigation potential, among other things) and temperature (including, but not limited to, the length of the growing season and the number of frost-free days).
 
@@
 The light green color indicates those that are in the niche.
-](../figures/threshold_map.png){#fig-thresholds fig-align="center" width="6in"}
+](../figures/threshold_map.jpg){#fig-thresholds fig-align="center" width="6in"}
 
@@
 But setting aside those methodological issues, some much more interesting explanations may be offered, at least from the perspective of theory.
-In particular, the spatial covariance structure in our model may be picking up on the structure of the streamflow network, specifically the accumulation of water run-off as a function of slope.
+In particular, the spatial covariance structure in our model may be picking up on the structure of the stream flow network, specifically the accumulation of water run-off as a function of slope.
 This would suggest that if there are more people in one watershed, there should be more more people in the downstream watershed, too, at least relative to other watersheds with similar precipitation and GDD values.
-In the language of the IFD model, the suitability of the watershed habitat should propagate through the streamflow network.
-Exploring that intuition in a more systematic way, however, would require much more careful consideration of streamflow.
+In the language of the IFD model, the suitability of the watershed habitat should propagate through the stream flow network.
+Exploring that intuition in a more systematic way, however, would require much more careful consideration of stream flow.
 Unfortunately, that is well beyond the scope of the current paper.
@@
 The left panel shows the relative density derived from observed site counts, and the right panel shows the relative density from the estimated site counts.
-](../figures/model-map.png){#fig-model-map fig-align="center" width="6in"}
+](../figures/model-map.jpg){#fig-model-map fig-align="center" width="6in"}
 

### 2023-06-19 R/fig-elevation-tradeoffs.R

@@ -133,3 +133,10 @@ ggsave(
   height = 3,
   dpi = 300
 )
+
+ggsave(
+  here("figures", "elevation-everything.jpg"),
+  width = 4,
+  height = 3,
+  dpi = 600
+)

### 2023-06-19 R/fig-overview-map.R

@@ -200,7 +200,7 @@ site_density <- ggplot() +
   theme(
     legend.background = element_rect(fill = "gray95", color = "transparent"),
     legend.justification = c("left", "bottom"),
-    legend.position = c(0.65, 0.19)
+    legend.position = c(0.62, 0.17)
   )
 
 # Save --------------------------------------------------------------------
@@ -217,8 +217,8 @@ everything <- bob + site_density
 
 ggsave(
   plot = everything,
-  filename = here("figures", "overview.png"),
-  width = 7,
-  height = 7 * (dy/(2*dx)),
-  dpi = 300
+  filename = here("figures", "overview.jpg"),
+  width = 5.75,
+  height = 5.75 * (dy/(2*dx)),
+  dpi = 600
 )

### 2023-11-25 R/fig-overview-map.R
update overview map to include road and city landmarks

@@ -17,6 +17,7 @@
 library(ggfx)
 library(ggspatial)
 library(here)
+library(magick)
 library(patchwork)
 library(sf)
 library(tidyverse)
@@ -30,6 +31,11 @@ gpkg <- here("data", "western-fremont.gpkg")
 sys.source(
   here("R", "fun-get_basemap.R"),
   envir = attach(NULL, name = "basemap")
+)
+
+sys.source(
+  here("R", "fun-prepare-image.R"),
+  envir = attach(NULL, name = "image")
 )
 
 # features ----------------------------------------------------------------
@@ -48,12 +54,15 @@ utah <- states |> filter(name == "Utah")
 
 window <- read_sf(gpkg, "window")
 
-roads <- read_sf(gpkg, "roads")
+roads <- read_sf(gpkg, "roads") |> 
+  filter(rttyp %in% c("U", "I")) |> 
+  st_intersection(st_union(utah, window))
 
-roads <- st_intersection(
-  roads, 
-  st_union(utah, window)
-)
+towns <- tigris::places("Utah") |> 
+  rename_with(tolower) |> 
+  select(name) |> 
+  filter(name %in% c("Salt Lake City", "Cedar City", "Delta", "Richfield")) |> 
+  arrange(name)
 
 watersheds <- read_sf(gpkg, "watersheds")
 
@@ -66,8 +75,9 @@ dx <- bb8[["xmax"]] - bb8[["xmin"]]
 
 basemap <- get_basemap(
   bb8, 
-  map = "imagery",
-  size = c(9000, 9000*(dy/dx)),
+  map = "World_Imagery",
+  size = c(6000, 6000*(dy/dx)),
+  dpi = 600,
   imageSR = 26912
 )
 
@@ -87,6 +97,22 @@ bob <- ggplot() +
       basemap,
       bb8[["xmin"]], bb8[["xmax"]],
       bb8[["ymin"]], bb8[["ymax"]]
+    ),
+    mask = ch_alpha("utah-mask")
+  ) +
+  with_mask(
+    geom_sf(
+      data = roads |> filter(rttyp == "I"),
+      color = "#f7782f",
+      linewidth = 0.4
+    ),
+    mask = ch_alpha("utah-mask")
+  ) +
+  with_mask(
+    geom_sf(
+      data = roads |> filter(rttyp == "U"),
+      color = "#f7782f",
+      linewidth = 0.2
     ),
[…]

### 2023-11-25 R/fun-get_basemap.R

@@ -1,10 +1,9 @@
 
-get_basemap <- function(x,
-                        map = "physical",
-                        size = c(16000,9000),
-                        dpi = 900,
-                        imageSR = 4326,
-                        native = FALSE){
+get_basemap <- function(x, 
+                        map = "physical", 
+                        size = c(16000,9000), 
+                        dpi = 300,
+                        imageSR = 4326) {
 
   x <- st_bbox(x)
 
@@ -22,48 +21,28 @@ get_basemap <- function(x,
 
   }
 
-  base_url <- "http://services.arcgisonline.com/arcgis/rest/services/"
-
-  map_name <- switch(
-    map,
-    "hillshade"= "Elevation/World_Hillshade",
-    "dark"     = "Elevation/World_Hillshade_Dark",
-    "natgeo"   = "NatGeo_World_Map",
-    "usa"      = "USA_Topo_Maps",
-    "imagery"  = "World_Imagery",
-    "physical" = "World_Physical_Map",
-    "shaded"   = "World_Shaded_Relief",
-    "street"   = "World_Street_Map",
-    "terrain"  = "World_Terrain_Base",
-    "topo"     = "World_Topo_map",
-    stop(paste0("The ", map, " map is not supported."), call. = FALSE)
-  )
-
-  map_service <- paste0(base_url, map_name, "/MapServer/export?")
-
-  query <- list(
-    bbox    = paste(x, collapse = ","),
-    bboxSR  = st_crs(x)$epsg,
+  req <- httr2::request("http://services.arcgisonline.com/arcgis/rest/services")
+  
+  req <- httr2::req_url_path_append(req, map, "MapServer", "export")
+  
+  req <- httr2::req_url_query(
+    req,
+    bbox = paste(x, collapse = ","),
+    bboxSR = st_crs(x)$epsg,
     imageSR = imageSR,
-    f       = "image",
-    format  = "png",
-    size    = paste(size, collapse = ","),
-    dpi     = dpi
+    format = "png",
+    dpi = dpi,
+    size = paste(size, collapse = ","),
+    pixelType = "U8",
+    noDataInterpretation = "esriNoDataMatchAny",
+    interpolation = "+RSP_BilinearInterpolation",
+    f = "image"
   )
-
-  request <- httr::GET(
-    map_service,
-    query = query
-  )
-
-  result <- httr::content(request)
-
-  if (native) {
-
[…]

### 2023-11-25 R/fun-prepare-image.R
add function to clean up all figures, mainly stripping outside white space

@@ -1,0 +1,22 @@
+
+# trim all white space around image, but then add a few white pixels back (dx, dy)
+prepare_image <- function(x, dx = 2, dy = 30, color = "white") {
+  
+  img <- magick::image_read(path = x)
+  
+  img <- magick::image_trim(img)
+  
+  info <- magick::image_info(img)
+  
+  new_width <- info$width + dx
+  new_height <- info$height + dy
+  
+  img <- magick::image_extent(
+    img, 
+    magick::geometry_area(new_width, new_height), 
+    color = color
+  )
+  
+  magick::image_write(img, path = x)
+  
+}

### 2023-11-28 R/fig-overview-map.R
add landmarks to maps for easier orientation, update font size in legend

@@ -253,7 +253,7 @@ site_density <- ggplot() +
     linewidth = 0.1
   ) +
   scale_fill_viridis(
-    name = "Log of\nSite Density",
+    name = "Observed\nLog Density",
     option = "mako"
   ) +
   geom_sf(
@@ -277,7 +277,8 @@ site_density <- ggplot() +
   theme(
     legend.background = element_rect(fill = "gray95", color = "transparent"),
     legend.justification = c("left", "bottom"),
-    legend.position = c(0.62, 0.17)
+    legend.position = c(0.62, 0.17),
+    legend.title = element_text(size = rel(0.9))
   )
 
 # Save --------------------------------------------------------------------

### 2023-11-28 R/analysis.qmd
minor changes to formatting, mostly added code to define function `prepare_image()`

@@
     embed-resources: true
-    fig-align: center
-    fig-dpi: 300
-    fig-responsive: true
     number-sections: true
@@
 
+```{css}
+/*| echo: false */
+figcaption {
+    display: none !important;
+}
+```
+
 ## Overview
@@
 
-**Data**: (i) site counts per watershed based on records from the State Historic Preservation Offices of Utah and Nevada and (ii) environmental data from the Oregon State University PRISM Group, as well as the USGS.
+**Data**: (i) site counts per watershed based on records from the State Historic
+Preservation Offices of Utah and Nevada and (ii) environmental data from the
+Oregon State University PRISM Group, as well as the USGS.
 
-**Method**: a GAMM with a negative binomial distribution for count data with dispersion, an an offset to account for the area of each watershed, and an exponential covariance matrix to account for residual spatial autocorrelation.
+**Method**: a GAMM with a negative binomial distribution for count data with
+dispersion, an offset to account for the area of each watershed, and an
+exponential covariance matrix to account for residual spatial autocorrelation.
 
@@
 library(tidyverse)
+library(tigris)
 library(viridis)
+
+```
+
+Some plotting defaults
+
+```{r}
+#| code-fold: true
+#| file: "fun-prepare-image.R"
 
@@
 
-Dependent variable:  
+Dependent variable:\
 - a count of **sites** per watershed
 
-Independent variables:  
-- **elevation**  
-- maize growing degree days (**gdd**, in Celsius)  
-- **precipitation** (millimeters)  
-- cost-distance (in hours) to **springs** and **streams**  
-- **protected** status  
+Independent variables:\
+- **elevation**\
+- maize growing degree days (**gdd**, in Celsius)\
+- **precipitation** (millimeters)\
+- cost-distance (in hours) to **springs** and **streams**\
+- **protected** status\
 - cost-distance (in hours) to **roads**
 
-The climate variables, precipitation and gdd, were hindcasted for each watershed using Kyle Bocinsky's [paleocar](https://github.com/bocinsky/paleocar) package and averaged over the approximately 1,000 year occupational sequence of the Fremont in the project area.
-The cost-distance variables are derived from elevation data using Campbell's hiking function and Djikstra's search algorithm and averaged over each watershed.
-Protected status refers to the proportion of each watershed that is federal land.
-Note that it and cost-distance to roads are not explanatory variables, but rather controls on potential sources of sampling bias, mostly due to taphonomic processes operating on the archaeological record - basically, modern farms and cities.
+The climate variables, precipitation and gdd, were hindcasted for each watershed
+using Kyle Bocinsky's [paleocar](https://github.com/bocinsky/paleocar) package
+and averaged over the approximately 1,000 year occupational sequence of the
+Fremont in the project area. The cost-distance variables are derived from
+elevation data using Campbell's hiking function and Djikstra's search algorithm
+and averaged over each watershed. Protected status refers to the proportion of
+each watershed that is federal land. Note that it and cost-distance to roads are
+not explanatory variables, but rather controls on potential sources of sampling
+bias, mostly due to taphonomic processes operating on the archaeological
+record - basically, modern farms and cities.
 
-Offsets:  
+Offsets:\
 - **area** of each watershed
 
-This is included to account for sampling bias owing to the size of each watershed.
-The other two are there to account for sampling bias owing to variable survey intensity, as well as the potential for human impacts to cultural resources.
+This is included to account for sampling bias owing to the size of each
+watershed. The other two are there to account for sampling bias owing to
+variable survey intensity, as well as the potential for human impacts to
+cultural resources.
 
-**Note:** not all of these variables make it into the final model as we first evaluate them for potential colinearity and concurvity.
-An additional concurvity test is performed after fitting the final model.
-Not all of them have smooths applied in the final model, either.
+**Note:** not all of these variables make it into the final model as we first
+evaluate them for potential colinearity and concurvity. An additional concurvity
+test is performed after fitting the final model. Not all of them have smooths
+applied in the final model, either.
 
@@
 
+watersheds <- read_sf(gpkg, "watersheds")
+
+# the following shapes are just for visualization purposes
 utah <- read_sf(gpkg, "utah")
 
-watersheds <- read_sf(gpkg, "watersheds") |> 
-  mutate(
-    huc4 = str_sub(id, 1, 4),
-    huc4 = factor(huc4)
-  )
+window <- read_sf(gpkg, "window")
+
+roads <- read_sf(gpkg, "roads") |> 
+  filter(rttyp %in% c("U", "I")) |> 
+  st_intersection(window)
+
+towns <- tigris::places("Utah", progress_bar = FALSE) |> 
+  rename_with(tolower) |> 
+  select(name) |> 
+  filter(name %in% c("Salt Lake City", "Cedar City", "Delta", "Richfield")) |> 
+  arrange(name) |> 
+  st_centroid()
 
@@
 
-```{r}
-#| fig-height: 4
-#| code-fold: true
-#| fig-align: center
+::: {#fig-overview}
+![](../figures/overview.jpg){width="100%" fig-alt="overview map of project area"}
 
-max_density_obs <- with(watersheds, max(sites/area))
-
-obs <- ggplot() +
-  geom_sf(
-    data = utah,
-    fill = "gray95"
-  ) +
-  geom_sf(
-    data = watersheds, 
-    aes(fill = (sites/area)/max_density_obs),
-    color = "gray90", 
-    linewidth = 0.2
-  ) +
-  scale_fill_viridis(
-    name = "Relative\nSite Density",
-    option = "mako",
-    limits = c(0, 1)
-  ) +
-  # coord_sf(expand = FALSE) +
-  theme_void() +
-  theme(
-    legend.background = element_rect(fill = "gray95", color = "transparent"),
-    legend.justification = c("left", "top"),
-    legend.position = c(0.66, 0.57)
-  )
-
-obs
-
-```
-
-This map shows the relative density of sites based on their observed frequency in the study area.
-If site density is the ratio of the number of sites in a watershed to the area of that watershed, the _relative_ density is the density estimate for each watershed divided by the maximum density estimate across all watersheds.
-It is more appropriate to visualize this value as it doesn't have the same strong implications as visualizing the counts.
-(1) By visualizing the _density_, it does not make assumptions about the absence of sampling bias owing to the area of each watershed.
-(2) By visualizing the _relative_ density, it does not make assumptions about the total occurrence rate of sites across the study area, i,e, it does not include an estimate of the intercept.
+Overview map.
+:::
 
@@
 ```{r}
+#| label: fig-correlations
+#| code-fold: true
+#| out-width: "75%"
 #| fig-height: 5
 #| fig-asp: 1
-#| code-fold: true
-#| out-width: "75%"
-#| fig-align: center
+#| fig-alt: "Pearson correlation matrix"
+#| fig-cap: "Pearson correlation matrix"
 
@@
 
-As you can see, there are strong, linear correlations between all the variables, including the familiar inverse relationship between precipitation and gdd, which is driven by elevation in the American West.
-Based on this and a previous post-hoc concurvity test, we will drop springs.
+As you can see, there are strong, linear correlations between all the variables,
+including the familiar inverse relationship between precipitation and gdd, which
+is driven by elevation in the American West. Based on this and a previous
+post-hoc concurvity test, we will drop springs.
 
@@
 ```{r}
+#| label: fig-covariates
+#| code-fold: true
 #| fig-width: 8
 #| fig-asp: 0.9
-#| code-fold: true
-#| fig-align: center
+#| fig-alt: "Map of covariates"
+#| fig-cap: "Map of covariates"
 
-make_map <- function(fill, name) {
+make_map <- function(x, fill, name) {
 
@@
     geom_sf(
-      data = watersheds, 
-      aes(fill = {{fill}}),
+      data = x, 
+      aes(fill = {{ fill }}),
       color = "gray90", 
@@
     ) +
+    geom_sf(
+      data = roads |> filter(rttyp == "I"),
+      color = "white",
+      linewidth = 0.4
+    ) +
+    geom_sf(
+      data = roads |> filter(rttyp == "U"),
+      color = "white",
+      linewidth = 0.2
+    ) +
+    geom_sf(
+      data = towns,
+      fill = "white",
+      color = "gray35",
+      shape = 21,
+      size = 2
+    ) +
     theme_void() +
@@
       legend.justification = c("left", "bottom"),
-      legend.position = c(0.65, 0.19)
+      legend.position = c(0.65, 0.19),
+      legend.title = element_text(size = rel(0.9))
     )
@@
 
-for_print <- 
-  (make_map(gdd, "GDD (°C)") + make_map(precipitation, "PPT (mm)")) /
-  (make_map(streams, "Streams\n(hours)") + make_map(protected, "Protected"))
+gdd <- make_map(
+  watersheds,
+  fill = gdd,
+  name = "GDD (°C)"
+)
+
+ppt <- make_map(
+  watersheds,
+  fill = precipitation,
+  name = "PPT (mm)"
+)
+
+streams <- make_map(
+  watersheds,
+  fill = streams,
+  name = "Streams\n(hours)"
+)
+
+protected <- make_map(
+  watersheds,
+  fill = protected, 
+  name = "Protected"
+)
 
@@
 
+fn <- here("figures", "covariates.jpg")
+
 ggsave(
-  plot = for_print,
-  filename = here("figures", "covariates.jpg"),
+  plot = (gdd + ppt) / (streams + protected),
+  filename = fn,
   width = 5.75,
@@
 
-(make_map(gdd, "GDD (°C)") + 
-  make_map(precipitation, "PPT (mm)") +
-  make_map(streams, "Streams\n(hours)")) /
-  (make_map(springs, "Springs\n(hours)") +
-     make_map(protected, "Protected") +
-     make_map(roads, "Roads\n(hours)"))
+prepare_image(fn)
+
+springs <- make_map(
+  watersheds,
+  fill = springs, 
+  name = "Springs\n(hours)"
+)
+
+routes <- make_map(
+  watersheds,
+  fill = roads, 
+  name = "Roads\n(hours)"
+)
+
+all_covariates <- (gdd + ppt + streams) / (springs + protected + routes)
+
+remove(gdd, ppt, streams, springs, routes, protected)
+
+all_covariates
 
@@
 ```{r}
+#| label: fig-thresholds
+#| code-fold: true
 #| fig-width: 8
 #| fig-asp: 0.45
-#| code-fold: true
-#| fig-align: center
+#| fig-alt: "Histograms of temperature and precipitation"
+#| fig-cap: "Histograms of temperature and precipitation"
 
@@
 
-threshold_dists <- ggplot() +
+ggplot() +
   geom_histogram(
@@
 
-ggsave(
-  plot = threshold_dists,
-  filename = here("figures", "threshold_distributions.jpg"),
-  width = 5.75,
-  height = 5.75 * (dy/(2*dx)),
-  dpi = 600
-)
-
-threshold_dists
-
 ```
@@
 ```{r}
-#| fig-width: 8
-#| fig-asp: 0.45
+#| label: fig-thresholds-map
 #| code-fold: true
-#| fig-align: center
+#| fig-width: 4
+#| out-width: "4in"
+#| fig-alt: "Map of maize niche thresholds"
+#| fig-cap: "Map of maize niche thresholds"
 
-threshold_map <- ggplot() +
+thm <- ggplot() +
   geom_sf(
@@
   ) +
+  geom_sf(
+    data = roads |> filter(rttyp == "I"),
+    color = "white",
+    linewidth = 0.4
+  ) +
+  geom_sf(
+    data = roads |> filter(rttyp == "U"),
+    color = "white",
+    linewidth = 0.2
+  ) +
+  geom_sf(
+    data = towns,
+    fill = "white",
+    color = "gray35",
+    shape = 21,
+    size = 2
+  ) +
   theme_void() +
@@
 
-obs + threshold_map
+thm
 
@@
 
-Is there a significant difference in elevation between those in the niche and those outside it?
+Is there a significant difference in elevation between those in the niche and
+those outside it?
 
@@
 
-Here, we go through iterations of model fitting and evaluation, with changes to correct for problems like autocorrelation and concurvity (the smoothed version of co-linearity in the predictors).
-For each fitted model, we provide a model summary, diagnostic plots, and visualizations of partial effects (sometimes called marginal responses) for each smooth term.
+Here, we go through iterations of model fitting and evaluation, with changes to
+correct for problems like autocorrelation and concurvity (the smoothed version
+of co-linearity in the predictors). For each fitted model, we provide a model
+summary, diagnostic plots, and visualizations of partial effects (sometimes
+called marginal responses) for each smooth term.
 
@@
 
-First, an h-test for scaled dispersion in a poisson model.
-In a poisson model, the variance is supposed to equal the mean.
-If this holds, then $\alpha$ in $Var(y) = E(y) + \alpha \cdot E(y)$ should equal zero.
-A simple, intercept-only linear model can test this idea.
+First, an h-test for scaled dispersion in a poisson model. In a poisson model,
+the variance is supposed to equal the mean. If this holds, then $\alpha$ in
+$Var(y) = E(y) + \alpha \cdot E(y)$ should equal zero. A simple, intercept-only
+linear model can test this idea.
 
@@
 
-Clearly, there is over-dispersion in this model.
-To correct for this, we re-run the model with a `negative binomial` error distribution, letting the model adjust to $\alpha$.
-We choose a negative binomial over a quasipoisson as it uses maximum likelihood.
+Clearly, there is over-dispersion in this model. To correct for this, we re-run
+the model with a `negative binomial` error distribution, letting the model
+adjust to $\alpha$. We choose a negative binomial over a quasipoisson as it uses
+maximum likelihood.
 
@@
 
-Monte Carlo simulations of Moran's I to test for spatial autocorrelation in the residuals (using sfdep).
+Monte Carlo simulations of Moran's I to test for spatial autocorrelation in the
+residuals (using sfdep).
 
@@
 
-There's significant spatial autocorrelation.
-To fix this, we will use a generalized additive mixed-model (GAMM) with an exponential spatial correlation structure.
-This is not strictly correct, as it assumes spatial continuity of points, not polygons, but it's a useful first approximation.
+There's significant spatial autocorrelation. To fix this, we will use a
+generalized additive mixed-model (GAMM) with an exponential spatial correlation
+structure. This is not strictly correct, as it assumes spatial continuity of
+points, not polygons, but it's a useful first approximation.
 
@@
 
-By accounting for spatial autocorrelation, the expected degrees of freedom for gdd and cost-distance to streams are now both 1, which suggests that there are no non-linear effects for those covariates.
+By accounting for spatial autocorrelation, the expected degrees of freedom for
+gdd and cost-distance to streams are now both 1, which suggests that there are
+no non-linear effects for those covariates.
 
@@
 
-Cost-distance to roads is non-significant, so we'll remove that variable from the model entirely.
+Cost-distance to roads is non-significant, so we'll remove that variable from
+the model entirely.
 
@@
 
-One more Moran's I test to see if we really got a hold of the spatial autocorrelation.
-We'll do this with the normalized residuals of the lme inside the GAMM as that incorporates the exponential covariance matrix.
+One more Moran's I test to see if we really got a hold of the spatial
+autocorrelation. We'll do this with the normalized residuals of the lme inside
+the GAMM as that incorporates the exponential covariance matrix.
 
@@
 
-There can't be concurvity in the model anymore, as there is only one smoothed term. However, we should now check for potential multi-collinearity in the parameteric terms.
+There can't be concurvity in the model anymore, as there is only one smoothed
+term. However, we should now check for potential multi-collinearity in the
+parameteric terms.
 
@@
 
-Here we estimate the marginal responses (also known as partial effects) for each covariate by letting the target covariate vary and holding the other covariates at their means and predicting the site count (the response).
-Here we use the {ggeffects} package.
+Here we estimate the marginal responses (also known as partial effects) for each
+covariate by letting the target covariate vary and holding the other covariates
+at their means and predicting the site count (the response). Here we use the
+{ggeffects} package.
 
 ```{r}
-#| code-fold: true
-#| fig-align: center
-#| fig-height: 6
-#| fig.asp: 0.93
-#| out-width: "70%"
 
-margins <- ggpredict(fremont) |>
+margins <- ggpredict(fremont, condition = list(area = mean(watersheds$area))) |>
   unclass() |>
@@
     "variable" = group
-  ) |> 
+  )
+
+```
+
+```{r}
+#| label: fig-partials
+#| code-fold: true
+#| fig-height: 6
+#| fig.asp: 0.93
+#| out-width: "70%"
+#| fig-alt: "Partial dependence plots"
+#| fig-cap: "Partial dependence plots"
+
+margins <- margins |> 
   mutate(
@@
 
+fn <- here("figures", "marginal_response.jpg")
+
 ggsave(
   plot = marginal_response,
-  filename = here("figures", "marginal_response.jpg"),
+  filename = fn,
   width = 5,
@@
 )
+
+prepare_image(fn)
 
@@
 ```{r}
+#| label: fig-model-map
+#| code-fold: true
 #| fig-width: 8
 #| fig-asp: 0.9
-#| code-fold: true
-#| fig-align: center
+#| fig-alt: "Map of estimated density"
+#| fig-cap: "Map of estimated density"
 
-basemap <- ggplot() +
-  geom_sf(
-    data = utah,
-    fill = "gray95"
-  ) +
-  theme_void() +
-  theme(
-    legend.background = element_rect(fill = "gray95", 
-                                     color = "transparent"),
-    legend.justification = c("left", "bottom"),
-    legend.position = c(0.65, 0.19)
+obs <- make_map(
+  watersheds,
+  fill = ifelse(sites == 0, log(0.00001), log(sites/area)), 
+  name = "Observed\nLog Density"
 )
 
-obs <- basemap +
-  geom_sf(
-    data = watersheds, 
-    aes(fill = (sites/area)/max_density_obs),
-    color = "gray90", 
-    linewidth = 0.1
-  ) +
-  scale_fill_viridis(
-    name = "Observed",
-    option = "mako",
-    limits = c(0, 1)
+est <- make_map(
+  watersheds,
+  log(estimate/area), 
+  "Estimated\nLog Density"
 )
 
-max_density_est <- with(watersheds, max(estimate/area))
-
-estimate <- basemap +
-  geom_sf(
-    data = watersheds, 
-    aes(fill = (estimate/area)/max_density_est),
-    color = "gray90", 
-    linewidth = 0.1
-  ) +
-  scale_fill_viridis(
-    name = "Estimated",
-    option = "mako",
-    limits = c(0, 1)
+res <- make_map(
+  watersheds,
+  fill = residuals, 
+  name = "Normalized\nResiduals"
 )
 
-res <- basemap +
-  geom_sf(
-    data = watersheds, 
-    aes(fill = residuals),
-    color = "gray90", 
-    linewidth = 0.1
-  ) +
-  scale_fill_viridis(
-    name = "Residuals",
-    option = "mako"
-  )
+fn <- here("figures", "model-map.jpg")
 
 ggsave(
-  plot = (obs + estimate),
-  filename = here("figures", "model-map.jpg"),
+  plot = (obs + est),
+  filename = fn,
   width = 5.75,
@@
 
+prepare_image(fn)
+
+fn <- here("figures", "threshold_map.jpg")
+
 ggsave(
-  plot = (estimate + threshold_map),
-  filename = here("figures", "threshold_map.jpg"),
+  plot = (est + thm),
+  filename = fn,
   width = 5.75,
@@
 
-(obs + estimate) / (res + threshold_map)
+prepare_image(fn)
+
+(obs + est) / (res + thm)
 
@@
 
-Please note that the residuals are still being shown for the count-level data.
+Please note that these are normalized residuals from the LME model.
 

### 2024-01-31 R/data-wrangling.R
moved fremont.gpkg into .misc/ folder

@@ -1,22 +1,5 @@
 
 # Site distribution for the western Fremont -- Data Wrangling
-# Vernon 2022
-
-##############################.
-#     Table of Contents
-# 01. Preamble
-# 02. Utah polygon
-# 03. Watersheds 
-# 04. Sites
-# 05. Digital Elevation Model (DEM)
-# 06. Land ownership
-# 07. Water features
-# 08. Roads
-# 09. PRISM
-# 10. Tree Rings (ITRDB)
-# 11. Cost-distance
-# 12. Extract all data into watersheds 
-##############################.
 
 # Note:
 # 1- Sensitive site data cannot be shared directly, so you'll need to get it
@@ -116,7 +99,7 @@ remove(bob)
 # 04) SITES ---------------------------------------------------------------
 
 sites <- read_sf(
-  "E:/fremont.gpkg", # <--- can't be run without site data
+  here(".misc", "fremont.gpkg"), # <--- can't be run without site data
   layer = "centroids"
 ) |> 
   st_filter(window)

### 2024-01-31 R/data-wrangling.R
don't source functions into opaque environment

@@ -23,19 +23,10 @@ gpkg <- here("data", "western-fremont.gpkg")
 # coordinate reference 
 # EPSG:26912
 
-# load custom functions
-# doing it this way to mask the functions in the global environment
-sys.source(
-  here("R", "fun-download_features.R"), # requires httr and jsonlite
-  envir = attach(NULL, name = "features")
-)
+# requires httr and jsonlite
+here("R", "fun-download_features.R") |> source()
 
-sys.source(
-  here("R", "fun-extractors.R"),
-  envir = attach(NULL, name = "extractors")
-)
-
-# use ls(features) or ls(extractors) to see the functions
+here("R", "fun-extractors.R") |> source()
 
 # 02) UTAH POLYGON --------------------------------------------------------
 

### 2024-01-31 manuscript/cover-letter-response-to-reviews.md
draft response to reviewers

We thank the reviewers for their many insightful comments and criticisms
regarding our paper.

All minor comments about grammatical errors, poor wording, and the like have
been addressed in the text. Major comments from reviewers are listed here with
comments about how we addressed them in the text. Reviewer two had the most
substantial comments and criticisms, so most of this involves responses to them.

### Reviewer #1

1.  Description of Fremont extent is misleading and vague.

    Response: We updated the description of the Fremont extent in the background
    section to make it clearer that there is evidence for Fremont foragers in ID
    and WY and Fremont farmers in NV and CO, rather than leaving it vague
    whether they might have all been farmers.

2.  Description of project area is inaccurate.

    Response: We updated this section to fix inaccurate descriptions. We also
    updated the overview map, adding major roads and some important nearby
    cities to give the reader more landmarks for orientation.

3.  The impossibility of dry farming does not necessarily entail irrigation.

    Response: This is correct, so we have added text to the discussion to
    clarify. In the subsection title, we also replaced 'irrigation' with 'water
    management', and in the text, we change "making irrigation ... necessary" to
    "making irrigation or some other form of water management ... necessary."

### Reviewer #2

1.  Why not high precipitation?

    Response: Our answer to this involves the trade-off with temperature in the
    project area. As elevation increases, precipitation increases but
    temperature decreases. So, the places with the most precipitation are also
    the ones most likely to experience frost and to have shorter growing
    seasons. This means the best place for maize is not necessarily the place
    with the most precipitation. We've updated the discussion, rearranging some
    comments into a new section entitled "Trade-off Between Precipitation and
    Temperature," to emphasize this point.

2.  The limits of rain-fed maize farming should be discussed more explicitly.

    Response: Our analysis of the Fremont maize farming niche is suggestive of
    its limits, but defining the true limits requires actualistic research,
    hence our reluctance to offer specific numbers on this subject. We do,
    however, discuss the results of some actualistic research, particularly the
    work referenced by Bocinsky in his 2014, and compare our results to that.
    This is the basis for our claim that the Fremont likely lived outside of the
    rain-fed maize farming niche, thus requiring some kind of water management
    on their part.

3.  Details of data collection are too vague.

    Response: We updated the Site Data section to more accurately represent how
    we came by these data.

    -   The NSF project alluded to in the previous version of this paper now
        includes its award number along with several citations to research
        coming out of that project.
    -   The "stratified random sample" comment was a bit misleading. That was
        done for a portion of the sites (we sampled a fraction of site forms in
        each county), but not systematically to produce the entire sample, which
        was mostly arrived at in a more opportunistic manner.
    -   To get room counts, we used the list of Fremont villages found in
        Coltrain and Leavitt (2002), and cross-referenced those with the
        original site forms, excavation reports, and Adrien Mooney's 2014 BYU MA
        thesis, which tries to reconstruct some poorly reported excavations. For
        many of these sites, the estimates can vary wildly, so we erred on the
        side of choosing the more conservative estimates.
    -   In spite of these limitations with our data collection scheme, we note
        that the current dataset represents one of the largest Fremont samples
        used for this sort of analysis to date.

4.  What kind of sites are these?

    Response: The reviewer requests additional information about site types, but
    we do not currently have the data to answer that question. Mainly, our data
    collection efforts were focused on identifying as many Fremont related sites
    as possible. In fact, we have done this now for a considerable portion of
    the state of Utah, though efforts are ongoing. We are also trying to go back
    and record as much attribute data as is available and will hopefully be able
    to report on these efforts in the near future.

5.  How do we know that these are farming and not foraging adaptations?

    Response: The reviewer notes that site types are needed to differentiate
    hunting sites from farming sites, and thus to determine whether settlement
    decisions were shaped by constraints on farming vs constraints on foraging.
    This is a really important point that goes right to the heart of our model
    interpretation. If the sites in our analysis are not primarily associated
    with farmers, then we cannot use them to understand the maize farming niche.
    Unfortunately, there's just not a lot of good data to really answer this
    question. The available stable isotope data that exists for the Fremont
    suggests that they were, for the most part, consuming maize at a rate
    comparable to maize farmers in the San Juan Basin, including Chaco Canyon.
    See, for example, Coltrain and Leavitt 2002 and Coltrain 1993. As Coltrain
    and Leavitt point out, the one exception to this general pattern is the
    Fremont around Bear River and the Willard Bay wetlands, but even there
    evidence of a largely hunting based diet is limited to the last few
    centuries in their sequence (around 900 CE their subsistence pattern becomes
    more mixed, and only around 1100 do they fully adopt foraging). This is not
    to say that foraging wasn't important to the Fremont. Just that farming was
    likely the critical constraint on settlement for most Fremont for most of
    their history (Yaworsky 2023; Spangler 2000, 2002).

6.  The Ideal Free Distribution section is overly detailed.

    Response: The reviewer thinks the discussion of the IFD model is overly
    detailed because "the analysis doesn't rank habitat suitability or site
    association with suitability." The reviewer suggests a simpler discussion
    that contextualizes the analysis "within the literature on SDMs." In
    response, we would like to note three things.

    -   First, we agree that the section was overly detailed. In fact, the main
        logic we wanted to convey is already in the intro, so the section is not
        just overly detailed but basically unnecessary, so we have removed it
        entirely.

    -   Second, distressingly, it's precisely the goal of this paper to estimate
        relative differences in habitat suitability across watersheds and
        environmental gradients, but the reviewer's comment suggests that we did
        not do a good job of making this clear. So, we have updated the
        introduction (particularly, the last paragraph) and the discussion to
        make this goal more explicit.

    -   Third, as the reviewer requested, we also explicitly connected this to
        the SDM literature by noting that ecologists will sometimes refer to
        SDMs as "Habitat Suitability Models" (e.g., Rowden et al 2017; Kort et
        al 2020; Bowden et al 2021).

7.  Goldilocks Zone concept is misleading.

    Response: The title of the subsection was changed to "Trade-off Between
    Precipitation and Temperature." The expression is now described explicitly
    as a "metaphor," and it's made clear that the reader should not
    over-interpret it for the reasons listed by the reviewer. We have also
    removed all but a single reference to it.

### Reviewer #4

1.  There are several reference errors and omissions.

    Response: These have been corrected. The more recent references to Fremont
    scholarship offered by the reviewer have also been added.

2.  But are they optimizers?

    Response: This paper is not meant to be a defense of the optimality
    assumption, just an example of its application, so we would prefer not to
    litigate that question here. We simply propose that Fremont farmers located
    themselves in places that yield the highest probabilities of a successful
    harvest. This assumption only requires that the Fremont had an intimate
    knowledge of their local ecology and leveraged this knowledge to make
    intelligent decisions regarding the placement of farming sites. Some of the
    wording in the introduction was changed to emphasize this.

3.  Trade-offs between spatial and temporal resolution?

    Response: We added an entire section to the discussion on this specific
    topic. It has the title "The Data Quality Problem in Fremont Research."
    Hopefully, it helps to assuage the reviewer's concern about a lack of
    temporal resolution in our data, which we fully admit is an issue. While
    having precise chronological control would be preferable, we argue that our
    results are not meaningless because "typically better watersheds," meaning
    watersheds that were warmer and wetter for most of the Fremont sequence,
    "would have attracted maize farmers more often and, over time, would have
    accumulated a larger archaeological maize farming assemblage, too." So, the
    results aren't perfect, but they are at least suggestive of the real drivers
    of Fremont residential choice behavior.

### 2024-01-31 manuscript/manuscript.qmd
all changes made to meet reviewer concerns

--- 2023-05-01 manuscript/manuscript.qmd
@@
           8000 Aarhus C, Denmark
-  - name: Weston McCool
+  - name: Weston C. McCool
     affiliations:
@@
 abstract: |
-  The Fremont provide an important case study to examine the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear.
-Using a tree-ring and simulation-based reconstruction of average annual precipitation and maize growing degree days, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin.
+  The Fremont provide an important case study for examining the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear.
+Using a tree-ring and simulation-based reconstruction of average annual precipitation and temperature, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin.
 The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
@@
   - Paleoclimate Reconstruction
+corresponding_author: Kenneth.Vernon@colorado.edu
+linenumbers: false
 bibliography: bibliography.bib
 format: 
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Quaternary International
-      formatting: preprint
-      model: 3p
+  simple-pdf: default
+  docx: default
 execute: 
@@
 
-Archaeological populations of subsistence maize farmers known collectively as "the Fremont" lived at the far northern periphery of maize farming in western North America, in an area encompassing much of the modern state of Utah north of the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989; @madsen1998].
-That the Fremont managed to thrive for as long as they did in this part of the world poses an interesting question, for conditions there are extremely harsh and inhospitable for maize farming.
-Archaeologists interested in that question have typically focused on strategies directly related to farming and subsistence, like irrigation methods, trade, and seasonal high elevation hunting [e.g., @barlow2008; @boomgarden2019; @janetski2002; @hart2021; @madsen1998; @metcalfe1985; @patterson2010; @spangler1993; @morgan2012].
-Here, however, we come at the problem from a slightly different angle.
-We seek to understand not so much *what* they did to make farming effective, but *where* they chose to do it, and why they chose those places over others [e.g., @bocinsky2014; @thomson2020].
-Those are not entirely separate questions, of course, as they trade-off each other, but we think a geographic analysis can actually help illuminate some of the costs and benefits of the other strategies adopted by the Fremont.
-A geographic or spatial model will also help us to investigate the ways that climate change might have structured those costs and benefits, and by extension the settlement decisions of Fremont farmers, which we consider a long-term record of human adaptation to climate change.
+Archaeological populations of subsistence maize farmers known collectively as
+"the Fremont" lived at the far northern periphery of maize farming in western
+North America, in an area encompassing much of the modern state of Utah north of
+the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989;
+@madsen_simms1998]. Conditions in this area are extremely harsh and inhospitable
+to maize farming, so the fact that the Fremont not only pulled it off but also
+thrived while doing so poses an interesting puzzle. How did they do it?
+Archaeologists interested in this question have typically focused on strategies
+directly related to farming and subsistence, like irrigation methods, trade, and
+seasonal high elevation hunting [e.g., @barlow_etal2008; @boomgarden_etal2019;
+@janetski2002; @hart_etal2021; @madsen_simms1998; @metcalfe_larrabee1985;
+@patterson_flanigan2010; @spangler1993; @morgan_etal2012]. Here, however, we
+come at the problem from a slightly different angle. We seek to understand not
+so much *what* they did to make farming effective, but *where* they chose to do
+it, and why they chose those places over others [e.g., @bocinsky_kohler2014;
+@thomson_macdonald2020]. Those are not entirely separate questions, of course,
+as they trade-off each other, but we think a geographic analysis can actually
+help illuminate some of the costs and benefits of the other strategies adopted
+by the Fremont. A geographic approach will also help us to investigate the ways
+that climate change might have structured those costs and benefits, and by
+extension the settlement decisions of Fremont farmers.
 
-As a test case, we focus on those Fremont living in the eastern Great Basin of western Utah, which we refer to as the western Fremont (see @fig-overview).
-The extreme desert conditions that the western Fremont adapted to provide a useful backdrop for exploring Fremont settlement, for variation in that environment exhibits abrupt and quite dramatic changes over short distances of space and time, thus heightening the differences between areas the western Fremont did and did not occupy.
-We explore those differences from a socio-ecological perspective, focusing on dynamic interactions between competing individuals on the one hand and individuals and their environment on the other [@bird2006; @kennett2006a].
-This dynamic is captured nicely by the Ideal Free Distribution model from Behavioral Ecology, as it represents individual decisions about where to live as a choice between habitats whose benefits to the individual may be constrained by what other individuals are doing.
-The technical term for these benefits is 'suitability', which the IFD defines as some function of both the habitat's intrinsic or pristine environmental quality and its current population size [@fretwell1969; @codding2015; @winterhalder2010].
-The fundamental assumption underlying the IFD is that individuals will behave optimally, that they will choose to settle the habitat with the highest suitability first, meaning the one that offers them the greatest ratio of benefits to costs relative to the available alternatives.
+As a test case, we focus on those Fremont living in the eastern Great Basin of
+western Utah, which we refer to as the western Fremont (see @fig-overview). The
+extreme desert conditions that the western Fremont adapted to provide a useful
+backdrop for exploring Fremont settlement, for variation in that environment
+exhibits abrupt and quite dramatic changes over short distances of space and
+time, thus heightening the differences between areas the western Fremont did and
+did not occupy. We explore those differences from a socio-ecological
+perspective, focusing on dynamic interactions between competing individuals on
+the one hand and individuals and their environment on the other
+[@bird_oconnell2006; @kennett_etal2006]. This dynamic is captured nicely by the
+Ideal Free Distribution model from Behavioral Ecology, as it represents
+individual decisions about where to live as a choice between habitats whose
+benefits to the individual may be constrained by what other individuals are
+doing. The technical term for these benefits is 'suitability', which the IFD
+defines as some function of both the habitat's intrinsic or pristine
+environmental quality and its current population size [@fretwell_lucas1969;
+@codding_bird2015; @winterhalder_etal2010].
 
-Assuming that western Fremont individuals were adept at responding to variation in their local ecology, in particular, that they optimized their choice of habitats, we can use an empirical or inductive species distribution model of Fremont sites to reconstruct the ecological niche for maize farming [@vernon2021; @yaworsky2020; @elith2009].
-This is sometimes referred to as ecological niche modeling.
-The basic idea here is that we can look at the distribution of Fremont sites as a reliable indicator of the sorts of available environmental conditions that would best promote maize farming in an arid landscape.
-In effect, we are helping ourselves to a little reverse engineering of the archaeological record [@dennett1995].
-We are assuming that whatever strategy - whatever habitat - the Fremont chose was the best alternative available to them, not in any absolute sense, of course, just the best given the constraints and trade-offs Fremont farmers faced in those environments [@jochim2022].
-In other words, we are treating their settlement decisions themselves as maize farming adaptations.
+The fundamental assumption underlying the IFD is that individuals will behave
+optimally, that they will choose to settle the habitat with the highest
+suitability first, meaning the one that offers them the greatest ratio of
+benefits to costs relative to the available alternatives. Of course, this comes
+with the usual caveats. The claim is not that individuals are infallible, that
+they will always do exactly the right thing, but that their choices will tend to
+approximate the best strategy given their constraints and trade-offs.
 
-![
-On the left is an overview map showing the project area with modern satellite imagery.
-Dark green colors represent high elevation, high precipitation areas.
-The large beige area in the northwest is the Bonneville Basin.
-What remains of the Great Salt Lake can be seen just west of Salt Lake City.
-For visualization purposes, the map on the right shows the relative density of archaeological sites (see the section on site data below for details on how we calculate relative density).
-Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
-](../figures/overview.png){#fig-overview fig-align="center" width="6in"}
+Assuming the western Fremont optimized their agricultural practices, we can use
+an empirical or inductive species distribution model [SDM, @elith_etal2006;
+@elith_leathwick2009] of Fremont sites to estimate relative differences in maize
+suitability across specific environmental gradients and geographic locations
+[@jochim2022; @vernon_etal2022a; @yaworsky_etal2020; @elith_leathwick2009]. In
+ecology, you will sometimes hear this approach referred to as "ecological niche
+modeling" [e.g., @feng_etal2019; @sillero_barbosa2021; @sillero_etal2021] or
+even more explicitly as "habitat suitability modeling" [e.g., @bowden_etal2021;
+@rowden_etal2017; @dekort_etal2020]. The basic idea here is that we can look at
+the distribution of Fremont sites as a reliable indicator of the sorts of
+available environmental conditions that would best promote maize farming in an
+arid landscape. To put that in more blatantly economic terms, we are assuming
+that their environmental preferences are shaped primarily by the constraints of
+maize farming and revealed in their residential choice behavior. This allows us
+to build a tentative model of the ecological niche for maize farming and to make
+some defeasible inferences about its limiting conditions.
+
+::: {#fig-overview fig-align="center"}
+![](../figures/overview.jpg){width="5in"}
+
+On the left is an overview map showing the project area with modern satellite
+imagery, so the green areas indicate plant growth, the beige areas, empty
+desert. The green areas are also the mountainous areas. Thick orange lines are
+interstate highways. Thin orange lines are major state highways. For
+visualization purposes, the map on the right shows the log-transformed density
+of feature-weighted archaeological sites. If no sites have been recorded in a
+watershed, log(1e-5) is shown.
+:::
 
@@
 
-Archaeologists have nominated a number of artifact types and artifact properties as candidate traits to distinguish the Fremont from their neighbors in space and time, including metates with secondary grinding surfaces, large semi-subterranean pithouses, various clay figurines, trapezoidal rock art, one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and elongated corner-notched arrow points [@madsen1998].
-None of these, however, nor any combination of them, has yet proven up to the task of bringing these diverse people under a single, all-encompassing definition [@madsen1998].
-Nevertheless, the Fremont do stand out, especially at their population apex roughly 1000 years BP.
+Archaeologists have nominated a number of artifact types and artifact properties
+as candidate traits to distinguish the Fremont from their neighbors in space and
+time, including metates with secondary grinding surfaces, large
+semi-subterranean pithouses, various clay figurines, trapezoidal rock art,
+one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and
+elongated corner-notched arrow points [@madsen_simms1998]. None of these,
+however, nor any combination of them, has yet proven up to the task of bringing
+these diverse people under a single, all-encompassing definition
+[@madsen_simms1998]. Nevertheless, the Fremont do stand out, especially at their
+population apex roughly 1000 years BP.
 
-Around this time, the Fremont likely reached their greatest geographic extent, inhabiting an area that encompassed most of the modern state of Utah [@janetski2011], with varying levels of support for occupations in Idaho [@dean1992prehistoric], Nevada [@hockett1998], Colorado [@baker1999], and Wyoming [@hakiel1987archery; @smith1992fremont].
-These modern political boundaries roughly overlap with the eastern Great Basin and Colorado Plateau physiographic provinces.
+Around this time, the Fremont likely reached their greatest geographic extent,
+inhabiting an area that encompassed most of the modern state of Utah
+[@janetski_etal2011], with varying levels of support for Fremont foragers in
+Idaho [@dean1992prehistoric] and Wyoming [@hakiel1987archery; @smith1992fremont]
+and Fremont farmers in Nevada [@hockett1998; @cole2012a] and Colorado
+[@baker1999]. These modern political boundaries roughly overlap with the eastern
+Great Basin and Colorado Plateau physiographic provinces.
 
-![
-Linear responses of average water-year precipitation (mm), average growing-season maize gdd (C), and cost-distance to streams (hours) to changes in elevation (m).
-See methods for how these values were calculated and results for more details.
-](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}
+::: {#fig-elevation fig-align="center"}
+![](../figures/elevation-everything.jpg){width="4in"}
 
-A basin and range topography dominates the landscape of the Great Basin, with large, dry valleys and endorheic basins punctuated by north-south trending mountain ranges.
-Changes in elevation in this setting are quite dramatic, on the order of several thousand meters over a short east-west transect.
-As elevation drives virtually every climatological process in the region, the local ecology is characterized by equally dramatic extremes [@billings1951; @grayson2011].
-As shown in @fig-elevation, net water-year precipitation in the eastern Great Basin tends to increase with elevation, while maize growing degree days (GDD) over the growing-season tends to decrease.
-An important trade-off thus exists between temperature and precipitation, with warmer and drier conditions at lower elevations and cooler and wetter conditions at higher elevations.
+Linear responses of the centered and scaled average water-year precipitation
+(mm), average growing-season maize gdd (C), and cost-distance to streams (hours)
+to changes in elevation (m). See methods for how these values were calculated
+and results for more details.
+:::
+
+A basin and range topography dominates the landscape of the Great Basin, with
+large, dry valleys and endorheic basins punctuated by north-south trending
+mountain ranges. Changes in elevation in this setting are quite dramatic, on the
+order of several thousand meters over a short east-west transect. As elevation
+drives virtually every climatological process in the region, the local ecology
+is characterized by equally dramatic extremes [@billings1951; @grayson2011]. As
+shown in @fig-elevation, net water-year precipitation in the eastern Great Basin
+tends to increase with elevation, while maize growing degree days (GDD) over the
+growing-season tends to decrease. An important trade-off thus exists between
+temperature and precipitation, with warmer and drier conditions at lower
+elevations and cooler and wetter conditions at higher elevations.
 
 These are the environments into which the Fremont first emerged in western Utah.
-Models for describing their origins, while diverse, tend to fall into the familiar categories of migration and diffusion.
-Where models of the former migration variety insist that Ancestral Puebloan farmers migrated northward from the US Southwest, bringing with them drought-tolerant maize varieties and maize farming [@madsen1975; @kidder1924], models of the latter diffusion variety suggest that the Fremont first emerged within these areas as local Archaic populations transitioned away from mobile foraging, adopting semi-sedentary agricultural practices from the southwest in a piecemeal fashion over several hundred years [@jennings1978; @winter1974].
-While archaeologists have marshaled an impressive array of arguments and counter-arguments for these two classes of models, a general consensus appears to be coalescing around the idea that the truth lies somewhere in between, that the interaction between a nascent Archaic population existing at extremely low population levels and a burgeoning community of migrant farmers together produced the complex we now call the Fremont [@patterson2015; @simms1986; @spangler1993; @spangler2000; @spangler2013].
+Models for describing their origins, while diverse, tend to fall into the
+familiar categories of migration and diffusion. Where models of the former
+migration variety insist that Ancestral Puebloan farmers migrated northward from
+the US Southwest, bringing with them drought-tolerant maize varieties and maize
+farming [@madsen_berry1975; @kidder1924], models of the latter diffusion variety
+suggest that the Fremont first emerged within these areas as local Archaic
+populations transitioned away from mobile foraging, adopting semi-sedentary
+agricultural practices from the Southwest in a piecemeal fashion over several
+hundred years [@jennings1978; @winter_wylie1974]. While archaeologists have
+marshaled an impressive array of arguments and counter-arguments for these two
+classes of models, a general consensus appears to be coalescing around the idea
+that the truth lies somewhere in between, that the interaction between a
+lingering Archaic population existing at extremely low population levels and a
+burgeoning community of migrant Southwest farmers together produced the complex
+we now call the Fremont [@patterson2015; @spangler1993; @spangler2000;
+@spangler2013; @searcy_talbot2015; @simms2008].
 
-Models that try to explain *why* Fremont maize farmers emerged in the far northern periphery of the US Southwest tend to agree - more or less - that at some point farming became a more economical alternative to foraging in the region.
-Where they differ, it is in whether things got better for farming starting roughly 2000 years BP [@benson2011; @coltrain2002; @matson1988] or worse for foraging [@barlow2002; @broughton2010; @cannon2000; @cannon2001].
-In support of the *farming-more-better* model, archaeologists draw on evidence for a suite of favorable climatic conditions that came together around 1600 years BP, including warmer temperatures and increased precipitation (both winter and summer), as well as the expansion of grasslands [@grayson2006; @hemphill1995; @rhode2000].
-The greatest expansion of the Fremont, in fact, occurred around 1000 years BP, around the time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures that would have increased the intensity of summer monsoons in the southwest, pushing them further north into the Great Basin and onto the Colorado Plateau [@grayson2006], a pattern reflected in numerous tree ring records [e.g., @graybill1990; @leavitt1994; @stine1990; @salzer2014], though some complications do exist for this suggested pattern [@hart2021].
+Models that try to explain *why* Fremont maize farmers emerged in the far
+northern periphery of the US Southwest tend to agree - more or less - that at
+some point farming became a more economical alternative to foraging in the
+region. Where they differ, it is in whether things got better for farming
+starting roughly 2000 years BP [@benson2011; @coltrain_leavitt2002;
+@matson_etal1988] or worse for foraging [@barlow2002; @broughton_etal2010;
+@cannon2000; @cannon2001]. In support of the *farming-got-better* model,
+archaeologists draw on evidence for a suite of favorable climatic conditions
+that came together around 1600 years BP, including warmer temperatures and
+increased precipitation (both winter and summer), as well as the expansion of
+grasslands [@grayson2006; @hemphill_wigand1995; @madsen2000]. The greatest
+expansion of the Fremont, in fact, occurred around 1000 years BP, around the
+time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures
+that would have increased the intensity of summer monsoons in the Southwest,
+pushing them further north into the Great Basin and onto the Colorado Plateau
+[@grayson2006], a pattern reflected in numerous tree ring records [e.g.,
+@graybill1990; @leavitt1994; @stine1990; @salzer_etal2014], though some
+complications do exist for this suggested pattern [@hart_etal2021].
 
-In response to these sorts of arguments, those who favor the *foraging-more-worse* model [@hart2021; @barlow2002] argue that conditions promoting maize farming would also tend to make foraging a more productive strategy.
-Plus, farming is an extremely costly endeavor, especially when compared to foraging, including lots of upfront investment of time and energy, so improving conditions favorable to maize agriculture are not by themselves sufficient to explain its origins.
-To fill that lacuna, archaeologists point to accumulating evidence suggesting that the Fremont transition to agriculture accompanied periods of sustained population growth [@barlow2008; @codding2021; @spangler2013] that likely resulted in the depressed availability of wild resources [@barlow2002; @janetski1997; @simms1986].
-So, demographic pressure reduced the efficiency of foraging strategies, thus making maize agriculture the optimal strategy.
-And that, in turn, led the Fremont to adopt farming as their primary mode of subsistence.
+In response to these sorts of arguments, those who favor the
+*foraging-got-worse* model [@hart_etal2021; @barlow2002] argue that conditions
+promoting maize farming would also tend to make foraging a more productive
+strategy. Plus, farming is an extremely costly endeavor, especially when
+compared to foraging, including lots of upfront investment of time and energy,
+so improving conditions favorable to maize agriculture are not by themselves
+sufficient to explain its origins. To fill that lacuna, archaeologists point to
+accumulating evidence suggesting that the Fremont transition to agriculture
+accompanied periods of sustained population growth [@barlow_etal2008;
+@codding_etal2022; @spangler2013] that likely resulted in the depressed
+availability of wild resources [@barlow2002; @janetski1997; @simms1986]. So,
+demographic pressure reduced the efficiency of foraging strategies, thus making
+maize agriculture a more profitable alternative. And that, in turn, led the
+Fremont to adopt farming as their primary mode of subsistence.
 
-Once they entered the farming niche, the Fremont quickly developed several adaptations for coping with the marginal and stochastic conditions in the eastern Great Basin and the Colorado Plateau, the most important almost certainly being irrigation (for more on this, see the discussion).
-On the Colorado Plateau, the eastern Fremont settled deep within the steep canyons and washes that flanked the tributaries of major rivers like the Green and Yampa [@yaworsky2021; see also, Yaworsky this issue], in places like Nine Mile Canyon [@spangler1993] and Range Creek Canyon [@hart2021].
-In the eastern Great Basin, the western Fremont chose to settle on open, alluvial fans and stream terraces along the lower slopes of most ranges [@janetski2000; @madsen1998; @simms2008], at places like Nephi Mounds [@sharrock1967], Pharo Village [@marwitt1968], Median Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock Village [@aikens1967].
+Once they entered the farming niche, the Fremont quickly developed several
+adaptations for coping with the marginal and stochastic conditions in the
+eastern Great Basin and the Colorado Plateau, the most important almost
+certainly being irrigation (for more on this, see the discussion). On the
+Colorado Plateau, the eastern Fremont settled deep within the steep canyons and
+washes that flanked the tributaries of major rivers like the Green and Yampa
+[@yaworsky2021], in places like Nine Mile Canyon [@spangler1993] and Range Creek
+Canyon [@hart_etal2021]. In the eastern Great Basin, the western Fremont chose
+to settle on open, alluvial fans and stream terraces along the lower slopes of
+most ranges [@janetski_talbot2000; @madsen_simms1998; @simms2008], at places
+like Nephi Mounds [@sharrock_marwitt1967], Pharo Village [@marwitt1968], Median
+Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock
+Village [@aikens1967].
 
-Given the extreme aridity of the west, they were still susceptible to drought, of course, even in those more suitable areas, with a single short growing season making agricultural returns highly variable.
-To cope with these uncertainties, Fremont farmers may have adopted additional risk-mitigation strategies like spreading farm plots among different micro-environments, a technique otherwise known as plot diversification [@spangler1993; @spangler2013; @patterson2015].
-The Fremont also relied heavily on crop storage both to provide a surplus during the winter season and to make-up for poor yields [@spangler2019].
-
-## The Ideal Free Distribution
-
-In population ecology, settlement decisions are represented as a choice between habitats, where each habitat $i$ has some net biological value or benefit, known as its suitability ($S_i$).
-Typically, though not always, this value exhibits negative density dependence, declining as a function of population size ($N_i$), perhaps owing to increased competition or resource depression.
-When $N_i$ is very small, the suitability of a habitat approaches its pristine or intrinsic quality ($Q_i$), or its gross value independent of all settlement costs [@greene2001].
-Together, these parameters suggest a very simple model of habitat suitability:
-
-$$
-S_i = Q_i - f(N_i)
-$$
-
-This is known as the Ideal Free Distribution (IFD) model [@fretwell1969].
-The IFD model assumes that individuals will choose habitats that maximize $S_i$, settling the most suitable habitat first then shifting to the next most suitable habitat when those become equal, and thereafter infilling each at an equal rate.
-The aggregate effect is a population distributed at equilibrium, meaning suitability is equal for all habitats.
-When conditions are ideal (individuals have perfect knowledge) and free (individuals face no additional costs to settlement), this distribution will also satisfy the "input matching" rule, with populations occurring at larger numbers in habitats with greater intrinsic value [@parker1978].
-
-Two circumstances provide important exceptions to this simple model.
-The first concerns economies of scale, or situations in which $S_i$ exhibits positive density dependence.
-In ecology, these are known as Allee effects, and they are assumed to occur at low density [@allee1949; @fretwell1969].
-The second concerns additional settlement costs, or costs associated with moving from one habitat to another.
-This is a violation of the free element of the IFD.
-When it occurs, the result is an ideal despotic distribution (IDD) [@fretwell1969].
-This has the consequence of reducing the total population in the more suitable habitat at equilibrium.
-
-Many social and ecological variables may be relevant to defining a general measure of $S_i$, but here we are focusing on ecological determinants of intrinsic habitat quality, $Q_i$, that are specific to maize farmers.
-Our reasoning here is fairly straightforward.
-If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters.
-That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium.
-So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.
+Given the extreme aridity of the west, they were still susceptible to drought,
+of course, even in those more suitable areas, with a single short growing season
+making agricultural returns highly variable. To cope with these uncertainties,
+Fremont farmers may have adopted additional risk-mitigation strategies like
+spreading farm plots among different micro-environments, a technique otherwise
+known as plot diversification [@spangler1993; @spangler2013; @patterson2015].
+The Fremont also relied heavily on crop storage both to provide a surplus during
+the winter season and to make-up for poor yields [@spangler_etal2019].
 
@@
 
-Because we are focusing on the western Fremont, we constrain the current project area to the eastern Great Basin in western Utah, as shown in @fig-overview, where the Great Basin is defined in the hydrologic sense using watershed boundaries [@grayson2011].
-On the far eastern edge of this region, a series of north-south ranges, including the Wasatch, Pahvant, and Tushar Mountains, together stand as the primary physiographic boundaries separating the Great Basin from the Colorado Plateau.
-On its far western edge, a series of north-south ranges straddle the Nevada state line or lie just inside it, including the Snake Range where Great Basin National Park is located.
-Its northern periphery is the Great Salt Lake and the West Desert, its southern boundary the Pine Valley Mountains that separate the Great Basin from the Virgin River watershed, just north of St.
-George, UT.
-This is a region approximately 107,000 km$^2$ in area, encompassing virtually all of the Bonneville Basin in western Utah.
+Because we are focusing on the western Fremont, we constrain the current project
+area to the eastern Great Basin in western Utah, as shown in @fig-overview,
+where the Great Basin is defined in the hydrologic sense using watershed
+boundaries [@grayson2011]. On the far eastern edge of this region, a series of
+north-south ranges, including the Wasatch Mountains in the north and the Sevier
+Plateau in the south, together stand as the primary physiographic boundaries
+separating the majority of the Great Basin from the Colorado Plateau. On its far
+western edge, a series of north-south ranges straddle the Nevada state line or
+lie just inside it, including the Snake Range where Great Basin National Park is
+located. Its northern periphery includes the Great Salt Lake (GSL) and the West
+Desert, its southern boundary the Bull and Pine Valley Mountains that separate
+the Great Basin from the Virgin River watershed just north of St. George, UT.
+This is a region approximately 107,000 km$^2$ in area, encompassing virtually
+all of the Bonneville Basin in western Utah.
 
-These boundaries also serve as geographic borders between the western Fremont and other contemporaneous populations.
-Around the Great Salt Lake and the Utah-Nevada border, Fremont farming gives way to Archaic populations engaged in intensive foraging modes of subsistence.
-The Wasatch Range and Tushar Mountains separate the western from the eastern Fremont on the Colorado Plateau, and the Pine Valley Mountains serve as a border between the Fremont and the Virgin Ancestral Puebloan to the south [@simms2008].
+These boundaries also serve as geographic borders between the western Fremont
+and other contemporaneous populations. Around the GSL and the Utah-Nevada
+border, Fremont farming gives way to Archaic populations engaged in intensive
+foraging modes of subsistence. The Wasatch Range and Tushar Mountains separate
+the western from the eastern Fremont on the Colorado Plateau, and the Pine
+Valley Mountains serve as a border between the Fremont and the Virgin Ancestral
+Puebloan to the south [@simms2008].
 
@@
 
-The unit of analysis for this research is the HUC10 watershed (n = 183), as defined by the Watershed Boundary Dataset (WBD) developed by the US Geological Survey and the Natural Resources Conservation Service within the US Department of Agriculture [@usgs2013].
-These provide a spatially-explicit proxy for specifically human habitats in the proposed study area.
-There are two reasons for this.
-First, the topography of the region is such that the cost of travel between watersheds is often quite significant.
-As a consequence, individuals are expected to spend more time traveling within watersheds than between them, all else being equal.
-Second, watersheds are by definition water sinks, funneling all available run-off into their respective stream networks.
-Given the extreme aridity of the proposed project area, this also makes them critical resource sinks, especially for maize farmers, as watersheds determine how much water might accumulate at a location, as well as its potential for irrigation.
+The unit of analysis for this research is the HUC10 watershed (n = 183), as
+defined by the Watershed Boundary Dataset (WBD) developed by the US Geological
+Survey and the Natural Resources Conservation Service within the US Department
+of Agriculture [@usgs_usdanrcs2013]. These provide a spatially-explicit proxy
+for specifically human habitats in the proposed study area. There are two
+reasons for this. First, the topography of the region is such that the cost of
+travel between watersheds is often quite significant. As a consequence,
+individuals are expected to spend more time traveling within watersheds than
+between them, all else being equal. Second, watersheds are by definition water
+sinks, funneling all available run-off into their respective stream networks.
+Given the extreme aridity of the proposed project area, this also makes them
+critical resource sinks, especially for maize farmers, as watersheds determine
+how much water might accumulate at a location, as well as its potential for
+irrigation.
 
-Aggregating to the level of the watershed also reduces the computational burdens of this modeling exercise.
-In particular, it makes paleoclimate reconstructions more tractable, as we are only estimating the means within each watershed through time (see below for details).
+While aggregating to the level of the watershed does reduce the spatial
+resolution of this analysis, depriving our model of a substantial amount of
+environmental variation, several arguments can be made for the simplification.
+First, it reduces the computational burdens of this modeling exercise. In
+particular, it makes paleoclimate reconstructions more tractable, as we are only
+estimating the means within each watershed through time (see below for details).
+Second, if there are good theoretical justifications for using certain polygons
+as spatially explicit habitats (like the ones outlined above), they can be
+relied upon to identify real, presumably causal, relationships in the data.
+Third, the aggregated site data can be shared without giving away site
+locations, making it easier for others to reproduce the analysis. Finally,
+depending on the question being asked, disaggregated site locations may offer
+only false precision. This is likely the case for the current analysis as
+Fremont were opportunistic hunters [@morgan_etal2012], so many of the sites in
+our database are likely temporary logistic hunting sites in proximity of their
+actual residential locations, but crucially the available data make it hard to
+differentiate these kinds of sites.
 
@@
 
-To identify Fremont sites in the project area, we rely on site records and cultural resource reports hosted by the cultural resource information systems in Nevada (NVCRIS) and Utah (Sego) with permission from the respective State Historic Preservation Offices.
-The Sego system also includes an attribute table with affiliation data that we use to filter Fremont sites where possible.
-For a previous project, we also performed a stratified random sample of all archaeological sites by county in western Utah and recorded affiliation and other attribute data from their site forms.
-For this analysis, we compare the results of that previous effort with the sites identified by filtering the Sego system as a form of quality control.
-The resulting dataset contains 2,248 individual Fremont sites.
+To identify Fremont sites in the project area, we rely on site records and
+cultural resource reports hosted by the cultural resource information systems in
+Nevada (NVCRIS) and Utah (Sego) with permission from the respective State
+Historic Preservation Offices. Data collection is part of a larger NSF-funded
+project (BCS-1921072) that is examining demographic reconstruction in the
+Bonneville Basin and surrounding areas [see @codding_etal2022;
+@codding_etal2023; @contreras_codding2023]. Unfortunately, our dataset does not
+represent all recorded Fremont sites in the study area, as we did not review all
+site forms in a systematic fashion. Some site forms were chosen at random.
+Others were selected opportunistically. Still, the resulting dataset contains
+2,248 individual Fremont sites, a considerable number representing one of the
+largest samples used for this sort of analysis to date.
 
-These site data include everything from small ceramic scatters to massive Fremont villages like Five Finger Ridge [@janetski1998; @janetski2000].
-For this analysis, we assume that these reflect different levels of population size and settlement intensity.
-All else being equal, scatters should follow from shorter stays by fewer people, villages from longer stays by more people.
-This is important because population size correlates with a habitat's total suitability according to the IFD model, and our attempts to reconstruct the ecological and geographic borders of the Fremont rely heavily on the input-matching rule, or the idea that habitats with greater intrinsic quality will have larger populations at equilibrium.
-We, therefore, weight the count of sites in each watershed by the total number of residential features that they contain (+1 for sites with no features).
-So, a site with no residential features counts as one site, a site consisting of, say, four pithouses counts as four unique sites, and a site containing a roomblock with seven rooms counts as seven sites.
-This brings the total weighted site count to 2,447, with minimum and maximum values of 0 and 119, respectively.
+These site data include everything from small ceramic scatters to massive
+Fremont villages like Five Finger Ridge [@janetski1998; @janetski_talbot2000].
+For this analysis, we assume that these reflect different levels of population
+size and settlement intensity. All else being equal, scatters should follow from
+shorter stays by fewer people, villages from longer stays by more people. This
+is important because population size correlates with a habitat's suitability
+according to the IFD model, and our attempts to reconstruct the ecological and
+geographic borders of the Fremont rely heavily on the idea that more suitable
+habitats will have larger populations at equilibrium.
 
-@fig-overview shows the relative density of Fremont sites across watersheds.
-The *relative* density refers to the density of feature-weighted sites ($D_i = N_i/area_i$) divided by the maximum density ($max\;D$) across all watersheds, thus scaling the density estimates to the unit interval [0,1].
-Using the relative density is just a way of avoiding any implication that we have an independent estimate of the absolute number of Fremont sites in the area, and not just an estimate based on a thinned sample.
+To get the architectural feature data, we relied primarily on summaries of well
+known and named Fremont villages in [@coltrain_leavitt2002] and cross-referenced
+those with original site forms and reports, as well as the work of
+[@mooney2014], which attempts to reconstruct poorly reported excavation efforts
+along the Wasatch Front. Storage features were not included in this count, only
+residential features described as pitstructures, pueblos, roomblocks, or
+wickiups. All told, this included 16 Fremont villages and 215 residential
+features. In several cases, the estimates of the number of features on a site
+vary wildly, so we erred on the side of taking the most conservative estimates.
+We then weight the count of sites in each watershed by those estimates. So, a
+site with no residential features counts as one site, a site consisting of, say,
+four pithouses counts as four unique sites, and a site containing a roomblock
+with seven rooms counts as seven sites. This brings the total weighted site
+count to 2,447, with minimum and maximum values of 0 and 119, respectively.
+
+@fig-overview shows the log density of Fremont sites across watersheds. The log
+is used mainly for visualization purposes as a number of high density watersheds
+swamped the variance shown on the untransformed scale. Note that if a watershed
+has zero sites in it, we used $log(1e-5)$ as $log(0)$ is undefined. It should be
+emphasized that we do not have an independent estimate of the absolute number of
+Fremont sites in the area, just an estimate based on a thinned sample, so the
+map is more appropriately interpreted as showing relative differences in the
+density of recorded sites across watersheds.
 
@@
 
-For this analysis, we use both topographic and climatological covariates.
-The topographic covariate is cost-distance to perennial streams (measured in hours).
-This provides a coarse grained estimate of water availability, as well as the potential costs of irrigation.
-To calculate this, we first use the R package FedData [@bocinsky2020] to download perennial stream features from the US National Hydrography Dataset [@usgs2022] and a digital elevation model (DEM) from the US National Elevation Dataset [@usgs2022a].
-We then apply Campbell's hiking function [@campbell2019] to slope estimates derived from the DEM using the R package hiker [@vernon2021a].
-This allows us to generate estimates of travel time between grid cells in the DEM and then calculate the accumulated cost of travel from each perennial stream to any grid cell within each watershed.
-We then aggregate these values to the watershed level by taking their mean.
+For this analysis, we use both topographic and climatological covariates. The
+topographic covariate is cost-distance or travel time to perennial streams
+(measured in hours). This provides a coarse grained estimate of water
+availability, as well as the potential costs of irrigation. To calculate this,
+we first use the R package FedData [@bocinsky2020] to download perennial stream
+features from the US National Hydrography Dataset [@usgs2022] and a digital
+elevation model (DEM) from the US National Elevation Dataset [@usgs2022a]. We
+then apply Campbell's hiking function [@campbell_etal2019] to slope estimates
+derived from the DEM. This allows us to estimate travel time between grid cells
+in the DEM and then calculate the accumulated cost of travel from each perennial
+stream to any grid cell within each watershed. We then aggregate these values to
+the watershed level by taking their mean.
 
-Climate covariates include net water-year precipitation (precipitation, in millimeters, mm) from October to September and maize growing degree days (GDD, in Celsius, C) over the growing season from May to September, as these have a large effect on maize productivity, thus providing important constraints on the patterns of settlement for those who relied on maize for subsistence.
-We caution that GDD is calculated in such a way as to be insensitive to extreme high and low temperatures (extreme heat and frost, in effect).
-It is also conceptually understood to be a measure of accumulated temperature between the last and first frost-free days, but in our analysis is calculated using hard start and end dates of May 1 and September 30, under the assumption that these dates will roughly coincide with those frost-free days.
+Climate covariates include net water-year precipitation (PPT, in millimeters,
+mm) from October to September and maize growing degree days (GDD, in Celsius, C)
+over the growing season from May to September, as these have a large effect on
+maize productivity, thus providing important constraints on the patterns of
+settlement for those who relied on maize for subsistence. We caution that GDD is
+calculated in such a way as to be insensitive to extreme high and low
+temperatures (extreme heat and frost, in effect). It is also conceptually
+understood to be a measure of accumulated temperature between the last and first
+frost-free days, but in our analysis is calculated using hard start and end
+dates of May 1 and September 30, under the assumption that these dates will
+roughly coincide with those frost-free days.
 
-![
-Distribution of covariates across watersheds. These include maize growing degree days (GDD), annual precipitation (precipitation), and cost-distance to streams (Streams).
-](../figures/covariates.png){#fig-covariates fig-align="center" width="6in"}
+::: {#fig-covariates fig-align="center"}
+![](../figures/covariates.jpg){width="5in"}
 
-We hindcast these climate covariates for each watershed and year over the Fremont sequence from 1550 to 550 years BP (400 to 1400 CE) using the Correlation Adjusted corRelation (CAR) method implemented in the R package paleocar [@bocinsky2019].
-This method, in effect, regresses modern climate data against the width of younger tree rings, then uses older rings to predict past climate trends [for further details, see @bocinsky2014; @bocinsky2016].
-In this case, we use interpolated precipitation and temperature estimates from the \~800 m resolution PRISM dataset [@prismclimategroup2014] and all tree-ring samples in North America, which were obtained from the International Tree Ring Database using FedData.
-For each watershed, we took the median values of precipitation and GDD over the Fremont sequence.
+Distribution of covariates across watersheds. These include maize growing degree
+days (GDD), annual precipitation (precipitation), and cost-distance to streams
+(Streams). Lighter values represent larger values, darker colors smaller values.
+:::
 
-@fig-covariates shows the geographic distributions of these covariates over watersheds.
-As these are all in one way or another driven by elevation, it is reasonable to expect that they will correlate.
-As part of the exploratory phase of this analysis, we, therefore, perform a series of pair-wise tests of correlation measured using Pearson's R.
-We also regress the scaled and centered values (or z-scores) of maize GDD, precipitation, and cost-distance to streams on elevation using ordinary least squares (OLS).
-No doubt, this violates many assumptions of OLS, particularly the identity and independence of the errors, but explanation was not our main goal with these simple linear models.
-We are only trying to get a rough sense of their general response to changes in elevation, which OLS offers (see @fig-elevation).
+We hindcast these climate covariates for each watershed and year over the
+Fremont sequence from 1550 to 550 years BP (400 to 1400 CE) using the
+Correlation Adjusted corRelation (CAR) method implemented in the R package
+paleocar [@bocinsky2019]. This method, in effect, regresses modern climate data
+against the width of younger tree rings, then uses older rings to predict past
+climate trends [for further details, see @bocinsky_kohler2014;
+@bocinsky_etal2016]. In this case, we use interpolated precipitation and
+temperature estimates from the \~800 m resolution PRISM dataset
+[@prismclimategroup2019] and all tree-ring samples in North America, which were
+obtained from the International Tree Ring Database using FedData. For each
+watershed, we took the median values of precipitation and GDD over the Fremont
+sequence.
 
-Given that our primary covariates all correlate with elevation (reported below), readers may wonder why we do not simply use elevation in our main analysis.
-The reason for this is two-fold.
-First, using elevation alone would obscure the trade-off between water availability and temperature.
-Second, elevation is not explanatory of settlement behavior.
-A maize farmer does not choose a location for farming *because* it is at, say, 1200 m above sea level.
-They choose that location *because* it has the right combination of water availability and temperature, that is to say, because conditions there are optimal for maize farming.
-Were those conditions found at some other elevation, the farmer would be expected to farm there.
+@fig-covariates shows the geographic distributions of these covariates over
+watersheds. Orographic effects (meaning the effects of mountains and elevation)
+are rampant in these data, so it is reasonable to expect that the covariates
+will exhibit some multicollinearity. As part of the exploratory phase of this
+analysis, we, therefore, perform a series of pair-wise tests of correlation
+measured using Pearson's R. We also regress the scaled and centered values (or
+z-scores) of maize GDD, precipitation, and cost-distance to streams on elevation
+using ordinary least squares (OLS). No doubt, this violates many assumptions of
+OLS, particularly the identity and independence of the errors, but explanation
+was not our main goal with these simple linear models. We are only trying to get
+a rough sense of their general response to changes in elevation, which OLS
+offers (see @fig-elevation).
+
+Given that our primary covariates all correlate with elevation (reported below),
+readers may wonder why we do not simply use elevation in our main analysis. The
+reason for this is two-fold. First, using elevation alone would obscure the
+trade-off between water availability and temperature. Second, elevation is not
+explanatory of settlement behavior. A maize farmer does not choose a location
+for farming *because* it is at, say, 1200 m above sea level. They choose that
+location *because* it has the right combination of water availability and
+temperature, that is to say, because conditions there are optimal for maize
+farming. Were those conditions found at some other elevation, the farmer would
+be expected to farm there instead.
 
@@
 
-The response variable is a count $N$ of sites per watershed $i$, so it should arise as the result of a Poisson process.
-Given the trade-offs outlined above, we also expect it to exhibit a non-linear response to precipitation, GDD, and cost-distance to streams, so we fit a generalized additive model (GAM) with a log-link to the data using the R package mgcv [@wood2004].
-Although we expect the outcome to be Poisson distributed, we use a negative binomial distribution here, as this includes a parameter, $r$, that can account for potential dispersion without giving up the inferential benefits of maximum likelihood.
+For this analysis, the outcome of interest is a count $N$ of archaeological
+sites per watershed $i$. Here $N$ is assumed to follow a negative binomial
+distribution:$$
+N_i \sim NB\,(\lambda_i, r)
+$$with $r$ being a dispersion parameter and $\lambda_i$ the expected number of
+sites per watershed, $E\,(N_i) = \lambda_i$. To estimate $\lambda_{i}$, we model
+it as a log response to the linear predictor:$$
+log(\lambda_i) = \alpha + \sum_{k=1}^{K}{s(x_{ik})} + \epsilon_i
+$$where $\alpha$ is the intercept, $\epsilon_i$ is the error term, and $s$ is a
+smooth function applied to the $K$ covariates - growing degree days,
+precipitation, and cost-distance to streams. The smooth is intended to account
+for potential non-linear responses to the covariates. In our model, $\epsilon_i$
+incorporates an exponential covariance model to account for spatial
+autocorrelation in the untransformed residuals. The complete model is fit within
+a generalized additive mixed model (GAMM) framework using the R package mgcv
+[@wood2004].
 
-$$
-\begin{aligned}
-N_i \sim& \;NB\,(\lambda, r)\\
-E\,(N_i) =& \;\lambda_i\\
-log\, (\lambda_i) =& \;log\, (area_i) + s(precipitation_i) + streams_i + gdd_i + protected_i
-\end{aligned}
-$$
+Importantly, we add to the model specification a constant offset for the log of
+the area of each watershed, in effect making this a model of population
+*density*. This is meant to address the idea that larger watersheds will have
+more sites just as a matter of chance. While our model as currently specified
+includes the intercept, we caution that such an estimate cannot be interpreted
+in terms of the abundance of Fremont sites, as that would assume that the
+absolute site counts can be inferred from this sample, which is implausible. At
+best, the intercept tells us about the abundance of *recorded* sites when
+$\sum_{k=1}^{K}{f(x_{ik})}=0$. On that note, we also include a parametric term
+for the proportion of each watershed classified as federally protected land.
+This is to account for anthropogenic impacts to archaeological resources that
+can bias our sample. Notably, for these data, those are almost entirely the
+result of Euro-American colonization of the area, in particular, the impacts of
+modern farming techniques and the development of the Wasatch Front as the urban
+core of the modern state of Utah. This is a coarse metric, but our expectation
+is that the greater the share of a watershed with federal protections, the
+smaller the effect of these anthropogenic impacts, the more sites we should
+observe.
 
-Importantly, we add to the model specification a constant offset for the log of the area of each watershed, in effect making this a model of population *density*.
-This is meant to address the idea that larger watersheds will have more sites just as a matter of chance.
-While our model as currently specified includes the intercept, we caution that such an estimate is not strictly meaningful for this kind of data, as it assumes that the absolute site counts can be inferred from the sample, which requires that the sample be unbiased - an implausible condition to meet.
-On that note, we also include a parametric term for the proportion of each watershed classified as federally protected land.
-This is meant to address potential sampling biases owing to taphonomic processes operating on the archaeological record.
-Notably, for these data, those are almost entirely the result of Euro-American colonization of the area, in particular, the impacts of modern farming techniques and the development of the Wasatch Front as the urban core of the modern state of Utah.
-This is a coarse metric, but our expectation is that the greater the share of a watershed with federal protections, the smaller the effect of these anthropogenic impacts, the more sites we should observe.
+Model evaluation includes checks for concurvity or non-linear correlation in the
+smooth terms of the GAMM, as well as a Variance Inflation Factor test on the
+parametric or linear terms. We also test for spatial autocorrelation in the
+untransformed residuals using Monte Carlo simulations of Moran's I. After
+fitting the full model, we found that several smooth terms had effective degrees
+of freedom (EDF) equal to one, suggesting no non-linear response in the data.
+We, therefore, remove the smooth terms for those covariates in the final model,
+leaving only precipitation with a potential non-linear effect.
 
-Model evaluation includes checks for concurvity or non-linear correlation in the smooth terms of the GAM, as well as a Variance Inflation Factor test on the parametric or linear terms.
-We also test for spatial autocorrelation in the untransformed residuals using Monte Carlo simulations of Moran's I and incorporate an exponential covariance matrix to remove that residual autocorrelation.
-Several smooth terms had effective degrees of freedom (EDF) equal to one, suggesting no non-linear response in the data.
-We, therefore, remove the smooth terms for those covariates in the final model, as expressed in the formula above.
-
-All analyses are conducted in the R programming language and environment [@rcoreteam2022]. For details of this analysis, please see the Supplement.
+All analyses are conducted in the R programming language and environment
+[@rcoreteam2022]. For details of this analysis, please see the Supplement.
 
@@
 
-As shown in @tbl-results, all linear (or parametric) coefficients are significant in the final model.
-The intercept `r q[["Intercept"]]` is close to zero when transformed back onto the response scale, likely reflecting the large number of watersheds with zero Fremont sites, though as mentioned above, the intercept estimate is not strictly meaningful for these data.
-The proportion of watersheds falling under federal protections `r q[["Protected"]]` has a positive effect on site counts, thus confirming our expectation that more sites occur in watersheds with a larger proportion of federal land.
-Site counts increase in watersheds with larger GDD values, though the effect is small `r q[["Maize GDD"]]`, and site counts decrease with distance from perennial streams, though again the effect is small `r q[["CD to Streams"]]`.
-The only smooth term in the final model is average precipitation over the water-year, which is significant `r q[["s(Precipitation)"]]`.
-The effective degrees of freedom (EDF) for precipitation suggests a strong non-linear effect of that covariate on site counts, which increases up to about 450 mm and decreases thereafter.
+As shown in @tbl-results, all linear (or parametric) coefficients are
+significant in the final model. The intercept `r q[["Intercept"]]` is close to
+zero when transformed back onto the response scale, likely reflecting the large
+number of watersheds with zero Fremont sites, though as mentioned above, the
+intercept estimate is only a measure of the recorded site abundance. The
+proportion of watersheds falling under federal protections `r q[["Protected"]]`
+has a positive effect on site counts, thus confirming our expectation that more
+sites occur in watersheds with a larger proportion of federal land. Site counts
+increase in watersheds with larger GDD values, though the effect is small
+`r q[["Maize GDD"]]`, and site counts decrease with distance from perennial
+streams, though again the effect is small `r q[["CD to Streams"]]`. The only
+smooth term in the final model is average precipitation over the water-year,
+which is significant `r q[["s(Precipitation)"]]`. The effective degrees of
+freedom (EDF) for precipitation suggests a strong non-linear effect of that
+covariate on site counts, which increases up to about 450 mm and decreases
+thereafter.
 
-A concurvity test on the final model was not conducted as there was only one smooth term.
-As expected, a VIF test for linear correlation in the parametric terms of the final model shows some evidence of correlation, particularly for GDD, which was close to 6, but this value is within acceptable limits at a moderate threshold.
-The Moran's I test suggests that there is no spatial autocorrelation in the untransformed residuals of the final model (I = 0.233, rank = 381, p = 0.476).
+A concurvity test on the final model was not conducted as there was only one
+smooth term. As expected, a VIF test for linear correlation in the parametric
+terms of the final model shows some evidence of correlation, particularly for
+GDD, which was close to 6, but this value is within acceptable limits at a
+moderate threshold. The Moran's I test suggests that there is no spatial
+autocorrelation in the untransformed residuals of the final model (I = 0.233,
+rank = 381, p = 0.476).
 
-![
-Marginal response plots. Each plot shows the response of site counts to changes in a target covariate while holding all other variables at their mean.
-](../figures/marginal_response.png){#fig-responses fig-align="center" width="5in"}
+::: {#fig-responses fig-align="center"}
+![](../figures/marginal_response.jpg){width="4.5in"}
+
+Partial dependence plots. Each plot shows the response of site counts to changes
+in a target covariate while holding all other variables at their mean.
+:::
 
@@
 
-Our results indicate that Fremont maize farmers generally preferred warmer watersheds with longer growing seasons, moderate levels of precipitation (in the range of 400-600 mm), and greater access to perennial streams, as shown by the marginal response plots in Fig.
-3.
-While site counts exhibit a positive linear response to maize GDD, for maize has an upper threshold temperature at which point it no longer grows.
-Plus, increased temperatures trade-off with water availability, with warmer conditions being better for plant growth but at the same time increasing water demand on the plant to keep up with growth [@ramankutty2002].
-The linear relationship between site counts and maize GDD is, thus, relative to the observed distribution of that variable within the study area and should not be used to generalize to a larger geographic extent.
+With this analysis, we seek to reconstruct the Fremont maize farming niche by
+estimating relative differences in suitability across watersheds and
+environmental gradients. Our approach involves the use of a Negative Binomial
+GAMM model of the feature weighted counts of archaeological sites, under the
+assumption that more suitable watersheds should have more sites at equilibrium.
+Results from that model indicate that Fremont maize farmers generally preferred
+warmer watersheds with longer growing seasons, moderate levels of precipitation
+(in the range of 400-600 mm), and greater access to perennial streams, as shown
+by the marginal response plots in @fig-responses.
 
-Another curious result of our analysis is that site counts do not increase linearly as a function of precipitation, but instead decline after 500 mm.
-We believe this is likely an artifact of trade-offs related to elevation, specifically the trade-off between precipitation and temperature.
-We discuss this more below, along with other important implications our results have for the regional prehistory of the eastern Great Basin, particularly for the spread and eventual collapse of maize agriculture.
+## The Data Quality Problem in Fremont Research
 
-## A Goldilocks Zone?
+Fremont research focused on individual, well-dated, and excavated sites can and
+often does describe the processes acting on those sites in great detail, but any
+attempt to generalize to a whole region runs into a familiar problem, namely
+that the Fremont sequence suffers from poor chronological resolution due to a
+paucity of direct dates and a more or less undefined ceramic chronological
+record. The consequence for this analysis is that we have to use (i) total site
+counts in each watershed, not the contemporaneous site counts, and (ii) the
+median values of the climate variables over the entire Fremont sequence rather
+than over the years that individual Fremont sites were occupied. While there are
+radiocarbon dates for Fremont sites in the study area, they represent only a
+small fraction of known Fremont sites, so we can use them to get better
+chronological control, but only at the expense of losing spatial resolution.
 
-Provided that the trade-off between precipitation and temperature is driven largely by elevation in this region, it would seem likely that the western Fremont chose site locations at elevations that maximized agricultural suitability, as evidenced by the estimated distribution of sites across watersheds (see Fig.
-4), though we caution that watersheds are coarse grained units of analysis.
-Given the topography of the study area, this trade-off would also suggest that areas suitable for maize agriculture take the form of a thin band around the lower slopes of mountain ranges (or lower canyons in the Uintas).
-Such a band would be akin to a Goldilocks zone for maize agriculture (Yaworsky et al, this issue), a place where the optima for GDD and precipitation overlap, though we should not make too much of this idea, for it is an area of overlap for continuous ecological gradients, not a region demarcated by anything like a perfect line.
-It also must be emphasized that what matters most for maize agriculture is not necessarily where precipitation falls, but where it accumulates once it enters the streamflow network.
-Of course, that will largely be a function of elevation in this region, or changes in elevation, so it will still be in proximity to the Goldilocks zone, though perhaps not perfectly coincident with it.
-At any rate, given these considerations, it would perhaps be more accurate to characterize the trade-off confronting the Fremont as one between what we might call water availability (including general precipitation across the streamflow network, water runoff, proximity to perennial surficial sources, and irrigation potential, among other things) and temperature (including, but not limited to, the length of the growing season and the number of frost-free days).
+Fortunately, there are good reasons to believe that we have wrestled a
+meaningful signal from the noise in these temporally unresolved spatial data.
+For one thing, it is unlikely that inter-watershed variability would have
+changed substantially over the Fremont sequence. Over the last two thousand
+years, wetter watersheds would have tended to be wetter than dryer watersheds,
+and warmer watersheds would have tended to be warmer than colder watersheds. The
+central tendency of the climate in each watershed also tells us which of them
+would, in general, have been better for maize farming, even if we cannot yet
+define diachronic variation in watershed suitability. So, the inference is
+simply that those typically better watersheds would have attracted maize farmers
+more often and, over time, would have accumulated a larger archaeological maize
+farming assemblage, too.
 
-## Irrigation and the Maize Farming Niche
+## The Relative Contribution of Foraging to Fremont Diets
 
-Based on previous research [@benson2011; @adams2006; @bellorado2010], Bocinsky and Kohler [@bocinsky2014] place minimum thresholds for precipitation and maize GDD at 300 mm and 1000°C GDD, respectively.
-Importantly, these are thresholds for dry-farming, which is typically defined as farming free of irrigation [@benson2011; @cordell2012; @varien1999], so they can be loosely interpreted as minimum temperature and precipitation levels required to grow maize through nothing more than planting and harvesting.
-The right panel in @fig-thresholds shows the watersheds whose median values for maize GDD and precipitation over the Fremont sequence are above those values.
-This is similar to Bocinsky's concept of "refugia," but rather than calculate the proportion of years in niche, we are using median values, so refugia are defined here as watersheds that spend at least half the Fremont sequence in the rain-fed maize farming niche ($\geq$ 500 years, in this case).
+Many Fremont scholars are convinced that foraging had to contribute a
+substantial portion to the diet of Western Fremont populations
+[@madsen_simms1998; @simms1999]. If true, this would pose a substantial obstacle
+to any inferences about maize suitability based on the distribution of Fremont
+archaeological materials, as those inferences would conflate settlement patterns
+owing to constraints on foraging and settlement patterns owing to constraints on
+farming. This is problematic because it strikes at the heart of the
+interpretation we propose for our model. In fact, if true, it would basically
+nullify the whole paper, so it is worth confronting the question head-on: to
+what extent did foraging contribute to the diet of Fremont maize farming
+populations in the eastern Great Basin?
 
-![
-The map on the left shows the relative density of feature-weighted site counts as estimated by the model, with lighter colors representing larger relative densities, darker colors representing smaller relative densities.
-The map on the right shows which watersheds are in the rain-fed maize farming niche for at least half of the Fremont sequence.
-The light green color indicates those that are in the niche.
-](../figures/threshold_map.png){#fig-thresholds fig-align="center" width="6in"}
+Unfortunately, the kinds of observations that might help us arrive at an
+unequivocal answer to this question are frustratingly few and far between. The
+most compelling evidence currently available is provided by stable isotope
+analysis of a small number of Fremont burials [@coltrain1993;
+@coltrain_leavitt2002]. The results of those analyses suggest that the
+contribution of maize to Fremont diets was on the order of 75 to 80%, a fraction
+comparable to what one would find in the San Juan Basin, where there is no doubt
+that maize farming was the primary mode of subsistence during the time period in
+question. The one exception to this pattern involves individuals from the GSL
+Wetlands, who were at least part time foragers. But those exceptions are limited
+to the latter half of the Fremont sequence, after 1100 years BP (850 CE), and
+they don't show signs of full reliance on foraging until nearly 800 years BP
+(1150 CE). This is consistent with the suggestion that substantial foraging (and
+not just opportunistic foraging) would have been relegated mostly to the margins
+of areas where maize farming occurs [@allison2008]. In the study area, that
+would include areas surrounding and to the north of the GSL, as well as the West
+Desert and northern Nevada.
 
-It appears that these models tend to agree on the best watersheds for maize farming, though there are some notable exceptions.
-The western slopes of the Stansbury Range in the north central part of the study area and parts of the Parowan Valley and surrounding areas in the southeast part of the study area are estimated by our model to have non-negligible relative densities, but they fall outside of the niche.
-Other areas, like much of the Salt Lake valley appear to be in the niche, but are estimated by our model to have relatively low densities.
-These discrepancies are at least partially explained by limitations of our methodology.
-For one thing, including the proportion of federal land as a covariate is a very crude way of accounting for sampling bias, so our model might not be picking up on the fact that the Salt Lake valley was actually a highly suitable area for maize agriculture, as evidence to that effect has largely been destroyed.
-The discrepancy may also be explained by the fact that we have aggregated to the watershed level, thus obscuring important environmental variation across both the landscape in general and site locations in particular.
-Were we to build a model of the disaggregated data, we would likely find that many agricultural sites do, in fact, fall in the rain-fed maize agricultural niche, even if the watersheds that contain them do not.
+There are two additional things to note here. First, our results have important
+parallels with the results reported in [@yaworsky_etal2023], where the Fremont
+being studied are unambiguous maize farmers. This suggests that we are also
+modeling maize farming populations. The other point to mention is that our model
+incorporates an exponential covariance matrix, so if there were any systematic
+deviations around the north and northwest periphery of the study area, those
+would at least partially be accounted for in the model. In other words, the
+primary trend driven by well understood maize farming sites in the core of the
+study area (following the I-15 corridor south from Salt Lake) should not be
+noticeably affected by those deviations at the periphery.
 
-But setting aside those methodological issues, some much more interesting explanations may be offered, at least from the perspective of theory.
-In particular, the spatial covariance structure in our model may be picking up on the structure of the streamflow network, specifically the accumulation of water run-off as a function of slope.
-This would suggest that if there are more people in one watershed, there should be more more people in the downstream watershed, too, at least relative to other watersheds with similar precipitation and GDD values.
-In the language of the IFD model, the suitability of the watershed habitat should propagate through the streamflow network.
-Exploring that intuition in a more systematic way, however, would require much more careful consideration of streamflow.
-Unfortunately, that is well beyond the scope of the current paper.
+## Trade-off Between Precipitation and Temperature
 
-Still, the distribution of the rain-fed niche may offer additional insights.
-Consider the fact that of the 183 watersheds in the study area, 87% (n=159) have median GDD values greater than 1000°C, but only 54% (n=98) have median precipitation values above 300 mm, and a scant 40% (n=74) meet both requirements.
-This would suggest that precipitation is the primary limiting factor when it comes to the choice of watersheds in which to dry farm, for there are many more watersheds that meet the minimum GDD requirements than meet the minimum precipitation requirements.
-One important implication of this is that dry farming would have been very nearly impossible across most of the study area over the Fremont sequence and irrigation more or less necessary.
-We note that even within the niche, which we have defined somewhat liberally as those watersheds that are included for at least half the Fremont sequence, irrigation would almost certainly have been a necessity [@boomgarden2019; @matson1990; @spangler2010].
-Of course, the type and intensity of irrigation would probably differ between sites, with more intensive irrigation occurring where there is less precipitation and less intensive irrigation where there is more precipitation, whether that precipitation falls in the watersheds themselves or in the upstream watersheds that feed into them.
+While site counts in our model exhibit a positive linear response to maize GDD,
+this should not be expected to continue indefinitely for the obvious reason that
+maize has an upper threshold temperature at which point it no longer grows.
+Plus, increased temperatures trade-off with precipitation, with warmer
+conditions being better for plant growth but at the same time increasing water
+demand on the plant to keep up with growth [@ramankutty_etal2002]. This
+trade-off probably also explains the curious result that site counts do not
+increase linearly as a function of precipitation, but instead decline after 500
+mm.
+
+Provided that the trade-off between precipitation and temperature is driven
+largely by elevation in this region, it would seem likely that the western
+Fremont chose site locations at elevations with the best combination of
+temperature and precipitation for maize farming in the region, as evidenced by
+the estimated distribution of sites across watersheds (see @fig-thresholds),
+though we caution that watersheds are coarse grained units of analysis. Given
+the topography of the study area, this trade-off would also suggest that areas
+suitable for maize agriculture take the form of a thin band around the lower
+slopes of mountain ranges.
+
+It is tempting to think of this elevation band as a Goldilocks Zone for maize
+agriculture [@yaworsky_etal2023], but we should take care not to over-interpret
+this metaphor, for the elevation band is an area of overlap for continuous
+ecological gradients, not a region demarcated by anything like a perfect line.
+There is also variation in site density within that band, perhaps suggesting
+missing covariates in our model. The unaccounted for variation could also be
+owing to the fact that what matters most for maize agriculture is not
+necessarily where precipitation falls, but where it accumulates once it enters
+the stream flow network. Of course, that will largely be a function of elevation
+in this region, or changes in elevation, so it will still be in proximity to
+this optimal elevation band, though perhaps not perfectly coincident with it.
+
+At any rate, given these considerations, it would perhaps be more accurate to
+characterize the trade-off confronting the Fremont as one between what we might
+call water availability (including general precipitation across the stream flow
+network, water runoff, proximity to perennial surficial sources, soil moisture,
+and irrigation potential, among other things) and temperature (including, but
+not limited to, the length of the growing season and the number of frost-free
+days).
+
+## Water Management and the Rain-fed Maize Farming Niche
+
+Based on previous research [@benson2011; @adams_etal2006; @bellorado2010],
+@bocinsky_kohler2014 place minimum thresholds for precipitation and maize GDD at
+300 mm and 1000°C GDD, respectively. Importantly, these are thresholds for
+dry-farming, which is typically defined as farming free of irrigation
+[@benson2011; @cordell_mcbrinn2012; @varien1999], so they can be loosely
+interpreted as minimum temperature and precipitation levels required to grow
+maize through nothing more than planting and harvesting. The right panel in
+@fig-thresholds shows the watersheds whose median values for maize GDD and
+precipitation over the Fremont sequence are above those values. This is similar
+to Bocinsky's concept of "refugia," but rather than calculate the proportion of
+years in niche, we are using median values, so refugia are defined here as
+watersheds that spend at least half the Fremont sequence in the rain-fed maize
+farming niche ($\geq$ 500 years, in this case).
+
+::: {#fig-thresholds fig-align="center"}
+![](../figures/threshold_map.jpg){width="5in"}
+
+The map on the left shows the log density of feature-weighted site counts as
+estimated by the model, with lighter colors representing larger relative
+densities, darker colors representing smaller relative densities. The map on the
+right shows which watersheds are in the rain-fed maize farming niche for at
+least half of the Fremont sequence. The light green color indicates those that
+are in the niche.
+:::
+
+It appears that these models tend to agree on the best watersheds for maize
+farming, though there are some notable exceptions. The western slopes of the
+Stansbury Range in the north central part of the study area (just west of Salt
+Lake City) and parts of the Parowan Valley and surrounding areas in the
+southeast part of the study area (the area roughly midway between Cedar City and
+Richfield) are estimated by our model to have non-negligible densities, but they
+fall outside of the niche. Other areas, like much of the Salt Lake valley appear
+to be in the niche, but are estimated by our model to have relatively low
+densities. These discrepancies are at least partially explained by limitations
+of our methodology. For one thing, including the proportion of federal land as a
+covariate is a very crude way of accounting for sampling bias, so our model
+might not be picking up on the fact that the Salt Lake valley was actually a
+highly suitable area for maize agriculture, as evidence to that effect has
+largely been destroyed. The discrepancy may also be explained by the fact that
+we have aggregated to the watershed level, thus obscuring important
+environmental variation across both the landscape in general and site locations
+in particular. Were we to build a model of the disaggregated data, we would
+likely find that many agricultural sites do, in fact, fall in the rain-fed maize
+agricultural niche, even if the watersheds that contain them do not.
+
+But setting aside those methodological issues, some much more interesting
+explanations may be offered, at least from the perspective of theory. Consider
+the fact that of the 183 watersheds in the study area, 87% (n=159) have median
+GDD values greater than 1000°C, but only 54% (n=98) have median precipitation
+values above 300 mm, and a scant 40% (n=74) meet both requirements. This would
+suggest that precipitation is the primary limiting factor when it comes to the
+choice of watersheds in which to dry farm, for there are many more watersheds
+that meet the minimum GDD requirements than meet the minimum precipitation
+requirements. One important implication of this is that dry farming would have
+been very nearly impossible across most of the study area over the Fremont
+sequence and irrigation or some other form of water management more or less
+necessary. We note that even within the niche, which we have defined somewhat
+liberally as those watersheds that are included for at least half the Fremont
+sequence, water management would almost certainly have been a necessity
+[@boomgarden_etal2019; @matson_etal1990; @spangler_etal2010]. Of course, the
+type and intensity of water management would probably differ between sites, with
+more intensive management occurring where there is less precipitation and less
+intensive management where there is more precipitation, whether that
+precipitation falls in the watersheds themselves or in the upstream watersheds
+that feed into them.
 
@@
 
-Our results comport with previous work suggesting that maize farmers in the eastern Great Basin preferentially targeted a certain elevation band that coincides with alluvial fans, forming a rim around the lower slopes of mountains and ridges on or near important drainages and floodplains.
-The ruggedness and aridity of this region led to extreme circumscription, with the costs of intensive farming outside these areas being so severe as to render that alternative virtually unsustainable.
-For maize farmers, the Great Basin should, thus, have the look and feel of an island biogeography, with mountains rising up like islands out of an ocean of desert sand and sagebrush.
+Our results comport with previous work suggesting that maize farmers in the
+eastern Great Basin preferentially targeted a certain elevation band that
+coincides with alluvial fans, forming a rim around the lower slopes of mountains
+and ridges on or near important drainages and floodplains. The ruggedness and
+aridity of this region led to extreme circumscription, with the costs of
+intensive farming outside these areas being so severe as to render that
+alternative virtually unsustainable. For maize farmers, the Great Basin should,
+thus, have the look and feel of an island biogeography, with mountains rising up
+like islands out of an ocean of desert sand and sagebrush.
 
-To account for the timing and tempo of island colonization among food producers in Oceania, @kennett2006 offer an extension of the IFD framework that might actually apply to the Great Basin.
-Their model depends crucially on the potential for farming populations to introduce economies of scale or Allee effects, with suitability increasing with increasing population at low densities.
-They also assume that individuals will prefer nearer habitats to those that are farther away, under the assumption that greater distances impose greater settlement costs, all else being equal.
-Together, these assumptions suggest a pulsating pattern of maize spread into the North American Southwest, with maize agriculture growing, then spreading, growing, then spreading, as appears to have happened in Oceania.
+To account for the timing and tempo of island colonization among food producers
+in Oceania, @kennett_etal2006 offer an extension of the IFD framework that might
+actually apply to the Great Basin. Their model depends crucially on the
+potential for farming populations to introduce economies of scale or Allee
+effects, with suitability increasing with increasing population at low
+densities. They also assume that individuals will prefer nearer habitats to
+those that are farther away, under the assumption that greater distances impose
+greater settlement costs, all else being equal. Together, these assumptions
+suggest a pulsating pattern of maize spread into the North American Southwest,
+with maize agriculture growing, then spreading, growing, then spreading, as
+appears to have happened in Oceania.
 
-This need not be a wholesale movement or adoption of agricultural practices, either.
-@madsen1998 argue that Fremont subsistence largely varies along two dimensions: levels of residential mobility and levels of cultigen adoption.
-While some were more mobile, moving from isolated patch to isolated patch, others were more sedentary, typically tethered to a productive wetland or riparian area.
-As local populations increase, the benefits of increased sedentism and cultigen adoption would slowly begin to outweigh the benefits of a more mobile and foraging-centered diet.
-According to our model, individuals moving in this direction would tend to concentrate more of their time and energy in the maize farming niche.
-This would then have the effect of increasing population size within a smaller area, possibly leading to Allee effects, which would invite further sedentism from those in the surrounding area.
-Following this logic, the patchy adoption of agriculture makes more sense.
-The degree and speed with which it occurs is simply a function of how fast habitat quality declines and how quickly populations grow.
+This need not be a wholesale movement or adoption of agricultural practices,
+either. Early on, subsistence would have varied along two dimensions: levels of
+residential mobility and levels of cultigen adoption. While some individuals
+would have been more mobile, moving from isolated patch to isolated patch,
+others would have been more sedentary, typically tethered to a productive
+wetland or riparian area. As local populations increase, the benefits of
+increased sedentism and cultigen adoption would slowly begin to outweigh the
+benefits of a more mobile and foraging-centered diet. According to our model,
+individuals moving in this direction would tend to concentrate more of their
+time and energy in the maize farming niche. This would then have the effect of
+increasing population size within a smaller area, possibly leading to scale
+effects, which would invite further sedentism from those in the surrounding
+area. Following this logic, the patchy adoption of agriculture makes more sense.
+The degree and speed with which it occurs is simply a function of how fast
+habitat quality declines and how quickly populations grow.
 
-![
-Geographic distribution of Fremont sites across watersheds.
-The left panel shows the relative density derived from observed site counts, and the right panel shows the relative density from the estimated site counts.
-](../figures/model-map.png){#fig-model-map fig-align="center" width="6in"}
+::: {#fig-model-map fig-align="center"}
+![](../figures/model-map.jpg){width="5in"}
+
+Geographic distribution of Fremont sites across watersheds. The left panel shows
+the relative density derived from observed site counts, and the right panel
+shows the relative density from the estimated site counts.
+:::
 
@@
 
-After farming these regions successfully for over a thousand years, the Fremont began an abrupt process of abandonment starting around 700 years BP.
-The cause of this event is still a matter of some dispute, though a growing consensus appears to be coalescing around climatic changes driving severe reductions in agricultural productivity [@benson2007; @spangler2019; @thomson2019; @thomson2020; @finley2020].
-We follow @lindsay1986 in thinking this was likely owing to climatic events that, in effect, tore the Goldilocks zone apart, pulling higher temperatures further down in elevation and pushing higher precipitation levels further up.
+After farming these regions successfully for over a thousand years, the Fremont
+began an abrupt process of abandonment starting around 700 years BP. The cause
+of this event is still a matter of some dispute, though most scholars agree that
+climate change drove severe reductions in agricultural productivity
+[@benson_etal2007; @spangler_etal2019; @thomson_etal2019;
+@thomson_macdonald2020; @finley_etal2020]. We follow @lindsay1986 in thinking
+this was likely owing to climatic events that, in effect, severed the connection
+between the optimal temperature and precipitation levels that made maize farming
+a reliable strategy, pulling higher temperatures further down in elevation and
+pushing higher precipitation levels further up.
 
-Lindsay suggests that the end of the Medieval Warm Period may have been responsible for this decoupling, as it involved higher temperatures and more growing-season precipitation owing to the northward intrusion of summer monsoons.
-This period lasted from roughly 1,000 to 600 years BP [@grayson2011; @graumlich1993], so it does coincide with the height of the Fremont complex.
-Given the logic of IFD, this likely led to population spillovers into less suitable habitats [@codding2021].
-However, given the extreme circumscription we have highlighted in the Great Basin, the available alternatives would have been limited, potentially leading to increasing population packing across all habitats during this time, thus making the Fremont especially sensitive to climatic downturns.
+Lindsay suggests that the end of the Medieval Warm Period may have been
+responsible for the initial coupling of these ecological gradients, as it
+involved higher temperatures and more growing-season precipitation owing to the
+northward intrusion of summer monsoons. This period lasted from roughly 1,000 to
+600 years BP [@grayson2011; @graumlich1993], so it does coincide with the height
+of the Fremont complex. Given the logic of IFD, this likely led to population
+spillovers into less suitable habitats [@codding_etal2022]. However, given the
+extreme circumscription we have highlighted in the Great Basin, the available
+alternatives would have been limited, potentially leading to increasing
+population packing across all habitats during this time, thus making the Fremont
+especially sensitive to climatic downturns.
 
-The well-documented megadrought at the end of the thirteenth century [@cook2004] would have been catastrophic in this context, reducing precipitation levels across the region [@benson2007], but also leading to reductions in water accumulation within the streamflow network.
-So, just as temperatures are beginning to cool, precipitation levels collapse.
-By extension, this would have decreased overall water availability and, thus, the potential for irrigation.
-Opportunities for the Fremont to recover from such a drought would have been limited, too, as the end of the Medieval Warm Period was also the beginning of the Little Ice Age, a time during which conditions were generally colder and wetter [@mann2009], further pulling apart the Goldilocks zone upon which the Fremont so long depended.
+The well-documented megadrought at the end of the thirteenth century
+[@cook_etal2004] would have been catastrophic in this context, reducing
+precipitation levels across the region [@benson_etal2007], but also leading to
+reductions in water accumulation within the streamflow network. So, just as
+temperatures are beginning to cool, precipitation levels collapse. By extension,
+this would have decreased overall water availability and, thus, the potential
+for irrigation. Opportunities for the Fremont to recover from such a drought
+would have been limited, too, as the end of the Medieval Warm Period was also
+the beginning of the Little Ice Age, a time during which conditions were
+generally colder and wetter [@mann_etal2009], further pulling apart the
+combination of temperature and precipitation upon which the Fremont so long
+depended.
+
+\pagebreak
+
+# References
+
+::: {#refs}
+:::

### 2024-02-07 R/fig-elevation-tradeoffs.R
delete redundant code

@@ -133,10 +133,3 @@ ggsave(
   height = 3,
   dpi = 300
 )
-
-ggsave(
-  here("figures", "elevation-everything.jpg"),
-  width = 4,
-  height = 3,
-  dpi = 600
-)

### 2024-02-07 R/fig-overview-map.R
just use source()

@@ -27,16 +27,8 @@ library(viridis)
 gpkg <- here("data", "western-fremont.gpkg")
 
 # load custom functions
-# doing it this way to mask the functions in the global environment
-sys.source(
-  here("R", "fun-get_basemap.R"),
-  envir = attach(NULL, name = "basemap")
-)
-
-sys.source(
-  here("R", "fun-prepare-image.R"),
-  envir = attach(NULL, name = "image")
-)
+here("R", "fun-get_basemap.R") |> source()
+here("R", "fun-prepare-image.R") |> source()
 
 # features ----------------------------------------------------------------
 

### 2024-02-07 R/rebuild-geopackage-database.R
code that will hopefully rebuild the gpkg

@@ -1,0 +1,213 @@
+
+# just putting this together so someone can source it to rebuild the
+# geopackage that i use for storing all the spatial and attribute data
+# not the most elegant way of doing this, but oh  well...
+# also, i haven't tested this... \shruggie
+
+# Load libraries
+library(FedData)
+library(here)
+library(sf)
+library(spData)
+library(terra)
+library(tidyverse)
+library(tigris)
+
+# geopackage database
+gpkg <- here("data", "western-fremont.gpkg")
+
+# coordinate reference 
+# EPSG:26912
+
+# requires httr and jsonlite
+here("R", "fun-download_features.R") |> source()
+
+# 02) UTAH POLYGON --------------------------------------------------------
+
+utah <- states() |> 
+  filter(NAME == "Utah") |> 
+  select(NAME) |> 
+  rename_with(tolower) |> 
+  st_transform(26912)
+
+write_sf(
+  utah, 
+  dsn = gpkg, 
+  layer = "utah"
+)
+
+# 03) WATERSHEDS ----------------------------------------------------------
+
+watersheds <- utah |> 
+  # to make sure we pick up the HUC12 inside the HUC8 but outside UT
+  st_buffer(30000) |> 
+  get_wbd(label = "western_fremont") |> 
+  st_transform(26912) |> 
+  select(HUC12) |> 
+  rename("id" = HUC12) |> 
+  mutate(id = substr(id, start = 1, stop = 10)) |> 
+  group_by(id) |> 
+  summarize()
+
+window <- watersheds |> 
+  mutate(id = substr(id, start = 1, stop = 8)) |> 
+  group_by(id) |> 
+  summarize() |> 
+  mutate(id = as.numeric(id)) |> 
+  filter(
+    id > 16010204,
+    id < 16040101, 
+    id != 16020307,
+    id != 16020309
+  ) |> 
+  st_union() |> 
+  st_as_sf()
+
+# weird holes introduced into this geometry
+bob <- st_geometry(window)[[1]][1]
+bob <- bob |> 
+  st_polygon() |> 
+  st_sfc(crs = 26912)
+
+window <- window |> st_set_geometry(bob)
+
+watersheds  <- watersheds |> st_filter(window, .predicate = st_within)
+
+write_sf(
+  window, 
+  dsn = gpkg, 
+  layer = "window"
+)
+
+remove(bob)
+
[…]
